---
title: "LLM word-learning experiments: methods, results, and paper claims"
subtitle: "GPT-2 vs children on a development-matched axis"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: kable
execute:
  warning: false
  message: false
---

```{r setup}
knitr::opts_knit$set(root.dir = here::here())  # resolve data paths from repo root
library(dplyr); library(readr); library(tidyr); library(ggplot2); library(purrr)
```

## What this notebook is

A walkthrough of the LLM side of the paper. The paper's two headline signatures
of children's vocabulary growth are **efficiency** (the intercept ξ_i — overall
learning rate) and **acceleration** (the slope κ_i — how fast learning speeds up
with experience), with between-child *variability* a secondary signature. The
model is M4: $\theta_i(t) = \xi_i + \kappa_i \log(t/a_0)$, and **κ = 1 is the
"unit accumulator"** (no acceleration); children sit at **κ_pop ≈ 10** (strong
acceleration).

The LLM section's claim (paper §"Large language models do not show acceleration")
is the **acceleration disanalogy**: GPT-2 — even trained on child-directed speech
— behaves as a **pure accumulator (κ ≈ 1)**, with very limited variability. This
notebook documents the experiments behind that claim, the statistic each uses,
and how each maps to the paper. Narrative log:
[`journal/experiments_llm.md`](../../journal/experiments_llm.md).

**Read §1 first** — it is the *linking hypothesis*: why an LM's per-word learning
curve has a slope we can identify with the model's κ. This is the non-obvious
step the whole comparison rests on.

::: {.callout-note}
## The five experiments at a glance
| | experiment | statistic | "experience" axis | answers |
|---|---|---|---|---|
| **L1** | CHILDES-trained GPT-2, 3 seeds (vs BookCorpus) | C&B per-word slope | **training steps** (within run) | input distribution ≠ the gap |
| **L2** | data-variance pilot, 2 disjoint 10M chunks | C&B per-word slope | **training steps** | data *identity* ≈ no variance |
| **L3** | developmental ladder, 1 seed × 6 budgets | best-val competence | **distinct input** (across runs) | *(pilot for L4)* |
| **L4** | developmental ladder, 5 seeds × 11 budgets | C&B per-word slope **on the budget axis** + aggregate | **distinct input** | acceleration (κ≈1), + variability, shape |
| **L5** | disjoint-halves control, 2 pools × 3 seeds | aggregate developmental slope | **distinct input** | convergence ≠ overlap artifact |
:::

## 1. The linking hypothesis: why a per-word slope *is* κ

This is the conceptual move the whole comparison depends on, so it's worth being
explicit. It has two steps: (a) why, *in our model*, a per-word learning curve has
slope κ; and (b) why we can read the *same* quantity off a language model.

### 1a. In the accumulator model, the per-word slope on log-experience **is** κ

In the model, a child's latent ability accumulates with experience as
$$\theta_i(t) = \xi_i + \kappa_i \,\log(t/a_0),$$
and each word *j* is produced with a Rasch/IRT probability that compares ability
to the word's difficulty $\delta_j$:
$$P(\text{produce } j \mid i, t) = \operatorname{logistic}\big(\theta_i(t) - \delta_j\big)
   = \operatorname{logistic}\big(\underbrace{\kappa_i}_{\text{slope}}\log(t/a_0) + \underbrace{(\xi_i - \delta_j)}_{\text{offset}}\big).$$

So for **any single word**, $P(\text{known})$ traces a **logistic (sigmoid) in
log-experience**. The word's difficulty $\delta_j$ and the child's efficiency
$\xi_i$ only set the **offset** — *when* the word crosses 50% (its midpoint). The
**slope** of that sigmoid on the log-experience axis is **κ_i** — the
acceleration, shared across all of a child's words.

That gives the boundary that organizes everything:

- **κ = 1** is the **unit accumulator** (McMurray): ability = log(cumulative
  evidence), so one log-unit of evidence buys exactly one logit of competence —
  *no acceleration*.
- **κ > 1**: each log-unit of evidence buys *more* than one logit — self-reinforcing
  **acceleration**. Children sit at κ ≈ 10.

### 1b. Chang & Bergen read the same slope off an LM

We can't see an LM's latent θ, but we *can* see its competence on each word as a
function of experience: the (negative) surprisal it assigns, which maps
monotonically to a probability. Chang & Bergen fit, **per word**, the same
4-parameter logistic to (mean surprisal) vs log-experience:
$$S_w(x) = \text{lower}_w + \frac{\text{upper}_w - \text{lower}_w}{1 + \exp((x - \text{mid}_w)/\text{scale}_w)},
\qquad \text{slope} = 0.434 / \text{scale}_w$$
(0.434 = $\log_{10}e$, converting the per-`log10` slope to logit-per-natural-log).
**The linking hypothesis is that this per-word slope is the LM's κ** — logits of
competence gained per log-unit of experience, in the same units as a child's κ_i.
Under it, a slope of 1 means the LM is a pure unit accumulator; a slope of 10
would mean it accelerates like a child. Implementation:
[`fit_per_word_sigmoid.py`](fit_per_word_sigmoid.py); C&B filters
`0.01 < scale < 10`, `(upper − lower) > 1` nat.

```{r linking-graphic}
#| fig-width: 7
#| fig-height: 3.2
#| fig-cap: "The linking hypothesis. Each curve is one word's P(known) (or scaled LM competence) vs log-experience. Difficulty/efficiency shift a word left/right (offset); the SLOPE through the midpoint is κ. κ=1 is the unit accumulator (LMs); children sit near κ=10."
x <- seq(-3, 3, length.out = 200)
sig <- function(x, kappa, mid) plogis(kappa * (x - mid))
df <- bind_rows(
  tibble(x, p = sig(x, 1, -1.2), kappa = "κ = 1  (unit accumulator — LMs)", word = "w1"),
  tibble(x, p = sig(x, 1,  0.4), kappa = "κ = 1  (unit accumulator — LMs)", word = "w2"),
  tibble(x, p = sig(x, 6, -1.0), kappa = "κ = 6  (accelerating — children)", word = "w1"),
  tibble(x, p = sig(x, 6,  0.6), kappa = "κ = 6  (accelerating — children)", word = "w2"))
ggplot(df, aes(x, p, group = interaction(kappa, word), color = kappa)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c("κ = 1  (unit accumulator — LMs)" = "#4575b4",
                                "κ = 6  (accelerating — children)" = "#d73027")) +
  labs(x = "log experience", y = "P(word known)", color = NULL,
       subtitle = "two words each; offset = when the word is learned, slope = κ = how fast") +
  theme_minimal(base_size = 11) + theme(legend.position = "top")
```

### 1c. Two wrinkles to keep explicit

1. **Per-child vs per-word.** Children's κ_i is *per-child* (one slope, shared
   across that child's words). The LM slope is *per-word* (it varies across words
   within one model). Both are "logit-per-log-experience" slopes; we compare the
   LM's per-word slope *distribution* to children's per-child κ_i *distribution*.
   Different "instances," same construct (acceleration rate).
2. **Which experience axis** (the L1-vs-L4 distinction):
   - **Training steps** (C&B original; L1, L2): surprisal vs `log10(steps)`
     *within one run*. Over 20 epochs the model re-sees the same corpus, so this
     mixes optimization convergence with data scaling — not strictly development.
   - **Distinct input** (the ladder; L4, L5): competence vs `log10(words)`
     *across separately-converged models*. The development-matched axis.

   Both yield a κ in the same units; the ladder is the cleaner developmental
   version. **Both land at κ ≈ 1** — which is the robustness that matters.

```{r kid-reference}
kid <- tibble(kappa_pop = 10.3, sigma_kappa = 3.5)  # English M_best (SM2 fits): median kappa_i, between-child SD
```

## 2. L1 — Training-axis slopes: CHILDES vs BookCorpus

Retrained GPT-2-small on **CHILDES** (Feng et al. spec, 3 seeds) and refit the
C&B per-word sigmoid on the training-step axis, vs Chang & Bergen's
BookCorpus/WikiText LMs. **Claim tested:** is the kid-vs-LM gap an
*input-distribution* artifact (adult text vs child-directed speech)?

```{r L1}
# C&B filters: meaningful transition only (0.01 < scale < 10, surprisal range > 1)
load_sig <- function(path, label) {
  read_tsv(path, show_col_types = FALSE) |>
    transmute(model = label, scale = ParamScale, slope = 0.434/ParamScale,
              rng = ParamUpper - ParamLower) |>
    filter(is.finite(slope), scale > 0.01, scale < 10, rng > 1)
}
childes <- map_dfr(c(42,0,123), ~load_sig(
  sprintf("fits/llm/sigmoids/gpt2_childes_seed%d_sigmoids.txt", .x), sprintf("CHILDES seed %d", .x)))
book <- map_dfr(c("bert","gpt2","bilstm","lstm"), ~load_sig(
  sprintf("data/chang_bergen_2022/%s_sigmoids.txt", .x), sprintf("BookCorpus %s", .x)))
bind_rows(childes, book) |> group_by(model) |>
  summarise(n = n(), median_slope = round(median(slope),3),
            iqr = sprintf("[%.2f, %.2f]", quantile(slope,.25), quantile(slope,.75)), .groups="drop") |>
  arrange(median_slope)
```

**Result.** CHILDES-trained seeds cluster at **0.72–0.74**, inside the BookCorpus
cluster (0.76–0.96) — all ~14× below children (κ ≈ 10). **CDS-matched training
does not move the slope**: input distribution explains ≈ 0 of the gap.
*(Also note the tight 3-seed agreement — first hint that LM between-instance
variance is ~0; σ across seed medians ≈ 0.01.)*

::: {.callout-important}
## Caveat carried by L1
This is the **training-step axis** — the slope is partly an optimization-rate, not
a pure developmental rate. L4 redoes it on the distinct-input axis and the
conclusion survives. The C&B figure in the deck (`chang_bergen_slope_summary.csv`)
is this analysis.
:::

## 3. L2 — Data-identity pilot (variance from *which* data)

Two GPT-2-small on **two disjoint 10M-word CHILDES chunks**, same seed/tokenizer/
eval. **Claim tested:** does the *identity* of the training data create
between-instance variance? (Gates whether we need many disjoint corpora.)

```{r L2}
chunkAB <- map_dfr(c("A","B"), ~{
  read_tsv(sprintf("fits/llm/sigmoids/gpt2_childes_chunk%s_seed42_sigmoids.txt", .x), show_col_types = FALSE) |>
    transmute(chunk = .x, Token, scale = ParamScale, slope = 0.434/ParamScale, rng = ParamUpper-ParamLower) |>
    filter(is.finite(slope), scale > 0.01, scale < 10, rng > 1)
})
paired <- chunkAB |> select(chunk, Token, slope) |>
  pivot_wider(names_from = chunk, values_from = slope) |> filter(!is.na(A), !is.na(B))
tibble(
  pearson = round(cor(paired$A, paired$B), 3),
  spearman = round(cor(paired$A, paired$B, method = "spearman"), 3),
  median_A = round(median(paired$A), 3), median_B = round(median(paired$B), 3),
  paired_median_diff = round(median(paired$A - paired$B), 3))
```

**Result.** Per-word slopes across *disjoint* data correlate **Pearson 0.76 /
Spearman 0.88**; the per-LM summary shifts only ~0.01–0.07 vs children's σ_κ ≈ 3.5.
**Data identity is a negligible variance source.** (Same C&B statistic, training
axis. Full write-up + figure: `pilot_data_variance_plot.R`.)

## 4. L4 — The developmental ladder (the main result)

5 seeds × 11 nested budgets (0.5M…24M words), each trained **to best-val
convergence**; competence = held-out CDI surprisal at the best-val model. Each
seed = an "individual" (its own init + input stream). This is the
**distinct-input axis**. Both the aggregate developmental slope and the per-word
C&B sigmoid **on the budget axis** are computed here.

```{r L4-load}
d <- read_csv("fits/llm/ladder_bestval.csv", show_col_types = FALSE) |>
  mutate(surprisal = as.numeric(surprisal), lnw = log(words), l10w = log10(words))
agg <- d |> group_by(seed, words, lnw) |> summarise(m = mean(surprisal), .groups="drop")
```

### 4a. Shape: development *decelerates* (vs children accelerate)

```{r L4-shape}
#| fig-width: 8
#| fig-height: 4
ggplot(agg, aes(words, m, color = factor(seed))) +
  geom_line(alpha=.8) + geom_point(size=1) +
  scale_x_log10(breaks=c(.5,1,2,4,8,16,24)*1e6, labels=c("0.5M","1M","2M","4M","8M","16M","24M")) +
  labs(x="distinct input (words, log)", y="best-val CDI surprisal (nats)", color="seed",
       title="Competence improves with diminishing returns (concave) — across all 5 seeds") +
  theme_minimal()
```

Surprisal 7.8 → 4.6 nats, **concave in log-input** (Δ per e-fold shrinks): the
neural scaling-law shape. Children *accelerate* (super-linear vocab growth). Same
competence axis, opposite curvature.

### 4b. Rate: per-word κ on the development axis ≈ 1.2

The C&B 4-PL, now fit per word to competence vs `log10(budget)` across the 11
converged models (this is `ladder_analysis_final.R`):

```{r L4-kappa}
four_pl <- function(x, y) tryCatch({
  f <- nls(y ~ lo + (up-lo)/(1+exp((x-mid)/sc)),
           start=list(up=max(y), lo=min(y), mid=mean(x), sc=0.5),
           lower=c(up=min(y)-5, lo=min(y)-5, mid=min(x)-3, sc=1e-3),
           upper=c(up=max(y)+5, lo=max(y)+5, mid=max(x)+3, sc=50),
           algorithm="port", control=nls.control(maxiter=200, warnOnly=TRUE))
  tibble(sc = unname(coef(f)["sc"]), rng = unname(coef(f)["up"]-coef(f)["lo"]))
}, error=function(e) tibble(sc=NA_real_, rng=NA_real_))

pw <- d |> group_by(seed, word) |> filter(n()==11) |>
  group_modify(~four_pl(.x$l10w, .x$surprisal)) |> ungroup() |>
  mutate(kappa = 0.434/sc) |> filter(sc>0.01, sc<10, rng>1)
kpseed <- pw |> group_by(seed) |> summarise(kappa_med = median(kappa), .groups="drop")
tibble(kappa_pop_LM = round(mean(kpseed$kappa_med),3),
       sigma_kappa_LM = round(sd(kpseed$kappa_med),4),
       CV_pct = round(100*sd(kpseed$kappa_med)/mean(kpseed$kappa_med),1),
       kids_kappa = 10.3, kids_sigma = 3.5, kids_CV_pct = 34)
```

**κ_dev ≈ 1.17** (per-seed medians 1.05–1.24) — matches the training-axis ~0.74
ballpark, ~9× below children. The shallow slope is **not** an axis artifact.

### 4c. Variability: between-seed σ_κ ≈ 0.08 (CV 7%) — categorically absent

The five individuals' per-word κ medians span 1.05–1.24: **σ_κ ≈ 0.08, CV 7%**, vs
children's σ_κ ≈ 3.5 / CV ≈ 34%. And the seeds **converge with input** — between-seed
competence SD shrinks 0.12 → 0.03 nats (0.5M → 16M), the opposite of children's
age-divergence.

```{r L4-converge}
agg |> group_by(words) |> summarise(between_seed_sd = round(sd(m),3), .groups="drop")
```

**Two different "variabilities" live here — and they are firmed up by different
runs.** This is the figure to look at when deciding on a new run:

```{r L4-variability-fig}
#| fig-width: 9
#| fig-height: 3.6
#| fig-cap: "Two variabilities, often conflated. LEFT — between-INDIVIDUAL: the 5 seeds' per-word κ medians (the paper's σ_κ analog). Tightly clustered (σ_κ ≈ 0.08); the per-seed median is robust at 11 budgets, so MORE SEEDS only tightens its CI. RIGHT — between-WORD within a model (C&B's per-word κ distribution). Jagged because each word's κ is a 4-PL sigmoid fit to only 11 budget points; a FINER LADDER cleans this. The proposed new run targets the right panel. Children sit off-scale (κ ≈ 10, σ_κ ≈ 3.5)."
library(patchwork)
pA <- ggplot(kpseed, aes(1, kappa_med)) +
  geom_hline(yintercept = mean(kpseed$kappa_med), linetype = "dashed", color = "grey50") +
  geom_jitter(width = .04, height = 0, size = 3, color = "#2c7fb8") +
  annotate("text", x = 1.18, y = mean(kpseed$kappa_med),
           label = sprintf("σ_κ = %.3f\nCV %.0f%%", sd(kpseed$kappa_med),
                           100*sd(kpseed$kappa_med)/mean(kpseed$kappa_med)), hjust = 0, size = 3.1) +
  coord_cartesian(xlim = c(.85, 1.7), ylim = c(0, 2)) +
  labs(title = "between-INDIVIDUAL (5 seeds)", subtitle = "paper's σ_κ analog — robust",
       x = NULL, y = "per-seed median κ") +
  theme_minimal(base_size = 10.5) + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
pB <- ggplot(pw, aes(kappa, color = factor(seed))) +
  geom_vline(xintercept = 1, linetype = "dotted") +
  geom_density() + coord_cartesian(xlim = c(0, 3)) +
  labs(title = "between-WORD (within a model)", subtitle = "C&B per-word distribution — noisy at 11 budgets",
       x = "per-word κ", y = "density", color = "seed") +
  theme_minimal(base_size = 10.5) + theme(legend.position = c(.82, .7))
pA | pB
```

- **Left = between-individual** (different seeds = different "children"): this is
  the σ_κ that maps to the paper. It is **already robust** — five seed medians at
  1.05–1.24, σ_κ ≈ 0.08. A finer ladder won't move it; only more seeds tightens
  its confidence interval, and it's already far below children's 3.5.
- **Right = between-word** (different words within one model): C&B's per-word κ
  distribution. The **median** is the acceleration headline (κ ≈ 1.17, robust);
  the **jagged tails are fitting noise** from 11 budget points, *not* real
  between-word structure. This is the only thing a finer ladder fixes.

## 5. L5 — Disjoint-halves control (convergence isn't an overlap artifact)

The L4 nested design has two seeds sharing ≈ B/24.5M of their data (~98% at 24M),
so some convergence is mechanical. L5: **2 disjoint CHILDES halves** (A ∩ B = ∅) ×
3 seeds × 8 budgets. Decompose **between-pool** (disjoint data) vs
**within-pool-between-seed** developmental slope.

```{r L5}
dd <- read_csv("fits/llm/disjoint_bestval.csv", show_col_types = FALSE) |>
  mutate(surprisal=as.numeric(surprisal), lnw=log(words))
slp <- dd |> group_by(pool, seed, words, lnw) |> summarise(m=mean(surprisal), .groups="drop") |>
  group_by(pool, seed) |> summarise(slope = coef(lm(m ~ lnw))[2], .groups="drop")
pool_mean <- slp |> group_by(pool) |> summarise(s = mean(slope), .groups="drop")
tibble(
  between_pool_slope_gap = round(abs(diff(pool_mean$s)), 4),          # disjoint data
  within_pool_seed_sd    = round(mean(c(sd(slp$slope[slp$pool=="A"]), sd(slp$slope[slp$pool=="B"]))), 4),
  L4_nested_seed_sd      = 0.021)
```

**Result.** Between-pool slope gap **0.021** (zero shared data) ≈ within-pool
seed SD **0.026** ≈ the L4 nested figure **0.021**. Disjoint data adds **no**
extra between-individual variability. The convergence is genuine input-averaging,
not the overlap confound.

## 6. Synthesis — mapping to the paper

The paper's LLM section is **"Large language models do not show acceleration"**:
its existing figure plots children's per-child κ_i against LM per-word slopes on a
common axis, with the **κ = 1 unit-accumulator** line as the reference. Everything
here either *is* that result or strengthens it. Ordered by the paper's signatures
(acceleration first, then efficiency/variability):

```{r synthesis}
tribble(
  ~"signature / claim", ~evidence, ~axis, ~"in paper?",
  "ACCELERATION: LMs are pure accumulators, κ≈1 (no acceleration)", "L1 (κ 0.72-0.74), L4 (κ_dev 1.17)", "training + distinct-input", "yes (fig-llm, training axis) — L4 strengthens",
  "  ...not an input-distribution artifact (CDS vs adult)", "L1 (CHILDES ≈ BookCorpus)", "training", "yes",
  "  ...not a training-vs-development artifact", "L4 (κ≈1 on distinct-input axis too)", "distinct-input", "NEW",
  "  ...LM curve decelerates where kids accelerate (shape)", "L4 (concave, all 5 seeds)", "distinct-input", "NEW",
  "VARIABILITY (demoted): LM between-instance σ_κ≈0 ", "L1 seeds, L2 data, L4 σ_κ 0.08 (CV 7%)", "both", "abstract mentions; not yet a figure",
  "  ...convergence is real, not a data-overlap artifact", "L5 (disjoint pools, gap 0.021)", "distinct-input", "NEW",
  "  ...holds for compositionally different data", "BabyLM register-mix control", "distinct-input", "PENDING (backlog)"
)
```

**The headline that maps cleanest to the paper:** LMs sit at the **κ ≈ 1 unit-accumulator
boundary** that children exceed by ~10×. The paper already makes this on the C&B
*training-step* axis; **L4 is the same result on the development-matched
distinct-input axis** (κ_dev ≈ 1.17), which closes the obvious "but that's just
optimization, not development" objection — the strongest single addition. The
*deceleration* (concave vs children's convex) is a second, sharper framing of the
same "no acceleration" point. Variability is the paper's demoted signature, so the
σ_κ result and L5 are supporting, not headline.

### On the figure noise (the right panel of `ladder_development_final.png`)

That panel is the **per-word κ distribution** (5 seed densities). It's jagged
because each per-word κ is a 4-PL sigmoid fit to only **11 budget points** — too
few to pin a 4-parameter logistic per word, so the right tail and bumps are
**fitting noise, not signal**. Two distinct fixes for two distinct quantities:

- **Per-word κ *distribution* (the noisy panel) → finer ladder.** 15–20 budgets
  would constrain each word's sigmoid far better and smooth the density. This is
  the right fix if we want to *show* the distribution.
- **Between-individual σ_κ (the variability number) → more seeds.** σ_κ is the SD
  of 5 per-seed medians; the *median* per seed is already robust (so σ_κ ≈ 0.08 is
  trustworthy), but more seeds would tighten its confidence interval.

Both are now cheap on the A40s (the pipeline is proven; ~a night each). **But note
what we actually need for the paper:** the acceleration headline rests on the
*central* κ (median ≈ 1.17, robust at 11 budgets) vs children's ≈ 10 — that does
**not** need the finer ladder. The noisy per-word density only matters if we
promote the *variability* signature to its own figure; given the paper demotes
variability, the cleanest move is to (a) report the median κ for the acceleration
claim now, and (b) run a finer ladder only if we decide the per-word distribution
earns a panel. My recommendation: **finer ladder (≈16 budgets) × current 5 seeds**
— it cleans the distribution *and* tightens σ_κ in one sweep, for one night of A40 time.

**Provenance.** Raw surprisal trajectories: `fits/llm/surprisal/{grid,disjoint}/`;
dispatch logs + sampling manifests: `fits/llm/provenance/`; the best-val extracts
the analyses read: `fits/llm/ladder_bestval.csv`, `fits/llm/disjoint_bestval.csv`.

**Promote-to-paper now:** the κ≈1 acceleration disanalogy on *both* axes (extends
the existing fig-llm with the development-axis point + the deceleration framing) and
the L5 overlap control. Hold the per-word κ *distribution* for a finer ladder; frame
the BabyLM control as the planned
strengthening.
```
