---
title: "Cross-sectional demographic decomposition: sex & maternal education"
subtitle: "Efficiency vs. acceleration of vocabulary growth across Wordbank languages"
format:
  html:
    toc: true
    fig-width: 10
    code-fold: true
    embed-resources: true
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(here); library(dplyr); library(tidyr); library(ggplot2)
  library(forcats); library(stringr)
})
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 130)
DIR     <- here("cross_sectional_demographics")
fits    <- readRDS(file.path(DIR, "cache", "fits.rds"))
scatter <- readRDS(file.path(DIR, "cache", "scatter.rds"))

ED_YEARS <- c("None"=0,"Primary"=6,"Some Secondary"=9,"Secondary"=12,
              "Some College"=14,"College"=16,"Some Graduate"=18,"Graduate"=20)
ed_group <- function(y) cut(y, breaks = c(-1,12,16,99),
                            labels = c("≤ Secondary","College","Graduate+"))
# order languages by cross-sectional sex-efficiency (most female-advantaged first)
sex_order <- fits$xsec |> filter(predictor=="sex") |> arrange(desc(eff)) |> pull(language)
# maternal-ed languages with adequate n (drop Korean/Cantonese/Hebrew/Arabic)
MATED_MIN_N <- 300
mated_langs <- fits$xsec |> filter(predictor=="matEd", n_kids >= MATED_MIN_N) |> pull(language)
```

Reproducible companion to `paper/standard_model.qmd`'s longitudinal
demographic analysis. Same model on cross-sectional Wordbank data:
`produces ~ predictor * log(age/a0) + (1|item) + (1|child)`, monolingual-TD,
one admin per child. Built by `00_build.R`.

# 1. Data check: production by age & sex

```{r fig-scatter-sex}
#| fig-width: 11
#| fig-height: 12
sx <- scatter |> filter(!is.na(sex), sex %in% c("Male","Female")) |>
  mutate(language = factor(language, levels = sex_order))
ggplot(sx, aes(age, prop, colour = sex, fill = sex)) +
  geom_point(alpha = 0.05, size = 0.3) +
  geom_smooth(method = "loess", se = TRUE, span = 0.9, linewidth = 0.7, alpha = 0.15) +
  facet_wrap(~ language, ncol = 5) +
  scale_colour_manual(values = c(Female="#d81b60", Male="#1e88e5"), name = NULL) +
  scale_fill_manual(values = c(Female="#d81b60", Male="#1e88e5"), name = NULL) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "age (months)", y = "proportion of words produced") +
  theme_minimal(base_size = 9) + theme(legend.position = "bottom")
```

# 2. Data check: production by age & maternal education

```{r fig-scatter-mated}
#| fig-width: 11
#| fig-height: 8
me <- scatter |> filter(language %in% mated_langs, !is.na(ed_years)) |>
  mutate(ed_grp = ed_group(ed_years))
ggplot(me, aes(age, prop, colour = ed_grp, fill = ed_grp)) +
  geom_point(alpha = 0.05, size = 0.3) +
  geom_smooth(method = "loess", se = TRUE, span = 0.9, linewidth = 0.7, alpha = 0.15) +
  facet_wrap(~ language, ncol = 4) +
  scale_colour_brewer(palette = "Dark2", name = "Maternal ed.", na.translate = FALSE) +
  scale_fill_brewer(palette = "Dark2", name = "Maternal ed.", na.translate = FALSE) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "age (months)", y = "proportion of words produced") +
  theme_minimal(base_size = 9) + theme(legend.position = "bottom")
```

# 3. Cross-sectional demographic effects (parallels `fig-demographics`)

Per-language effects on **efficiency** (intercept at a0) and **acceleration**
(log-age slope), with a random-effects meta-analytic estimate. Sex coded as
*Female* effect (positive = female advantage); maternal ed per SD of years.

```{r fig-forest-xsec}
#| fig-width: 10
#| fig-height: 9
forest_df <- fits$xsec |>
  filter(predictor == "sex" | (predictor == "matEd" & language %in% mated_langs)) |>
  pivot_longer(c(eff, acc), names_to = "component", values_to = "estimate") |>
  mutate(se = if_else(component == "eff", eff_se, acc_se),
         component = recode(component, eff = "Efficiency", acc = "Acceleration"),
         component = factor(component, levels = c("Efficiency","Acceleration")),
         lo = estimate - 1.96*se, hi = estimate + 1.96*se,
         predictor = recode(predictor, sex = "Sex (female vs male)",
                            matEd = "Maternal education (per SD)"))
meta_df <- fits$meta |>
  mutate(component = recode(component, efficiency = "Efficiency", acceleration = "Acceleration"),
         component = factor(component, levels = c("Efficiency","Acceleration")),
         predictor = recode(predictor, sex = "Sex (female vs male)",
                            matEd = "Maternal education (per SD)"))

plot_forest <- function(pred) {
  d <- forest_df |> filter(predictor == pred) |>
    mutate(language = fct_reorder(language, estimate, .fun = function(x) x[1]))
  m <- meta_df |> filter(predictor == pred)
  ggplot(d, aes(estimate, language)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
    geom_pointrange(aes(xmin = lo, xmax = hi), size = 0.3, colour = "grey30") +
    geom_vline(data = m, aes(xintercept = estimate), colour = "#c41e37", linewidth = 0.5) +
    geom_rect(data = m, aes(xmin = ci.lb, xmax = ci.ub, ymin = -Inf, ymax = Inf),
              inherit.aes = FALSE, fill = "#c41e37", alpha = 0.12) +
    facet_wrap(~ component, scales = "free_x") +
    labs(x = "estimate (logits)", y = NULL, title = pred,
         subtitle = "grey = per-language; red band = random-effects meta-analytic 95% CI") +
    theme_minimal(base_size = 9) +
    theme(plot.title = element_text(face = "bold"))
}
plot_forest("Sex (female vs male)")
plot_forest("Maternal education (per SD)")
```

```{r meta-table}
fits$meta |>
  mutate(across(c(estimate, ci.lb, ci.ub, pval, tau2), ~round(., 3))) |>
  knitr::kable(caption = "Cross-sectional random-effects meta-analysis")
```

# 4. Composite figure: cross-sectional **and** longitudinal (paper candidate)

The paper-candidate figure (built by `composite_figure.R`, shared with the
manuscript). Panel A: per-language efficiency + acceleration for both
predictors, cross-sectional (● this work) and longitudinal (◆ paper BLUPs,
same units). Panel B: random-effects meta-analytic summary. Language set is
consistent across sex and maternal ed (maternal-ed n ≥ 300, French dropped —
see anomaly section).

```{r fig-composite}
#| fig-width: 8
#| fig-height: 13
source(here("cross_sectional_demographics", "composite_figure.R"))
make_demographics_composite(fits)
```

# 5. Anomaly checks

Flagged from the cross-sectional table: French (French) maternal-ed
*acceleration* < 0, and Spanish (European/Mexican) maternal-ed *efficiency*
< 0. Inspect the maternal-ed coding and the raw curves.

```{r anomaly-coding}
# what maternal-ed categories actually appear, and how the curves separate
chk <- scatter |> filter(language %in% c("French (French)","Spanish (European)","Spanish (Mexican)"),
                         !is.na(ed_years)) |>
  mutate(ed_grp = ed_group(ed_years))
chk |> count(language, ed_years, ed_grp) |>
  pivot_wider(names_from = language, values_from = n, values_fill = 0) |>
  knitr::kable(caption = "Maternal-ed (years) composition in the anomaly languages")
```

```{r fig-anomaly}
#| fig-width: 9
#| fig-height: 3.5
ggplot(chk, aes(age, prop, colour = ed_grp, fill = ed_grp)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.9, linewidth = 0.8, alpha = 0.15) +
  facet_wrap(~ language, scales = "free_x") +
  scale_colour_brewer(palette = "Dark2", name = "Maternal ed.") +
  scale_fill_brewer(palette = "Dark2", name = "Maternal ed.") +
  labs(x = "age (months)", y = "proportion produced") +
  theme_minimal(base_size = 9) + theme(legend.position = "bottom")
```
