---
author:
  - name: Michael C. Frank
    affiliations: 
      - name: Department of Psychology, Stanford University

title: Children but not language models learn words via accelerating accumulation 
#Children but not language models learn vocabulary via accelerating accumulation
#Efficiency and acceleration in the accumulation of children’s early vocabulary
title-short: Accelerating accumulator models
keywords: Early word learning; Bayesian data analysis; Large language models
abstract: |
  \noindent Children learn hundreds of words over the first few years of their life, in a process that begins slowly but quickly picks up speed. Prior models of word learning describe vocabulary growth as a process of accumulation of evidence over time. Here we show that the process is best characterized as *accelerating* accumulation, with children varying widely in both the overall efficiency with which they accumulate vocabulary and their acceleration rate. In contrast to children, language models (LMs) -- even those trained on child-directed speech -- show *deceleration* in their rate of learning and very limited variability, consistent with known laws governing LM scaling. Children learn using orders of magnitude less training data than language models do; acceleration is a candidate explanation. Creating acceleration in LMs -- perhaps by instilling children's ability to "learn to learn" -- is thus an open challenge. 

bibliography: standard_model.bib
# Dataset citations appear as plain text in Table 1 (kable can't render
# [@key] markers); list their bib keys here so they still appear in the
# References. Add @marchman / @smith dataset keys once they exist in the bib.
nocite: |
  @thal2013, @simonsen2014, @hagihara2023
format:
  pdf:
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
    linestretch: 1.05
    colorlinks: true
    keep-tex: true
    # Pin figures where they appear in the source instead of letting LaTeX float
    # them to a later page. Needs BOTH halves: fig-pos puts [H] on every figure,
    # and the float package below is what defines [H].
    fig-pos: 'H'
    # pdflscape: needed by kableExtra::landscape() (Table 1). Declared
    # explicitly because kableExtra's automatic dependency injection is
    # unreliable in Quarto (dropped on cached-chunk replay), which makes
    # the compile fail with "Environment landscape undefined."
    # float: defines the [H] specifier used by fig-pos above and by the tables'
    # kableExtra latex_options = "HOLD_position". The capitals matter: the
    # lowercase spelling of that option only emits [!h], which still permits
    # floating, whereas the uppercase one emits a true [H].
    include-in-header:
      text: |
        \usepackage{pdflscape}
        \usepackage{float}
        % Tables need this as well as the kableExtra option. When a chunk has a
        % tbl-cap, Quarto emits its OWN \begin{table} wrapper around the kable
        % output and kableExtra's placement is discarded, so the float came out
        % with no specifier at all (i.e. default [tbp], free to drift).
        % \floatplacement sets the default for every table float regardless of
        % who emits it. Long tables rendered as `longtable` are unaffected,
        % since longtable is not a float.
        \floatplacement{table}{H}
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(here); library(dplyr); library(tidyr); library(readr)
  library(ggplot2); library(patchwork); library(latex2exp)
  library(knitr); library(scales); library(stringr); library(forcats)
  library(broom); library(png); library(grid); library(gridExtra)
  library(ggrepel)
})
source(here("paper", "_helpers.R"))



# Cached data for the three main-text figures — produced by
# `Rscript paper/build_cache_short.R`.
CACHE <- here("paper", "cache")
fig1_fan   <- readRDS(file.path(CACHE, "fig1_fan.rds"))        # Fig 1: M0-M3 schematic + per-dataset M3 fan
fig2       <- readRDS(file.path(CACHE, "fig2_efficiency.rds")) # Fig 2: per-word exposures-to-learn (M3)
llm_slopes <- readRDS(file.path(CACHE, "fig6_llm_slopes.rds")) # Fig 3: child vs LLM acceleration
# QC cache
TR <- readRDS(file.path(CACHE, "qc_spaghetti_data.rds"))


# Inline text values (raw numbers in si_inline.rds, built by build_cache_short
# section 7); formatted here to the precision they render at in the text.
inl <- readRDS(file.path(CACHE, "si_inline.rds"))
.k1  <- function(x) sprintf("%.1f", x)                        # kappa: one decimal
.pct <- function(x) if (x >= 10) sprintf("%.0f", x) else sprintf("%.1f", x)
age_lo   <- inl$age_lo; age_hi <- inl$age_hi                  # integer months
loo_min  <- format(signif(inl$loo_min, 2), big.mark = ",")   # 2 sig figs
kap_lo   <- .k1(inl$kappa_lo); kap_lo_q5 <- .k1(inl$kappa_lo_q5); kap_lo_q95 <- .k1(inl$kappa_lo_q95)
kap_hi   <- .k1(inl$kappa_hi); kap_hi_q5 <- .k1(inl$kappa_hi_q5); kap_hi_q95 <- .k1(inl$kappa_hi_q95)
en_kappa <- .k1(inl$en_kappa); no_kappa <- .k1(inl$no_kappa)
en_sd    <- .k1(inl$en_sd);    no_sd    <- .k1(inl$no_sd)

# Random-effect summaries for main-text reporting (see tbl-ranef). sigma_b = between-
# child SD of acceleration; rho = efficiency-acceleration correlation; pct_gt1 = share
# of children above the pure-accumulator value kappa = 1; kappa_between_sd = SD of
# kappa ACROSS the five samples, the honest uncertainty scale given that CDI items are
# not locally independent (it is ~13x the narrowest within-sample interval).
sb_lo    <- .k1(inl$sb_lo);  sb_hi  <- .k1(inl$sb_hi)
rho_lo   <- sprintf("%+.2f", inl$rho_lo); rho_hi <- sprintf("%+.2f", inl$rho_hi)
pct_gt1  <- sprintf("%.1f", inl$pct_kappa_gt1)
kappa_between_sd <- .k1(inl$kappa_between_sd)

# Longitudinal sample + QC exclusions (bayes_long_sample.rds; built from the bundles by
# studies/bayes_long/qc_exclusion_report.R, so it renders independently of the fit caches).
bl_samp  <- readRDS(file.path(CACHE, "bayes_long_sample.rds"))
.bl      <- function(s) bl_samp$sample[bl_samp$sample$dataset == s, ]
qc_mar   <- .pct(.bl("marchman")$pct_kids_excluded)   # % children excluded, Marchman
qc_nor   <- .pct(.bl("norwegian")$pct_kids_excluded)  # % children excluded, Norwegian
n_total  <- format(.bl("total")$n_kids, big.mark = ",")      # kids retained, all datasets
n_excl   <- .bl("total")$kids_excluded                       # kids excluded, all datasets
pct_excl <- .pct(.bl("total")$pct_kids_excluded)

# dev = "cairo_pdf": the default pdf() device cannot encode non-ASCII glyphs. It fails
# with an mbcsToSbcs conversion error and substitutes ".", so theta, kappa and the
# up-arrow were rendering as "Latent Ability (.)", "Pure (.=1)" and ". Efficiency" in
# Fig 1B (and kappa/gamma in the LM figure). cairo_pdf handles UTF-8; it needs XQuartz
# for its X11 libraries, which is now installed. Verify with capabilities("cairo").
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE, cache=TRUE,
  fig.align = "center", out.width = "100%", dpi = 300, dev = "cairo_pdf"
)
```

Children’s language grows rapidly during early childhood [@frank2021;@kidd2020;@fenson1994], and the "explosion" of expressive vocabulary is a signature human cognitive achievement [@bloom2002;@spelke2003]. Providing quantitative characterization of this process is an important scientific goal [@frank2009;@mcmurray2012;@trueswell2013;@kachergis2012;@jones2017;@goodman2008;@hidaka2013;@roy2015;@braginsky2019;@kachergis2022;@kachergis2022b], but until recently, language acquisition could only be studied in human learners. Language models (LMs) show increasing linguistic sophistication, however, allowing them to become a model system for the study of human language [@huebner2021;@futrell2025;@frank2025;@evanson2023;@misra2024].

The most successful LMs differ from human learners in an important way: they are trained on vastly more data than human learners receive during childhood. While even open-source models are routinely trained on trillions of words of data [@gao2020], children hear on the order of 100k -- 1M words of input per month, with 36M words as an upper bound by age 3 [@frank2023;@warstadt2023;@hu2024]. Why do models require so many more orders of magnitude of training data to achieve basic linguistic competence? 

Here we study this "data gap" by comparing vocabulary growth in children and LMs. The number of words a child produces is an important measure of children's expressive abilities that can be measured with relative precision. Early vocabulary is also very tightly coupled to grammatical development and other aspects of early language [@bates1994;@marchman1994;@frank2021]. Vocabulary size varies widely between children of the same age [@frank2021],
and relative vocabulary size is a predictor of later academic success [@marchman2008;@catts2002;@muter2004]. 

Total vocabulary for young children can be approximated effectively using parent report forms, such as the MacArthur Bates Communicative Development Inventories (MB-CDI) [@marchman2021]. CDIs are a widely-used, reliable, and valid method for taking an in-depth snapshot of children's early vocabulary, and can be used longitudinally to track the growth of language over time [@fenson1994;@fenson2007;@marchman2021;@frank2021]. Here we use data from Wordbank, an open access database of cross-linguistic parent report data on the vocabularies of thousands of children ages 8 -- 36 months [@frank2017].

Our goal is to shed light on the process of vocabulary acquisition in both children and LMs by creating a framework in which they can be directly compared. We begin by developing a psychometric model of vocabulary growth in children -- an "accelerating accumulator" model. This model shows good fit to both individual growth trajectories and their variation across children. Next, we develop methods for measuring the acceleration of vocabulary growth in language models. Making use of transformer language models trained on human developmental data, we show that -- in contrast to children -- these models *decelerate* across training and show very limited variation between datasets or model instances. This analysis exposes a critical disanalogy between LMs and children and suggests new avenues for the study of both. 

# An accelerating accumulator model of vocabulary growth

Recent models describe children's vocabulary growth as a process of accumulation [@mcmurray2007; @mitchell2009; @hidaka2013; @mollica2017; @kachergis2022]. Each distinct word (type) can be conceptualized as a "bucket," and every token of language input then is a "drop" that falls into its respective bucket. When each bucket is full, the word is learned [@mcmurray2007;@kachergis2022]. Accumulator models have been used to recover interpretable differences in "bucket size," often based on word properties including concreteness and syntactic category [@hidaka2013;@kachergis2022]. They also provide a graded explanation for the widely-observed "vocabulary spurt" phenomenon -- in which children's rate of vocabulary learning appears to increase dramatically [@mcmurray2007;@frank2021;@ganger2004;@gomezdiaz2024]. 

We begin with the model of @kachergis2022, an accumulator model that operates over ratio-scale variables with real units: words (tokens heard) per hour and total words (types) in the child's vocabulary. The presence of a true zero in each of these variables means that they can be log transformed, which is critical for capturing the shape of their growth over development [@frank2026;@kail1991]. This feature distinguishes our modeling approach from other models of early word learning, which typically establish either schematic or variability-based linkages to developmental data [@frank2009;@mcmurray2012;@trueswell2013;@kachergis2012]. 

We use the Rasch model from psychometrics [@rasch1960;@bond2013], which assumes that each child $i$ has a learning ability $\theta_i$ and each word $j$ has a difficulty $\delta_j$.^[A variety of previous work has focused on estimating the properties of individual words  [@goodman2008;@hidaka2013;@roy2015;@braginsky2019;@kachergis2022;@kachergis2022b]. These analyses reveal that there is predictable word-level difficulty information but that only a modest portion of the variation between words is carried by frequency information. Words can be hard for a child to acquire because they vary also in the complexity and concreteness of their meanings, their syntactic category and syntactic context, their phonology, and many other factors. In the current work, we estimate word difficulty as a free parameter, and focus instead on variation between children.] The probability that child $i$ produces word $j$ is: 

$$P(w_{i,j}=1 \mid \theta_i, \delta_j) =\frac{\exp(\theta_i - \delta_j)}{1 + \exp(\theta_i - \delta_j)}.$$
  
We then describe the child's ability as

  $$\theta_i(t) = \xi_i + \kappa_i ~ \log(t/a_0) + \log(H),$$
\noindent which contains three terms: an intercept $\xi_i$ describing the child's overall learning efficiency, an acceleration term, $\kappa_i ~ \log(t/a_0)$ that describes the rate of accumulation growing exponentially in the log of the child's age, and a constant offset $H$ for the number of hours over which accumulation occurs (See @fig-schematic A). 

This set of terms describes an "accelerating accumulator" model, in which accumulating language "counts for more" later in development. Children vary on both their efficiency of accumulation $\xi_i$ and their acceleration $\kappa_i$, creating a variant of a standard longitudinal growth model in which children vary on their slope and intercept [@grimm2016]. Our model is close to the formulations of @hidaka2013 and @kachergis2022, but has not yet been proposed (see SI: Comparison with Prior Models).

Under this model, we can use the mean value of $\kappa$ to distinguish pure accumulator behavior ($\kappa \approx 1$) from accelerating accumulator behavior ($\kappa > 1$). It is critical to distinguish between supra-linear growth in children's *actual vocabulary* (in terms of count) -- a pattern that is robustly observed across children worldwide [@frank2021;@fenson1994] -- and the idea that children's *internal learning ability* accelerates across development. The latter is the substantive claim we test here (see @fig-schematic B for an illustration of model predictions in both latent space and vocabulary space).

# Fitting the accumulator model 

Our analysis depends on the availability of longitudinal data at the individual word level. Summed vocabulary scores do not allow for the estimation of word difficulties $\delta$, and without this component, different difficulty distributions can be posited to explain the accelerating shape of vocabulary growth [@mcmurray2007; @mitchell2009]. Further, cross-sectional data do not allow for identification of whether individual children's variation is due to differences in efficiency (intercept) or acceleration (slope) (see SI: Non-identifiability in Cross-Sectional Data). Thus, to constrain our model, we use item-level longitudinal data from CDIs archived in Wordbank. We analyze reported production, which tends to be a more reliable measure than reported comprehension [@frank2021]. 

We analyze five longitudinal datasets of monolingual, typically-developing children ages `r age_lo`--`r age_hi` (three from American English and one each from Norwegian and Japanese, $N_{total} = `r sum(fig1_fan$meta$n_kids)`$; see SI: Datasets). We fit the accelerating accumulator model (M3) to each dataset using Bayesian inference in Stan [@stancite]. To test whether each model component was necessary, we additionally fit three model ablations, removing individual variation in acceleration (removing $\kappa_i$; M2), removing individual variation in efficiency (removing $\xi_i$; M1), and finally removing acceleration entirely (removing $\kappa$; M0). Our primary results come from children with three or more longitudinal administrations, but all results hold on the larger dataset including children with two administrations. 

The accelerating accumulator provided the best fit to children's growth trajectories (@fig-schematic C, see SI: Full Model Comparison Results), with substantial differences in reliable improvements in item-level expected log predictive density between the accelerating accumulator and the next best model (LOO ELPD $>`r loo_min`$ for all five datasets). Indeed, fits from M3 corresponded closely to curves derived from a non-parametric GAMLSS beta regression, a flexible model class used to produce percentile norms for the CDI instruments [@marchman2021] (see SI: Comparison to a Non-Parametric Quantile Model).

Critically, for all datasets, estimates of acceleration were high, robustly rejecting the pure accumulator ($\kappa=1$) in every case :`r kap_lo` [`r kap_lo_q5`, `r kap_lo_q95`] -- `r kap_hi` [`r kap_hi_q5`, `r kap_hi_q95`].  $\kappa$ estimates were robust across model and dataset variants, including a 2PL IRT model and a hierarchical model fit to all datasets simultaneously (SI: 2PL Model Comparison and SI: Robustness Across Model and Data Settings). Linear (rather than logarithmic) growth models did not provide better fit (see SI: Linear vs. Logistic Age). 

Variability is a ubiquitous feature of early language across a vast array of phenomena and a wide range of languages [@frank2021;@kidd2020]. The between-child standard deviation in acceleration ($\sigma_b$) ranged from `r sb_lo` to `r sb_hi` across the five samples; `r pct_gt1`% of children were estimated above the pure-accumulator value of $\kappa = 1$. The distribution appeared unimodal across samples, but the increasing spread of children's trajectories with age was especially noticeable for low-vocabulary older children [who may be classified as "late talkers"; @fernald2012; @rescorla2011; @pinstock2026;@singletary2025] (see SI: Variation in Acceleration).

```{r fig-schematic, fig.width=7.5, fig.height=8.6}
#| label: fig-schematic
#| fig-cap: "(A) Generative model: each child $i$ has an efficiency $\\xi_i$ and an acceleration $\\kappa_i$ that set their latent ability; whether child $i$ produces word $j$ depends on $\\theta_i(t)$ relative to the word's difficulty $\\delta_j$. (B) Model predictions in latent-ability ($\\theta$) space and, projected through the logistic item response model, in words-produced (CDI) space. Both pure accumulator ($\\kappa = 1$, dashed) and accelerating accumulator predictions are plotted. Greater efficiency ($\\xi$) lifts the curve's level, while greater acceleration ($\\kappa$) fans it open. (C) Individual children's observed trajectories (gray) for each of the five longitudinal datasets, with quantiles computed across the per-child parameters of the fitted models (colored percentile bands)."

## Fig 1 precomputed in paper/build_cache_short.R (fig1_fan.rds). Three-block
## composite via cowplot: (A) graphical model image; (B) conceptual theta -> CDI
## mechanism (two facets); (C) per-dataset model-implied fan over trajectories
## (five facets, legend in the empty 6th cell). Only block-level A/B/C labels.
library(cowplot)
pal <- unname(WORDBANK_PALETTE)
col_scn <- c("Pure (κ=1)" = "grey55", "Baseline" = "#222222",
             "↑ Efficiency" = "#0072B2", "↑ Acceleration" = "#D55E00")
lty_scn <- c("pure" = "21", "accel" = "solid")

## Block B: conceptual theta -> CDI mechanism.
## Built as TWO single-facet plots stitched side by side rather than one two-facet
## facet_wrap. facet_wrap shares one y-axis title across its panels, but these panels
## are in different units (logits vs proportion) and each needs its own. Keeping a real
## facet_wrap(~quantity) inside each plot preserves the bold strip headers and the
## grouped appearance; patchwork only supplies the independent y titles.
## [MCF: the two y-axis titles below are agent-set -- reword if you'd prefer others.]
## NB: name this `qsel`, NOT `q` -- fig1_fan$conceptual carries a column called `q`
## (pivot_longer names_to), and dplyr data-masking would resolve `quantity == q`
## against that column instead of the local variable, silently yielding zero rows.
## Display-only short strip titles. At 7.5in the two block-B panels are ~1.4in wide and
## the cached labels ("Latent Ability (θ)", "Words Produced (CDI)") are wider than that,
## so they were being clipped mid-word. The y-axis titles already say which units each
## panel is in, so the strips can be terse. [MCF: reword if you prefer.]
STRIP_B <- c("Latent ability", "Words produced")
panelB <- function(i, ylab, ylim = NULL) {
  qsel <- levels(fig1_fan$conceptual$quantity)[i]
  d <- dplyr::filter(fig1_fan$conceptual, quantity == .env$qsel) |>
    dplyr::mutate(quantity = factor(STRIP_B[i], levels = STRIP_B[i]))
  ggplot(d, aes(age, value, color = scenario, linetype = kind, group = scenario)) +
    (if (is.null(ylim)) NULL else
       geom_blank(data = data.frame(age = 9, value = ylim),
                  aes(age, value), inherit.aes = FALSE)) +
    geom_line(linewidth = 0.8) +
    ## Direct labels on the LEFT panel only. The two panels share a colour scale, and in
    ## words-produced space the accelerating curves converge near 1.0, so their end labels
    ## collided there however they were repelled. Labelling once keeps the right panel clean.
    (if (i != 1) NULL else ggrepel::geom_text_repel(
      data = dplyr::filter(fig1_fan$conceptual_lab, quantity == .env$qsel) |>
              dplyr::mutate(quantity = factor(STRIP_B[i], levels = STRIP_B[i])),
      aes(label = scenario),
      size = 2.7, hjust = 0, direction = "y", nudge_x = 0.8,
      segment.size = 0.2, xlim = c(36, NA), box.padding = 0.3, min.segment.length = 0)) +
    facet_wrap(~ quantity) +
    scale_color_manual(values = col_scn, guide = "none") +
    scale_linetype_manual(values = lty_scn, guide = "none") +
    ## lower limit 8 to match the DATA range in block C (Marchman and Norwegian both
    ## start at 8), so B and C share an x-origin. Curve data now spans 8-36 (cache).
    scale_x_continuous(breaks = c(12, 24, 36)) +
    coord_cartesian(xlim = c(8, if (i == 1) 49 else 38), clip = "off") +
    labs(x = "Age (Months)", y = ylab) +
    theme_minimal(base_size = 11) +
    theme(strip.text = element_text(face = "bold", size = 11),
          panel.grid.minor = element_blank(),
          plot.margin = margin(16, 10, 4, 6))
}
## left = latent ability (free y); right = CDI proportion pinned to the full [0, 1]
pB <- panelB(1, "Latent ability (logits)") |
      panelB(2, "Proportion of items produced", ylim = c(0, 1))

## Block C: per-dataset fits (one faceted plot; legend flush in the empty 6th cell)
pC <- ggplot() +
  geom_line(data = fig1_fan$spag, aes(age, prop, group = interaction(lang, child)),
            color = "grey30", alpha = 0.14, linewidth = 0.22) +
  geom_line(data = fig1_fan$fan, aes(age, vocab, color = qf, group = qf,
            linewidth = ifelse(qf == "50th", 0.9, 0.5))) +
  facet_wrap(~ lang, scales = "free_x", nrow = 2) +
  scale_color_manual(values = pal, name = "Child\nPercentile", drop = FALSE) +
  scale_linewidth_identity() +
  scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1)) +
  labs(x = "Age (Months)", y = "Proportion of Items Produced") +
  guides(color = guide_legend(override.aes = list(linewidth = 1.1))) +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "bold", size = 11),
        panel.grid.minor = element_blank(), panel.spacing = unit(1, "lines"),
        plot.margin = margin(16, 8, 4, 6),
        legend.position = c(0.70, 0.26), legend.justification = c(0, 0.5),
        legend.key.width = unit(15, "pt"), legend.title = element_text(size = 10.5),
        legend.background = element_rect(fill = "white", color = NA))

## Block A: graphical model (title-less asset built from paper/mockups/fig1_gm_notitle.tex)
gm <- cowplot::ggdraw() +
  cowplot::draw_grob(grid::rasterGrob(png::readPNG(here("paper", "mockups", "fig1_gm_notitle.png")),
                                      interpolate = TRUE, width = grid::unit(0.98, "npc")))

## compose: [A | B B] over [C C C], only block-level A/B/C labels
## rel_widths: block A is a RASTER image whose internal text does not scale with
## base_size, so at 1/3 of a 7.5in row it became unreadable. Give it a larger share.
top <- cowplot::plot_grid(gm, pB, ncol = 2, rel_widths = c(1, 1.55),
                          labels = c("A", "B"), label_size = 13, vjust = 1.1)
cowplot::plot_grid(top, pC, ncol = 1, rel_heights = c(1, 1.85),
                   labels = c("", "C"), label_size = 13, vjust = 1.1)
```


# Growth in ability leads to rapid changes in learning times for individual words

```{r fig-efficiency, fig.width=7, fig.height=5}
#| label: fig-efficiency
#| fig-cap: "Estimated cumulative exposures to a word plotted by the age at which 50% of the children in our three English-language datasets were estimated to produce the word. Colors show different word classes, and lines show simple linear fits."

items <- fig2$items
k <- 8
.vis <- items |> filter(N_word >= 50, N_word <= 1e6, t_50 >= 10, t_50 <= 42)
labels_df <- bind_rows(
  slice_min(.vis, t_50,   n = k),   # far left  (earliest)
  slice_max(.vis, t_50,   n = k),   # far right (latest)
  slice_min(.vis, N_word, n = k),   # bottom    (fewest exposures)
  slice_max(.vis, N_word, n = k)) |># top       (most exposures)
  distinct(item, .keep_all = TRUE)
ggplot(items, aes(t_50, N_word, colour = lexical_class)) +
  geom_point(alpha = 0.5, size = 1.1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8, formula = y ~ x) +
  ggrepel::geom_text_repel(data = labels_df,
                           aes(label = sub("[ ]*\\(.*\\)$", "", word)), size = 2.7,
                           max.overlaps = Inf, min.segment.length = 0,
                           segment.size = 0.2, show.legend = FALSE) +
  scale_y_log10(labels = scales::label_number(big.mark = ","),
                breaks = c(100, 1000, 10000, 100000),
                limits = c(50, 1000000)) +
  scale_x_continuous(limits = c(10, 42)) +
  scale_colour_brewer(palette = "Dark2", name = NULL,
                      labels = c("Nouns", "Action Words", "Descriptive Words", "Function Words")) +
  labs(x = "Age (months) when 50% of children produce word",
       y = "Cumulative exposures at age of acquisition") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

The accelerating accumulator model captures a salient fact about language acquisition: the number of exposures necessary to learn a word decreases dramatically over childhood. A toddler may have heard a word like "book" thousands of times before they say it; a few years later a preschooler may hear "ibex" a handful of times on a trip to the zoo and repeat it the next day. 

To quantify this phenomenon, we combined our fitted model with corpus frequencies to compute the average proportion of exposures a child has likely heard for a newly acquired word at different ages. Focusing on English, the language for which we had the most data, we computed each word's age of acquisition (AoA) $A$ from our fitted models, with AoA defined as the point at which 50% of children produce the word -- in other words, the age at which child ability and word difficulty balance: $A(j) = a_0 \exp\!\big((\delta_j - \log H - \mu_\xi)/\kappa\big)$. Then for each word, we made an estimate of the cumulative word frequency for an individual child based on frequency estimates from the CHILDES archive, a large set of transcripts of speech to children [@macwhinney2000;@sanchez2019].

@fig-efficiency shows the relationship between cumulative exposures and age of acquisition for the accelerating accumulator. In contrast, the fitted pure accumulator predicts a far shallower decline, and cannot reproduce the observed range of acquisition ages (see SI: Age of acquisition in the accelerator model). Instead, within each word class (e.g., nouns, verbs, and function words), the number of examples needed for learning drops exponentially with development.^[Indeed, this analysis likely understates the degree of change in efficiency as it assumes that each exposure to a word is equally meaningful. In fact, individual exposures vary widely in how much information they give. Sometimes words are overheard rather than directed to the child, or lack physical context to allow for clear inference [@cartmill2013;@trueswell2013;@mollica2017].] Thus, in line with both intuition and experiments, the total number of exposures required on average to learn a new noun might indeed be only a small handful by age three [@markson1997;@bloom2002].

# Language models show no acceleration in their vocabulary growth

```{r fig-llm-acceleration, fig.width=7.5, fig.height=6.6}
#| label: fig-llm-acceleration
#| fig-cap: "(A) Density of the acceleration exponent $\\kappa$ for both children (per child, English and Norwegian) and LMs (by word). (B) Variation in acceleration (kappa) across children and LMs. (C) Fractional return on training (the ratio of decrease in loss to excess loss) for both children and LMs. Blue lines show model trajectories; red lines show fitted trajectories from children. Three children are highlighted in dark red with horizontal uncertainty bands indicating +/- 1 SD in calibration of their input."
## LAYOUT/FONT SIZING. out.width="100%" scales every figure to the 6.5in text width, so
## the on-page font size is base_size * 6.5/fig.width. At the old fig.width=14 that made
## base_size 11 render at 5.1pt. Dropping to 7.5 (scale 0.87) puts base_size 11 at ~9.5pt
## and inline annotations at ~9pt once their sizes are bumped to match (geom_text size is
## in mm: rendered_pt = size * 2.845 * 0.87, so size 3.6 ~ 9pt).
## Three panels no longer fit side by side at that width, so A and B share the top row and
## C -- a horizontal dotplot with five population rows that benefits from width -- spans
## the bottom. Panel letters keep their A/B/C order so the text references still hold.

# Data cached by paper/build_cache.R §6 + build_cache_short.R §8:
#   sl   -- per-instance kappa (children by child [_a3 M3 BLUPs], LMs by word);
#           C&B architectures dropped here (SI: Acceleration in Other Architectures)
#   f3   -- fractional-return curves (panel B)
#   vy   -- per-learner developmental kappa for the variability panel (panel C):
#           children (BLUPs) vs one point per LM training run (CHILDES / BabyLM /
#           ClimbMix ladders)
sl <- llm_slopes$slopes |>
  dplyr::filter(group != "LMs: C&B 2022 (4 architectures)") |> droplevels()
f3 <- llm_slopes$fig3; scaling_par <- llm_slopes$scaling_par
vy <- llm_slopes$vary
col_child <- "#c41e37"; col_ex <- "#7a0e1f"; col_lm <- "#2c7fb8"
slope_pal <- c("Children (English)"="#c41e37","Children (Norwegian)"="#fb9a99",
               "LMs: CHILDES (training)"="#2c7fb8","LMs: CHILDES (development)"="#41ab5d")

## Panel A: acceleration (kappa) densities -- children high & variable, LMs near 1
pA <- ggplot(sl, aes(slope_natural, fill=group, colour=group)) +
  geom_density(alpha=0.32, linewidth=0.5) +
  geom_vline(xintercept=1, linetype="dashed", colour="grey55", linewidth=0.4) +
  annotate("text", x=1.6, y=0.92, hjust=0, vjust=1, size=3.6, colour="grey45",
           label="kappa = 1\nunit accumulator") +
  scale_fill_manual(values=slope_pal, name=NULL) + scale_colour_manual(values=slope_pal, name=NULL) +
  coord_cartesian(xlim=c(0,22)) +
  labs(subtitle="Acceleration: children vs. LMs",
       x="kappa: slope on log(experience)", y="density") +
  theme_minimal(base_size=11) +
  theme(legend.position=c(0.66,0.62), legend.text=element_text(size=8.5), legend.key.height=unit(11,"pt"),
        legend.background=element_rect(fill="white",colour=NA), plot.subtitle=element_text(face="bold"))

## Panel B: fractional return gamma = return / excess loss, at matched absolute input.
## LMs: gamma = beta (flat). Children: gamma_i(t) -> kappa_i. All eligible English
## children (wash) + 3 exemplars (low/med/high kappa) with +/-1 SD input-rate ribbons.
n_child <- dplyr::n_distinct(f3$child_bg$ckey)
ex_lab <- f3$child_ex |> group_by(ckey) |> slice_max(tokens, n=1) |> ungroup() |>
  mutate(lab=sprintf("κ=%.0f", kappa))
pB <- ggplot() +
  geom_hline(yintercept=1, linetype="dashed", colour="grey55", linewidth=0.4) +
  geom_line(data=f3$child_bg, aes(tokens, gamma, group=ckey), colour=col_child, alpha=0.04, linewidth=0.35) +
  geom_line(data=f3$lm_gamma, aes(words, gamma, group=seed), colour=col_lm, alpha=0.5, linewidth=0.5) +
  geom_ribbon(data=f3$child_ex, aes(y=gamma, xmin=lo, xmax=hi, group=ckey),
              fill=col_ex, alpha=0.15, orientation="y") +
  geom_line(data=f3$child_ex, aes(tokens, gamma, group=ckey), colour=col_ex, linewidth=0.85) +
  ggrepel::geom_text_repel(data=ex_lab, aes(tokens, gamma, label=lab), colour=col_ex, size=3.4,
                           nudge_x=0.12, direction="y", segment.size=0.2, min.segment.length=0) +
  annotate("text", x=3.3e6, y=0.60, hjust=0, size=3.5, colour=col_lm, fontface="italic", label="LMs (10 seeds)") +
  annotate("text", x=3.3e6, y=1.22, hjust=0, size=3.2, colour="grey45", label="unit accumulator") +
  annotate("text", x=3.3e6, y=16, hjust=0, size=3.5, colour=col_child, fontface="italic",
           label=sprintf("all children (n=%d);\n3 exemplars ± input SD", n_child)) +
  scale_x_log10(breaks=c(3e6,5e6,1e7,2.4e7), labels=c("3M","5M","10M","24M")) +
  scale_y_log10(breaks=c(0.2,0.3,0.5,1,2,5,10,20)) +
  coord_cartesian(xlim=c(3e6,2.6e7), ylim=c(0.2,25)) +
  labs(subtitle="Fractional return across cumulative input",
       x="cumulative word tokens of input", y="fractional return  γ") +
  theme_minimal(base_size=11) +
  theme(plot.subtitle=element_text(face="bold"), panel.grid.minor=element_blank())

## Panel C: between-learner VARIABILITY in developmental kappa. Each point = one
## learner: a child (fitted kappa_i) or one LM training run (per-ladder median
## per-word kappa; CHILDES = 10 seeds, BabyLM / ClimbMix = 3 disjoint subsets x
## 8 seeds each -- the composition control). CV annotations quantify dispersion.
vpal <- c("Children (English)"="#c41e37","Children (Norwegian)"="#fb9a99",
          "LMs: CHILDES"="#2c7fb8","LMs: BabyLM"="#238b45","LMs: ClimbMix"="#e6550d")
## Use the cached bootstrap CIs (build_cache_short.R) rather than recomputing the bare
## point estimate: a CV from 8-10 seeds, with both mean and variance estimated from fitted
## kappa, is unstable, so the interval is the honest summary. Resampled over learners
## within population. Kept terse -- this panel is now half-width.
## Labels are right-aligned in a gutter beyond every row's data: the children's clouds
## reach ~55 while the LM rows stop near 3, so the axis is extended to leave clear space
## on the right. A log axis absorbs the extra decade cheaply, and one shared alignment
## reads better than five ragged ones.
lab_cv <- llm_slopes$vary_cv |>
  mutate(lab = sprintf("%.0f%% [%.0f-%.0f]", cv, cv_lo, cv_hi))
# CV is computed on ALL learners; the log axis display drops the handful of
# children whose posterior-median kappa_i is non-positive (8 of 1745).
pC <- ggplot(vy |> dplyr::filter(kappa > 0),
             aes(kappa, forcats::fct_rev(population), colour=population)) +
  geom_jitter(position=position_jitter(height=0.22, seed=1), alpha=0.25, size=0.9) +
  stat_summary(fun=median, geom="crossbar", width=0.5, linewidth=0.5, colour="grey20") +
  geom_text(data=lab_cv, aes(x=380, y=forcats::fct_rev(population), label=lab),
            inherit.aes=FALSE, size=3.0, hjust=1, colour="grey30") +
  geom_vline(xintercept=1, linetype="dashed", colour="grey55", linewidth=0.4) +
  scale_x_log10(breaks=c(0.5,1,2,5,10,20,50)) +
  scale_colour_manual(values=vpal, guide="none") +
  coord_cartesian(xlim=c(0.4,400)) +
  labs(subtitle="Variability in acceleration",
       x="developmental kappa (log scale)", y=NULL) +
  theme_minimal(base_size=11) +
  theme(plot.subtitle=element_text(face="bold"), panel.grid.minor=element_blank())

## Fractional return takes the full-width bottom row, per MCF: it is the paper's most
## original child-LM comparison, so it gets the space. The kappa density and the
## learner-variability dotplot share the top row.
##
## [MCF: panel LETTERS follow POSITION, keeping the usual A-B-C reading order, so the
##  content behind B and C has swapped. The fig-cap above still describes the OLD order
##  and needs its (B) and (C) clauses exchanged -- a corrected version is drafted in the
##  fenced block just below this chunk. Also worth grepping the main text for any
##  reference to Fig 3 panels by letter.]
cowplot::plot_grid(
  cowplot::plot_grid(pA, pC, nrow = 1, labels = c("A", "B"), label_size = 12),
  cowplot::plot_grid(pB, nrow = 1, labels = "C", label_size = 12),
  nrow = 2, rel_heights = c(1, 0.82))
```

Do LMs show the same signatures of acceleration and variability as children? To evaluate this question, we begin by building a bridge between the well-known "scaling laws" for large language models [@hoffmann2022;@kaplan2020] and the model we describe here for vocabulary growth. The form of one popular scaling law is that the loss $L$ of a neural network language model on a dataset size $D$ is $L = E + \frac{B}{D^\beta}$ (holding the size of the the network constant), where $B$ and $\beta$ are the free parameters of a power law and $E$ is the entropy of the text being modeled. This power law scaling behavior can emerge from a set of independently learned "quanta" that are themselves power-law distributed [@michaud2024]; the words of natural language follow such a power law and are potential quanta of this type [@zipf1949]. 

Following this idea, we explore the emergence of individual words in neural language models. If their emergence indeed follows power-law dynamics, then it would follow that their accumulation *decelerates* rather than accelerating (as we have shown in children). In particular, each further order of magnitude of data should produce a smaller gain in word knowledge. To trace the emergence of individual words in language models, we fit sigmoidal curves to the trajectory of surprisal values (negative log probability of individual words) across training [following @chang2022]. These sigmoids provide a link to the logistic curves in the Rasch model we used with children. The slope of the sigmoid for each word is analogous to our acceleration parameter $\kappa$ -- it captures how fast evidence for that particular word accumulates with training data (see SI: Computing Acceleration in LMs). By computing the distribution of these slopes, we can compare the average speed of evidence accumulation in LMs to that of children.

We used GPT-2 for this analysis [@radford2019] as it has become the preferred model system for such investigations [@feng2024;@misra2024;@evanson2023;@choshen2022] (see SI: Acceleration in Other Architectures). Matching training data between models and children is critical for a valid comparison. Thus, we trained models on a corpus of 24M words of English child-directed speech from the CHILDES archive [@macwhinney2000;@feng2024;@huebner2021], approximating a three-year-old child's total language input. We first computed per-word surprisal across training checkpoints. $\kappa$ estimates from these clustered around 1 (see @fig-llm-acceleration A): knowledge about individual words accumulated gradually.
<!-- In contrast, across all five of our samples, children's $\kappa$ values were both high and highly variable.  -->

This analysis treats learning as the result of multiple training passes through the same dataset, however, an approach that is fundamentally different from learning in children. Word surprisal changes as a function of total training data experienced (rather than total passes through the same dataset) provide a better analogue. We thus trained "developmental" sequences of GPT-2 models. Each model in a sequence was trained to convergence on an increasingly larger subsample of the CHILDES data, ranging from .5M words to 24M words, and we trained 10 such sequences with different random seeds. $\kappa$ values computed across these runs also showed no evidence of acceleration.

Next, to make a direct comparison between the growth of individual models (rather than words) and individual children, we computed learners' fractional return on receiving new data. Scaling law analyses describe the change in a model's loss $L$ with respect to the fundamental entropy $E$ of the training data. These can be manipulated to compute a *fractional return* on training data, describing the total amount of remaining possible loss that has been removed by a particular amount of data: $\gamma(D) = \frac{-,dL/d\ln D}{L - E}$. To compute this quantity in children, we computed the child's "loss" as the summed surprisal across all words in the child's vocabulary and took the derivative across time; this revealed a function that converges to $\kappa$. We aligned these to children's estimated input rate over time (see SI: Input Estimation). As predicted by neural scaling laws, LMs were precisely constant (flat) in their fractional returns, while individual children's trajectories showed positive acceleration (@fig-llm-acceleration C).

Children were highly variable in their trajectories, but we observed almost no variation in acceleration between model runs with different random seeds. We investigated whether this lower level of variability is a fundamental feature of standard LM training regimes or instead reflects the use of a single training dataset that was shared across all models. Because sufficient child-directed training data was not available, we trained new developmental sequences of models on disjoint samples of either the BabyLM mix (a varied dataset designed to mimic the varied registers heard by children) or a standard LM training dataset (ClimbMix). Across all datasets, cross-model variation in acceleration was far more limited than for children  (@fig-llm-acceleration B).


# Discussion

How does children's vocabulary grow over early childhood? We proposed that this process is characterized by accelerating accumulation, in which variation between children is produced by differences in the efficiency of the accumulation process and the rate of acceleration. This cognitive process leads to a simple growth model that nevertheless provides a good description of cross-linguistic longitudinal data. In contrast to children, LMs follow neural scaling laws, meaning that they decelerate across training rather than accelerating. This difference in returns offers one candidate contributor to the data gap: children increasingly capitalize on each additional unit of experience, whereas conventional LMs do not.

Where does the acceleration in children's learning come from? Some may come from developmental maturation of general abilities such as learning, memory, or attention. In the language of our model, these changes would mean that children's "buckets" require fewer drops to fill them. However, the degree of acceleration we show means that word learning becomes more efficient by several orders of magnitude, seemingly beyond measured changes in domain general cognitive processes [@hartshorn1998;@bauer2000]. 

A second potential source of acceleration is that children "learn to learn": their increasingly sophisticated inferential abilities allow them to extract progressively more information from the linguistic signal [@smith2002;@markman1988;@bohn2019;@meylan2022]. They learn "overhypotheses" that guide their inferences based on general regularities in how their language works -- for example, that many nouns are organized by shape [@kemp2007;@smith2002]. They also use the language they already know in the moment to extract more information from new utterances -- leveraging word meanings [@markman1988;@clark1988], syntactic knowledge [@gleitman1990], and pragmatic inference [@bohn2019] to make sophisticated inferences about gaps in their knowledge [@frank2009;@bloom2002]. Indeed, formally, these sorts of "leverage" mechanisms can change the slope of acceleration in accumulator models [though they cannot create it; @mitchell2009].

The accelerating accumulator is a parsimonious, low-dimensional description of variation in children's vocabulary growth that accords with the structure of the learning problem faced by children. But our evidence is based only on statistical fit and model comparison, rather than focused causal study. Further work is required to test the accumulator as a mechanistic account. 

In sum, our study suggests that the presence of acceleration is a fundamental difference between children and LMs that may be responsible for the "data gap" between the two. Characterizing the nature of this acceleration and understanding the circumstances under which it can arise in artificial systems should thus be a focus for future investigation.

# Methods

## Data and code

Reproducible code for this paper is available at \href{}{http://github.com/langcog/standard-model-2}. Data were retrieved from Wordbank using the `wordbankr` package [@frank2017] and CHILDES using the `childesr` package [@sanchez2019].

All longitudinal datasets with three or more observations per child were included. We applied a quality filter to exclude administrations that showed either decreases of more than 25% of productive vocabulary or increases that exceeded 40% of total vocabulary per month. This excluded `r n_excl` children in total (`r pct_excl`%), nearly all from the English (Marchman) dataset (`r qc_mar`%; Norwegian `r qc_nor`%, none from the others) (See SI: Data Exclusions). 

## Bayesian models

Bayesian models were fit separately to each dataset using Stan [@stan] using four independent chains, per the specification described in the text. The model was parameterized to capture individual children's variation from the population mean such that $\xi_i = \mu_\xi + a_i$ and $\kappa_i = \mu_\kappa + b_i$, where $(a_i, b_i) \sim \mathrm{MVN}(\mathbf 0, \Sigma)$. To facilitate interpretation of the accumulator mechanism, we included the constant $\log H = \log(365) \approx 5.90$ (waking hours/month) and the anchor age $a_0 = 18$ months (so $\log(t/a_0)=0$ at 18 months).

We chose weakly informative priors throughout, including on the efficiency and acceleration parameters $\mu_\xi \sim \mathcal N(-6,\ 5)$ and $\mu_\kappa \sim 1 +  \mathcal N(0,\ 5)$ (centered around the pure accumulator model) as well as the individual child standard deviations $\sigma_a \sim \mathrm{Half\text{-}Normal}(0,\ 3)$ and 
$\sigma_b \sim \mathrm{Half\text{-}Normal}(0,\ 5)$ and the item difficulties $\tau_\delta \sim \mathrm{Half\text{-}Normal}(0,\ 5)$. We also chose a prior that mildly regularized the correlations between $a_i$ and $b_i$ ($\mathbf L \sim \mathrm{LKJ}(2)$).

## Learning efficiency

Age of acquisition for each word was computed based on the weighted average of word difficulties $\delta_j$ from the three fitted M3 models for English. For the population level of $\xi$ and $\kappa$, we computed the earliest age at which there was at least a .5 probability of producing the word. Average exposures per month to each word were computed as $N_j = R \cdot t_{50}(j) \cdot p_j$, where $R$ is the rate of input (words/month), $t_{50}(j)$ is the word level age of acquisition computed above, and $p_j$ is the word's corpus probability. We used average input estimates derived from @sperry2019 and @hart1995 for our estimate of $R$ (see SI: Input Estimation), and computed $p_j$ from empirical word frequencies in all North American English corpora in childes-db release 2021.1. 

## Language model training

Following @feng2024, we fit randomly-initialized \texttt{GPT-2-small} (124M parameters) models, using the 24M word (~47M token) preparation of English-language CHILDES data from that study. We also fit models to the 2nd BabyLM challenge dataset (downloaded from [osf.io/ad7qg](), with CHILDES data removed) [@hu2024] and the ClimbMix dataset (available at [https://huggingface.co/datasets/karpathy/climbmix-400b-shuffle]()) [@diao2025]. We constructed three disjoint subsets of 24M words from each.

For comparability, all data were tokenized using a standard BPE tokenizer trained on the full CHILDES dataset (all 24M). Surprisal was computed as mean per-word negative log likelihood over ~50 held-out validation contexts (last 128 tokens each) for the 609 CDI words that were used in @chang2022 and that appeared in the training data with sufficient frequency in the CHILDES held-out contexts. Surprisal values for sigmoid fits were read out from the epoch with lowest validation loss.

For the computation of surprisal across training runs ("training"), we trained models using 3 random seeds on the full dataset. For the computation across different dataset sizes ("development"), the data were segmented into 18 nested datasets ranging in size from .5M to the full 24M. We then chose 10 random seed values and trained 10 models on each dataset for a total of 180 models. For the computation of variability across model runs ("composition"), we ran 8 random seed values across 12 nested dataset sizes for three disjoint subsets of both BabyLM and ClimbMix, resulting in 8 (seeds) x 12 (sizes) x 3 (subsets) x 2 (datasets) = 576 distinct models.

Models were trained for a total of 20 epochs with sequence length 1024 and batch size 8, using AdamW for optimization and learning rate of 1e-4 with a linear decay, no warmup, weight decay of 0, and no in-epoch shuffling. Training was performed on NVIDIA A40 GPUs. All trained models are available at [https://huggingface.co/mcxfrank/childes-gpt2-ladder]() and [https://huggingface.co/mcxfrank/gpt2-composition-control]().

## Fractional return on input

Fractional return on input was computed for models based on mean surprisal $L(D)$ of the CDI words after training on
$D$ word tokens of input. This follows the form from @hoffmann2022 of $L(D) = E + B\,D^{-\beta}$, with $E$ the irreducible entropy floor and $B, \beta$ fit across budgets (here $E \approx 2.94$ nats, $B \approx 312$, $\beta \approx 0.32$,
$R^2 = 0.998$). The excess loss $L(D) - E = B\,D^{-\beta}$ is the reducible
part that training removes; it falls as a power law of slope $-\beta$. Marginal return is the loss removed per $e$-fold of data, 

$$R(D) = -\frac{\mathrm{d}L}{\mathrm{d}\ln D} = \beta\,\big(L(D) - E\big),$$

\noindent obtained by differentiating the scaling law. We compute the ratio of this quantity to the excess loss, obtaining the fractional return: the share of the remaining reducible loss removed
per $e$-fold, $\gamma_{\text{LM}}(D) = \frac{R(D)}{L(D) - E} = \beta.$
For a power law, this is constant: $\gamma_{\text{LM}} = \beta \approx 0.32$
at every budget, recapitulating the scaling law.

We can compute the same quantity for children. We define a vocabulary production
loss, $L_i(t) = \operatorname{mean}_j[-\log p_{ij}(t)]$, whose floor is zero
(a known word is produced, $p \to 1$). Its marginal return is
$R_i(t) = -\mathrm{d}L_i/\mathrm{d}\log t = \kappa_i\,\operatorname{mean}_j(1 - p_{ij})$,
so the fractional return is
$$\gamma_i(t) = \frac{R_i(t)}{L_i(t)}
 = \kappa_i\,\frac{\operatorname{mean}_j(1 - p_{ij})}{\operatorname{mean}_j(-\log p_{ij})},$$
which rises toward the child's own $\kappa_i$ as the vocabulary saturates.

# Acknowledgements

Some of the computing for this project was performed on the Sherlock and Marlowe clusters. We would like to thank Stanford University and Stanford Research Computing for providing computational resources and support that contributed to these research results. Funds for computational experiments were also provided by a gift from Meta to support research on children's home language environment. Claude Code was used for literature review, code generation, visualization, and simulation infrastructure. 

# References

::: {#refs}
:::


<!-- :::{.suppinfo} -->


<!-- ::: -->

```{=latex}
\newpage
\begin{center}
{\LARGE\bfseries Supplemental Information}
\end{center}
```

{{< include supplemental.qmd >}}