---
author:
  - name: Michael C. Frank
    affiliations: 
      - name: Department of Psychology, Stanford University

title: Efficiency and acceleration in the accumulation of children’s early vocabulary
title-short: Efficiency and acceleration 
keywords: Early word learning; Bayesian data analysis; Large language models
abstract: |
  \noindent Children learn hundreds of words over the first few years of their life. This process begins slowly but increases until only a few exposures to a word are necessary for a child to learn it. Yet there are also wide variations between children, and the distribution across children has not been accurately characterized. Here we present a statistical model describing this process as accelerating accumulation, with children varying in both the slope and intercept of their accumulation. Using data from three languages, this model provides a good characterization of longitudinal growth and provides insights into the sources of age and socioeconomic variation. Using dense observations of children's early language environment, the model also can be used to estimate how much variation between children is due to differences in language input, finding estimates that converge with recent meta-analyses. In contrast to children, large language models are not well-described by this model: they show standard accumulator dynamics and very limited variability. 

bibliography: standard_model.bib
format:
  pdf:
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
    linestretch: 1.05
    colorlinks: true
    keep-tex: true
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(here); library(dplyr); library(tidyr); library(readr)
  library(ggplot2); library(patchwork); library(latex2exp)
  library(knitr); library(scales); library(stringr); library(forcats)
  library(broom); library(png); library(grid); library(gridExtra)
  library(ggrepel)
})
source(here("paper", "_helpers.R"))

# load cached data — produced by `Rscript paper/build_cache.R`
# (Table 1 is built inline in its own chunk, not cached.)
CACHE <- here("paper", "cache")
ladder     <- readRDS(file.path(CACHE, "fig2_glmer_ladder.rds"))
demo_blups <- readRDS(file.path(CACHE, "blups_demographics.rds"))
io_summary <- readRDS(file.path(CACHE, "fig5_io_summary.rds"))
aic        <- readRDS(file.path(CACHE, "aic_summary.rds"))
exposure   <- readRDS(file.path(CACHE, "fig4_exposure.rds"))

# Inline range: how much worse the best (D) model becomes when age enters
# linearly rather than as log-age, across the four languages (dAIC of
# D_lin relative to the best model, D_log).
.dlin <- aic |> filter(model == "D_lin") |> pull(dAIC)
aic_loglin_range <- sprintf("%s--%s",
  format(round(min(.dlin)), big.mark = ","),
  format(round(max(.dlin)), big.mark = ","))

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE, cache=TRUE,
  fig.align = "center", out.width = "100%", dpi = 300
)
```



Children’s early vocabulary grows very rapidly during early childhood, with large variation between children [@fenson1994;@fenson2007;@frank2021]. The growth has theoretical, practical, and methodological importance. From a theoretical perspective, growth mechanisms shed light on children’s fundamental learning mechanisms and how they change across childhood [@mcmurray2007;@bloom2002;@frank2021]. From a practical perspective, early vocabulary is an important precursor -- and predictor -- of academic achievement [@marchman2008;CITES]. From a methodological perspective, vocabulary also affords a unique psychometric opportunity to study human variation because, unlike nearly all important human psychological metrics, there is a true zero point -- a time before which children know no words. Thus, absolute levels of variability can be calculated. 

One dominant viewpoint is that children's vocabulary growth can be characterized as a process of accumulation [@mcmurray2007; @mitchell2009; @hidaka2013; @mollica2017; @kachergis2022]. Each distinct word (type) can be conceptualized as a "bucket," and every token of language input then is a "drop" that falls into its respective bucket. When each bucket is full, the word is learned [@mcmurray2007;@kachergis2022]. These models have been used to recover interpretable differences in "bucket size," often based on word properties including concreteness and syntactic category [@hidaka2013;@kachergis2022].

The accumulator view connects with a viewpoint in which the quantity of language that children hear plays a strong causal role in the speed of their learning [@fernald2012]. Indeed, there is systematic evidence that variation in children's input is related to variation in vocabulary [@fernald2012;@weisleder2013;CITES]. Studies of input have identified systematic variation in *quantity* in terms of sheer amount of tokens as well as variation in *quality*. Measuring what makes input high quality has been challenging, but higher quality input likely has higher lexical diversity, longer sentences, more declarative and interrogative content relative to imperatives, and more elaborated discourse [LOTS OF CITES]. 

A naïve interpretation of the accumulator view would suggest that most of the variance in vocabulary growth should be related to input. Two recent meta-analyses complicate this view, however. @anderson2021 showed that only a modest proportion of the variance in vocabulary outcomes is related to input quantity (XYZ) and quality (XYZ). A newer meta-analysis of quantity specifically confirms and extends the proportion of variance due to quantity to 4--7% of total variation in vocabulary [@coffey2026]. Thus -- as we might have supposed based on other research on human variation -- not all variation in vocabulary between children can be attributed to input differences. 

As this discussion indicates, despite the promise of the accumulator viewpoint, it has not been used successfully to understand variation between individuals. Instead, the majority of applications have focused on variation between words. Further, the rapidity of children's early word learning stands as a rebuke to this view. Sometimes a small number of examples is enough to form a lasting representation [@markson1997;@woodward1994]. Perhaps only a small number of high-quality instances is sufficient for learning [@trueswell2013;@mollica2017], but if so, why do some words take much longer to learn?

The goal of this work is to address these puzzles through systematic investigation and extension of accumulator models. Following previous work, we consider that children's learning ability might accelerate across development [@hidaka2013]; in the bucket metaphor, the drops get bigger as children get older. We begin by extending the psychometric accumulator model of @kachergis2022: our key move is to model variation in individual children's learning efficiency and the acceleration of their learning. This extension defines a longitudinal growth model that can be fit to data, and that provides a simple reconciliation between slow accumulation and fast mapping: the number of examples of a word necessary for learning declines by multiple orders of magnitude from age one to age three. 

To constrain our model, we use data from the MacArthur-Bates Communicative Development Inventories [MB-CDI; @fenson1994;@marchman2021], a popular parent report instrument in which parents are asked to indicate which words their child produces and understands. Here we focus exclusively on production, which is used with older children and which tends to be a more reliable measure [@frank2021;others]. We make use of longitudinal data from Wordbank [@frank2017], an open repository of data from the MB-CDI. 

To address questions about the relationship between language input and growth, we fit a Bayesian version of our growth model. We make use of new data resources to estimate the magnitude of linkages between input  and vocabulary growth. First, we estimate plausible population values based on independent estimates of population variation in language input [@coffey2024]. In addition, we use measures of individual children’s language input from egocentric videos [@long2024] and day-long home recordings [@egan2025;@adams2018;@fernald2013] to estimate the input-uptake relationship more precisely for a subsample of children.

Overall, our analysis supports a view in which children’s vocabulary growth is not a pure process of accumulation but rather one that accelerates rapidly over time. We end by comparing the fit of our models on human data to their fit on large-language models. Making use of GPT-2 models trained on human developmental data, we show that – in contrast to children – these models do appear to function as pure accumulators. This analysis exposes a critical disanalogy between LLMs and human learners and suggests a potential route for understanding the vast "data gap" between children and large language models [@frank2023;@warstadt2023]. 

# Model 


```{r fig-schematic, fig.width=7, fig.height=4.5}
#| label: fig-schematic
#| fig-cap: ""
# Schematic walk through the model ladder. Four columns are model
# variants (A → D). Top row: latent ability theta_i(t). Bottom row: the
# expected VOCABULARY curve, summed over a distribution of word
# difficulties — V(t) = mean_j plogis(theta_i(t) - delta_j). This is the
# McMurray (2007) construction: even a pure accumulator (A, kappa=1)
# produces an accelerating vocabulary curve because theta sweeps through
# the difficulty distribution. The SAME Gaussian difficulty distribution
# is used in every panel (the same words), so the columns differ only in
# the latent dynamics (kappa) and between-child variance (sigma_alpha,
# sigma_kappa), not in the word set. Five quantiles (10/25/50/75/90) in
# the Wordbank palette match Fig 2.
ages <- seq(8, 30, length.out = 80)
A0   <- 8                                  # anchor age: log(t/A0) >= 0
L    <- log(ages / A0)
qs   <- c(0.1, 0.25, 0.5, 0.75, 0.9)
z    <- qnorm(qs)

XI0     <- -2.0                            # median intercept (log r + log alpha)
KAPPA_A <- 2.5                             # acceleration for B/C/D
# shared word-difficulty distribution (Gaussian, McMurray 2007)
DELTA   <- qnorm(ppoints(150), mean = 0, sd = 1.5)
vocab_of <- function(th) vapply(th, function(x) mean(plogis(x - DELTA)), numeric(1))

# population parameters chosen to be illustrative, not estimated.
variants <- tibble::tribble(
  ~name,                          ~kappa_pop, ~sigma_alpha, ~sigma_kappa, ~eq,
  "(A) pure accumulator",         1,           0,            0,            r"($\theta = \log(r) + \log(t)$)",
  "(B) + acceleration",           KAPPA_A,     0,            0,            r"($\theta = \log(r) + \kappa\,\log(t)$)",
  "(C) + efficiency variance",    KAPPA_A,     1.1,          0,            r"($\theta = \log(\alpha_i) + \log(r) + \kappa\,\log(t)$)",
  "(D) + per-child acceleration", KAPPA_A,     1.1,          0.7,          r"($\theta = \log(\alpha_i) + \log(r) + \kappa_i\,\log(t)$)"
)

# For each model x quantile x age: theta and the summed vocabulary curve
sim <- variants |>
  mutate(name = factor(name, levels = name), panel_order = seq_along(name)) |>
  rowwise() |>
  do({
    v <- .
    bind_rows(lapply(seq_along(qs), function(i) {
      th <- XI0 + z[i] * v$sigma_alpha +
            (v$kappa_pop + z[i] * v$sigma_kappa) * L
      tibble(
        name = v$name, panel_order = v$panel_order,
        qprob = qs[i], qprob_f = factor(as.character(qs[i]),
                                        levels = c("0.1","0.25","0.5","0.75","0.9")),
        age   = ages, theta = th, vocab = vocab_of(th), eq = v$eq
      )
    }))
  }) |>
  ungroup()

# annotation table — one equation per panel
eq_lab <- variants |>
  mutate(name = factor(name, levels = name)) |>
  select(name, eq)

# vertical line size: median bold, others thinner
lwidth <- function(qf) ifelse(as.character(qf) == "0.5", 0.95, 0.55)

p_theta <- ggplot(sim,
                  aes(age, theta, color = qprob_f, group = qprob_f)) +
  geom_line(aes(linewidth = lwidth(qprob_f))) +
  geom_label(data = eq_lab,
             aes(x = -Inf, y = Inf,
                 label = latex2exp::TeX(eq, output = "character")),
             inherit.aes = FALSE, parse = TRUE,
             hjust = -0.05, vjust = 1.4, size = 2.4, label.size = 0,
             fill = scales::alpha("white", 0.8)) +
  facet_wrap(~ name, nrow = 1) +
  scale_color_manual(values = WORDBANK_PALETTE, name = "Percentile",
                     labels = c("10th", "25th", "50th", "75th", "90th"),
                     drop = FALSE) +
  scale_linewidth_identity() +
  coord_cartesian(ylim = c(-3.5, 4.5)) +
  labs(x = NULL, y = expression(theta[i](t))) +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(face = "bold", size = 9),
        legend.position = "bottom",
        panel.grid.minor = element_blank())

p_vocab <- ggplot(sim,
                  aes(age, vocab, color = qprob_f, group = qprob_f)) +
  geom_line(aes(linewidth = lwidth(qprob_f))) +
  facet_wrap(~ name, nrow = 1) +
  scale_color_manual(values = WORDBANK_PALETTE, name = "Percentile",
                     labels = c("10th", "25th", "50th", "75th", "90th"),
                     drop = FALSE) +
  scale_linewidth_identity() +
  scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1)) +
  labs(x = "age (months)", y = "expected vocabulary proportion") +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_blank(),  # suppress strip on lower row
        legend.position = "bottom",
        panel.grid.minor = element_blank())

(p_theta / p_vocab) + plot_layout(guides = "collect", heights = c(1, 1)) &
  theme(legend.position = "bottom")
```

We present a nested set of accumulator-type models, many of which have been presented in the literature. As we show below, each refinement of the model contributes to its ability to describe the variability between children. We start with the Rasch model from psychometrics [@rasch1960;@bond2013], which assumes that each child $i$ has a learning ability $\theta_i$ and each word $j$ has a difficulty $\delta_j$. The probability that child $i$ produces word $j$ is

$$P(w_{i,j}=1 \mid \theta_i, \delta_j) =\frac{\exp(\theta_i - \delta_j)}{1 + \exp(\theta_i - \delta_j)}$$
  
  
Following @kachergis2022, we can write our first model as a Rasch-style accumulator model (Model 1), specifying that $\theta_i$ is not a single parameter but instead a function of accumulation over time. In the simplest form: 
  
  $$\text{M1: }\theta_i(t)  = \log(r_i) + \log(t/a_0)$$
  
\noindent where $r_i$ is the child's rate of language input (in tokens per month) and $t$ is the child's age (in months, centered at $a_0$); note that we use log age to simulate linear accumulation because $\exp(\theta) = r \cdot t$. This form instantiates the basic hypothesis that the only thing that changes over time is more input, and the only source of variation between children is variation in rate. 

We compare this model to an accelerating accumulator model (Model 2), in which the amount learned from each exposure increases with time by an exponent $\kappa$.

$$\text{M2: }\theta_i(t)  = \log(r_i) + \kappa ~ \log(t/a_0)$$
\noindent When $\kappa>$, each successive month yields more accumulation than the last.

Neither of these models has any way to capture variation across children beyond variation in input rate. Thus, we augment the basic accelerating accumulator, rewriting it as the sum of either one or two individually variable quantities:

$$\text{M3: }\theta_i(t) = \xi_i + \kappa ~ \log(t/a_0)$$

$$\text{M4: }\theta_i(t) = \xi_i + \kappa_i ~ \log(t/a_0)$$

The intercept term in both models $\xi_i = \log(r_i) + \log(\alpha_i)$ can then decompose individual variation into both input rate-related and other variation in children's efficiency. Similarly, in M4, the slope term $\kappa_i = \kappa + \gamma \log r_i + \zeta_i$ decomposes acceleration into a population term, input-related acceleration, and a child‑specific residual acceleration $\zeta_i$ (the part not explained by input). The coefficient $\gamma$ then captures the strength of the relationship between input and acceleration. 

Model 4 instantiates the idea that children vary in their base rate of efficiency and their acceleration, instantiating  a longitudinal growth model in which children vary on their slope and intercept. Note that, while we define the relationship of $\xi_i$ and $\zeta_i$ to a child's input $r_i$, we can nevertheless fit the simplest form of the model by just allowing these parameters to vary as random child-level slopes and intercepts. 

As we will show, this model provides a good description of variation in children's longitudinal growth trajectories. @fig-schematic shows schematic predictions from a pure accumulator model (M1), an accelerating accumulator (M2) in comparison to models including child-level variation in efficiency (M3) and both efficiency and acceleration (M4). Each model's predictions are represented as ability scores ($\theta$) across age and then also projected into proportion correct scores on a discrete set of vocabulary items, representing an observed accuracy score that is subject to compression at floor and ceiling. 

A variety of previous work has focused on estimating the difficulty of individual words  [@goodman2008;@hidaka2013;@roy2015;@braginsky2019;@kachergis2022]. These analyses reveal that there is predictable word-level difficulty information but that only a modest portion of the variation between words is carried by frequency information. Words can be hard for a child to acquire because they vary also in the complexity and concreteness of their meanings, their syntactic category and syntactic context, their phonology, and many other factors. In the current work, we estimate word difficulty as a free parameter, and focus instead on variation between children. 

Our analysis generally depends on the availability of longitudinal data at the item level. The different models described above are not distinguishable from the cross-sectional, aggregate data that are typically considered in many analyses of early vocabulary. First, aggregate data do not allow for the estimation of individual word difficulties, thus the distribution of difficulties remains latent and different difficulty distributions can be posited to explain the accelerating shape of vocabulary growth [@mcmurray2007; @mitchell2009]. Second, cross-sectional data do not allow for accurate estimation of whether variation is due to differences in efficiency (intercept) or acceleration (slope) (see Supplemental Figure 1). 

# Results

## Children's longitudinal growth reflects differences in both efficiency and acceleration

We first retrieved longitudinal data on monolingual, typically-developing children's vocabulary from Wordbank [@frank2017], an open archive of data from the MacArthur-Bates Communicative Development Inventory and its international adaptations. @tbl-datasets shows the four longitudinal datasets in Wordbank that contained more than $>100$ children with multiple observations.  

```{r tbl-datasets}
#| label: tbl-datasets
#| tbl-cap: "Characteristics of the datasets."
# Wordbank rows are pulled from outputs/glmer_ladder/00_language_survey.csv;
# IO rows are hardcoded from each bundle's actual extracts. Citation column
# is a [@CITE] placeholder for the author to fill in.
# Table 1 is built inline (not cached) so the numbers and names are easy
# to edit here. Wordbank longitudinal rows are derived from the language
# survey CSV (outputs/glmer_ladder/00_language_survey.csv); input-observed
# (IO) rows are hardcoded from each dataset's extract. The Citation column
# holds [@CITE] placeholders for the author to fill in.

# --- Wordbank longitudinal rows (from the survey) ---
survey <- read_csv(
  here("outputs", "glmer_ladder", "00_language_survey.csv"),
  show_col_types = FALSE)

wb_rows <- survey |>
  filter(language %in% LANG_LABELS[PAPER_LANGS]) |>
  # keep the paper's language order
  mutate(language = factor(language, levels = LANG_LABELS[PAPER_LANGS])) |>
  arrange(language) |>
  transmute(
    dataset        = as.character(language),
    citation       = "[@CITE]",
    n_kids         = long_kids,                 # longitudinal kids only
    admins_per_kid = round(admins / total_kids, 1),
    age_range      = paste0(age_min, "–", age_max, " mo"),
    input_info     = "—"
  )

# --- Input-observed (IO) rows (hardcoded from each dataset's extract) ---
io_rows <- tibble::tribble(
                  ~dataset,    ~citation, ~n_kids, ~admins_per_kid, ~age_range, ~input_info,
  "BabyView",  "@long2024", 22L,     5.0,  "6–32 mo",  "Headcam (n = 5,739 videos)",
  "SEEDLingS", "@egan2025", 44L,     12.0, "6–17 mo",  "LENA (n = 525 recordings)",
  "AM2018",    "@adams2018", 66L,     3.0,  "16–24 mo", "LENA (n = 126 recordings)",
  "FMW2013",   "@fernald2013", 51L,     3.0,  "18–30 mo", "LENA (n = 51 recordings)"
)

table1 <- bind_rows(wb_rows, io_rows) 
table1$citation[table1$dataset == "Norwegian"] <- "@simonsen2014"
table1$citation[table1$dataset == "Japanese"] <- "@hagihara2023"
table1$citation[table1$dataset == "English (American)"] <- "@thal2013,@marchman2013, L. Smith"


table1 |>
  mutate(n_kids = format(n_kids, big.mark = ","),
         admins_per_kid = sprintf("%.1f", admins_per_kid)) |>
  knitr::kable(
    col.names = c("Dataset", "Citation",
                  "N", "Admins (mean)",
                  "Ages", "Recordings"),
    align = "lllrlll") |>
  kableExtra::kable_styling(font_size = 8)
```

Taking advantage of the equivalence of Rasch models to generalized linear mixed effects models [@deboeck2011], we fit the set of models shown in @fig-schematic as a series of logistic regressions. As described above, this initial set of fits do not attempt to estimate the relationship with input but instead simply attempt to describe the distribution of vocabulary growth curves. Each model predicted the probability of an individual child producing a particular word as a function of age (either unmodified in model A or multiplied by a constant in model B) and a random effect of word difficulty. Successive models add random intercepts (model C) and slopes (model D) by child.

Model D, which incorporated variation in acceleration and efficiency, was the best fitting model across all datasets, with AIC and BIC differences shown in @tbl-aic. @fig-glmer-ladder shows the fit of the four model variants; the increasing spread of children's trajectories with age is especially noticeable for low-vocabulary older children [who may be classified as "late talkers"; @fernald2012; @rescorla2011].^[Following the literature, we primarily investigate linear accumulator models, which show power-law growth of ability over time because $\theta$ is a function of $\log(t)$. However, it is also possible that ability grows linearly in time, which would lead to exponential increases in the probability of word production. We fit linear-time models, but found that they were dispreferred statistically, though differences were relatively small within the restricted age range for which we have data ($\delta$AIC `r aic_loglin_range`; see Supplemental Figure 2 for model fits).]

```{r tbl-aic}
#| label: tbl-aic
#| tbl-cap: ""
# Model-comparison table: dAIC and dBIC of each model variant relative to
# the best model within each language (the best is always D log-age, dAIC
# = dBIC = 0). Models without a per-child slope (A, B) have no log/linear
# distinction at the accumulation step; we tabulate B/C/D in both age
# bases. Values built from aic_summary (paper/build_cache.R <- glmer ladder).
lang_order <- c("english_american","norwegian","french_quebecois","japanese")
lang_short <- c(english_american="English", norwegian="Norwegian",
                french_quebecois="French (Q.)", japanese="Japanese")
mod_lab <- c(A="M1. accumulator",
             B_lin="M2. + acceleration", B_log="M2. + acceleration",
             C_lin="M3. + efficiency var.", C_log="M3. + efficiency var.",
             D_lin="M4. + acceleration var.", D_log="M4. + acceleration var.")
age_lab <- c(A="—", B_lin="linear", B_log="log", C_lin="linear",
             C_log="log", D_lin="linear", D_log="log")
mod_order <- c("A","B_lin","B_log","C_lin","C_log","D_lin","D_log")

tab <- aic |>
  mutate(Model = mod_lab[model], Age = age_lab[model],
         lang_slug = factor(lang_slug, levels = lang_order)) |>
  select(model, Model, Age, lang_slug, dAIC) |>
  tidyr::pivot_wider(names_from = lang_slug, values_from = dAIC) |>
  mutate(model = factor(model, levels = mod_order)) |>
  arrange(model) |>
  select(Model, Age, all_of(lang_order))

# dBIC, same shape, to append as a second column block
tabB <- aic |>
  mutate(lang_slug = factor(lang_slug, levels = lang_order),
         model = factor(model, levels = mod_order)) |>
  select(model, lang_slug, dBIC) |>
  tidyr::pivot_wider(names_from = lang_slug, values_from = dBIC) |>
  arrange(model) |> select(all_of(lang_order))

fmt <- function(x) ifelse(x == 0, "0", format(round(x), big.mark = ","))
out <- cbind(tab[, c("Model","Age")],
             lapply(tab[, lang_order], fmt) |> as.data.frame(),
             lapply(tabB,             fmt) |> as.data.frame())

knitr::kable(out, align = c("l","l", rep("r", 8)),
             col.names = c("Model", "Age", rep(unname(lang_short[lang_order]), 2))) |>
  kableExtra::add_header_above(c(" " = 2, "$\\\\Delta$AIC" = 4, "$\\\\Delta$BIC" = 4),
                               escape = FALSE) |>
  kableExtra::kable_styling(latex_options = "scale_down", font_size = 8)
```

```{r fig-glmer-ladder, fig.width=8, fig.height=5}
#| label: fig-glmer-ladder
#| fig-cap: "Individual children's longitudinal vocabulary growth trajectories across age in four languages. The same data are plotted in each column of panels, with overlaid models of increasing complexity (colored lines). Transparency is higher for English (American) and Norwegian panels to avoid overplotting."
# Replicates glmer_ladder/04b_plot.R style: empirical kid trajectories
# (grey lines + points) overlaid with the per-kid quantile predictions
# from each model. C_log and D_log get the full 10/25/50/75/90 fan
# (Wordbank palette); A and B_log have no per-kid variation so the
# quantiles collapse — we show only the median for them.
qtiles <- ladder$qtiles |>
  filter(model %in% c("A", "B_log", "C_log", "D_log")) |>
  mutate(qprob_f = factor(qprob, levels = c(0.1, 0.25, 0.5, 0.75, 0.9),
                          labels = c("0.1", "0.25", "0.5", "0.75", "0.9")))
emp <- ladder$emp |>
  mutate(child_lang = paste(lang_slug, child_id, sep = "::"))

model_lab <- c("1"     = "(M1) accumulator",
               "2_log" = "(M2) + pop. acceleration",
               "3_log" = "(M3) + per-child efficiency",
               "4_log" = "(M4) + per-child acceleration")
qtiles$model_label <- factor(model_lab[qtiles$model], levels = model_lab)
lang_lab <- ladder$lang_labels
emp$lang_label    <- factor(lang_lab[emp$lang_slug],    levels = lang_lab)
qtiles$lang_label <- factor(lang_lab[qtiles$lang_slug], levels = lang_lab)
# replicate empirical data per model for facetting (grey background)
emp_facet <- bind_rows(lapply(names(model_lab), function(m)
  emp |> mutate(model_label = factor(model_lab[m], levels = model_lab))))

# split quantiles by model: full fan for C_log/D_log, median only for A/B_log
qt_fan    <- filter(qtiles, model %in% c("M3_log", "M4_log"))
qt_median <- filter(qtiles, !model %in% c("M3_log", "M4_log"),
                    qprob_f == "0.5")

ggplot() +
  geom_line(data = filter(emp_facet, 
                          language %in% c("English (American)", "Norwegian")),
            aes(age, vocab, group = child_lang),
            color = "grey25", alpha = 0.03, linewidth = 0.25) +
  geom_line(data = filter(emp_facet, 
                          !(language %in% c("English (American)", "Norwegian"))),
            aes(age, vocab, group = child_lang),
            color = "grey25", alpha = 0.25, linewidth = 0.25) +
  # geom_point(data = emp_facet, aes(age, vocab),
  #            color = "grey25", alpha = 0.20, size = 0.25) +
  geom_line(data = qt_fan,
            aes(age, vocab, color = qprob_f, group = qprob_f,
                linewidth = ifelse(as.character(qprob_f) == "0.5",
                                   0.95, 0.55))) +
  geom_line(data = qt_median,
            aes(age, vocab, color = qprob_f, group = qprob_f),
            linewidth = 0.95) +
  facet_grid(lang_label ~ model_label, scales = "free_y", space = "fixed",
             switch = "y") +
  scale_color_manual(values = WORDBANK_PALETTE, name = "Percentile",
                     labels = c("10th", "25th", "50th", "75th", "90th"),
                     drop = FALSE) +
  scale_linewidth_identity() +
  labs(x = "Age (months)", y = "Productive vocabulary") +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(size = 8, face = "bold"),
        strip.placement = "outside",
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        panel.spacing = unit(0.4, "lines"))
```


## Efficiency and acceleration relate to separate demographic predictors


```{r fig-demographics, fig.width=7, fig.height=4.5}
#| label: fig-demographics
#| fig-cap: "Regression coefficients predicting individual children's efficiency and acceleration as a function of gender and maternal education. Black dots show meta-analytic estimates from a random-effects meta analysis. Error bars show 95\\% confidence intervals."
# Per-kid BLUPs (xi = intercept, zeta = log-age slope) from each language's
# D_log fit, joined with Wordbank demographics. For each (language, outcome)
# we regress the z-scored random effect on sex + maternal education.
#
# Maternal education enters as CONTINUOUS years (Wordbank's harmonized
# categories mapped to approximate years of schooling), standardized
# within-language. This (a) puts the differently-grained category sets on
# one scale (Norwegian has 4 levels, English/French 7), (b) uses 1 df
# instead of k-1, and (c) yields one slope per language we can pool.
# Birth order is dropped (sparse outside Norwegian). Sex is kept for all
# four languages; Japanese has NO maternal-education data, so it
# contributes to the sex term only.
ED_YEARS <- c("None" = 0, "Primary" = 6, "Some Secondary" = 9,
              "Secondary" = 12, "Some College" = 14, "College" = 16,
              "Some Graduate" = 18, "Graduate" = 20)

fit_one <- function(d, outcome) {
  d <- d |>
    mutate(y = as.numeric(scale(.data[[outcome]])),
           ed_years = unname(ED_YEARS[as.character(caregiver_ed)])) |>
    filter(!is.na(y), !is.na(sex))
  if (nrow(d) < 30) return(NULL)
  has_ed <- sum(!is.na(d$ed_years)) >= 30
  if (has_ed) {
    dd <- filter(d, !is.na(ed_years))
    m  <- lm(y ~ sex + scale(ed_years), data = dd)
  } else {
    dd <- d
    m  <- lm(y ~ sex, data = dd)   # Japanese: sex only
  }
  broom::tidy(m, conf.int = TRUE) |> mutate(outcome = outcome, n = nrow(dd))
}
combos <- expand.grid(
  lang_slug = unique(demo_blups$lang_slug),
  outcome   = c("xi", "zeta"),
  stringsAsFactors = FALSE
)
coefs <- bind_rows(lapply(seq_len(nrow(combos)), function(i) {
  d <- filter(demo_blups, lang_slug == combos$lang_slug[i])
  res <- fit_one(d, combos$outcome[i])
  if (is.null(res)) NULL else mutate(res, lang_slug = combos$lang_slug[i])
})) |>
  filter(term != "(Intercept)")

# Random-effects meta-analysis per (outcome, term) across the languages
# that contribute that term (4 for sex, 3 for education).
meta_summary <- coefs |>
  group_by(outcome, term) |>
  group_modify(~ {
    if (nrow(.x) < 2) return(tibble())
    res <- tryCatch(
      metafor::rma(yi = .x$estimate, sei = .x$std.error,
                   method = "REML", control = list(stepadj = 0.5)),
      error = function(e) NULL)
    if (is.null(res)) return(tibble())
    tibble(estimate = as.numeric(res$beta),
           conf.low = res$ci.lb, conf.high = res$ci.ub,
           std.error = res$se, k = res$k, 
           p = res$pval)
  }) |>
  ungroup() |>
  mutate(lang_slug = "_meta_")

coefs_all <- bind_rows(coefs, meta_summary) |>
  mutate(
    term_label = recode(term,
      "sexMale"         = "Male\n(vs Female)",
      "scale(ed_years)" = "Maternal education\n(per SD, years)") |>
      factor(levels = rev(c("Male\n(vs Female)",
                            "Maternal education\n(per SD, years)"))),
    outcome = factor(outcome, levels = c("xi", "zeta"),
                     labels = c("Efficiency~(xi)",
                                "Acceleration~(zeta)")),
    lang_label = factor(c(LANG_LABELS, "_meta_" = "Meta-analytic")[lang_slug],
                        levels = c(LANG_LABELS[ladder$langs],
                                   "Meta-analytic")))
plot_pal <- c(setNames(unname(LANG_PAL[ladder$langs]),
                       unname(LANG_LABELS[ladder$langs])),
              "Meta-analytic" = "black")

coefs_all <- coefs_all |>
  mutate(is_meta = lang_label == "Meta-analytic")

ggplot(coefs_all,
       aes(estimate, term_label, color = lang_label,
           group = lang_label, size = is_meta)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey55") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
                  position = position_dodge(width = 0.70),
                  fatten = 3) +
  facet_wrap(~ outcome, ncol = 2, scales = "free_x",
             labeller = label_parsed) +
  scale_color_manual(values = plot_pal, name = NULL, drop = FALSE) +
  scale_size_manual(values = c(`FALSE` = 0.35, `TRUE` = 0.65),
                    guide = "none") +
  guides(color = guide_legend(nrow = 1)) +
  labs(x = "regression coefficient (z-scored outcome)", y = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold", size = 11),
        axis.text.y = element_text(size = 9))
```

Children's vocabulary is known to be related to a number of sociodemographic predictors, including sex and maternal education (which is interpreted as a proxy for socioeconomic status). In particular, female children show consistently larger vocabularies than male children across dozens of languages and countries [@eriksson2012;@frank2021]. Similarly, children of more educated mothers tend to show larger vocabularies, though this effect is somewhat more heterogeneous across countries and languages [@frank2021].

Our longitudinal model offers a summary of individual children's growth in terms of both their efficiency (intercept) and their acceleration (slope). This decomposition allows us to ask whether these two demographics effects -- sex and maternal education -- are seen in efficiency, acceleration, or both. For each child in our analysis, we extracted the best linear unbiased prediction (BLUP) from our models. We then regressed intercept and slope terms against each demographic predictor separately for each language. @fig-demographics shows the coefficients from this analysis, along with a cross-language meta-analytic estimate. 

Sex effects show the predicted female advantage across all four languages, but this effect manifests itself only in efficiency  ($\beta = `r round(meta_summary$estimate[2],2)`$, $p = `r papaja::apa_p(meta_summary$p)[2]`$) not acceleration ($\beta = `r round(meta_summary$estimate[4],2)`$, $p = `r papaja::apa_p(meta_summary$p)[4]`$). In contrast, though maternal education data were only available for three languages, effects were numerically present in both measures, though stronger in acceleration ($\beta = `r round(meta_summary$estimate[3],2)`$, $p = `r papaja::apa_p(meta_summary$p)[3]`$) than in efficiency ($\beta = `r round(meta_summary$estimate[1],2)`$, $p = `r papaja::apa_p(meta_summary$p)[1]`$). One interpretation of these findings is that sex effects reflect a small constant offset in average ability. In contrast, socioeconomic differences as reflected in maternal education might continue to compound across development [@fernald2013;@hart2003].


## Growth in ability leads to rapid changes in learning times for individual words

```{r fig-efficiency, fig.width=7, fig.height=4}
#| label: fig-efficiency
#| fig-cap: ""
# Per-word "exposures to 50% production" for a typical English child,
# from the M_best fit + CHILDES frequencies (data in fig4_exposure.rds;
# see paper/build_cache.R section 5 and model/scripts/exposure_to_learn.R).
# Each point is a word; x = predicted age at 50% production, y = cumulative
# exposures of that word by that age (log scale). Lines: per-class lm fits.
items <- exposure$items
meta  <- exposure$meta

# label the extreme word per class on log-exposure (balanced across classes)
labels_df <- items |>
  mutate(log_N = log10(N_word)) |>
  group_by(lexical_class) |>
  arrange(log_N) |>
  slice(c(1, n())) |>
  ungroup()

ggplot(items, aes(t_50, N_word, colour = lexical_class)) +
  geom_point(alpha = 0.55, size = 1.1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7, formula = y ~ x) +
  geom_label_repel(data = labels_df, aes(label = item),
                   size = 2.8, alpha = 0.95, show.legend = FALSE,
                   box.padding = 0.4, max.overlaps = Inf,
                   min.segment.length = 0) +
  scale_y_log10(labels = scales::label_number(big.mark = ",")) +
  scale_x_continuous(limits = c(10, 40)) +
  scale_colour_brewer(palette = "Dark2", name = "Class") +
  labs(x = "Age (months) when word is produced at 50%",
       y = "Cumulative exposures of word at that age") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(size = 9, colour = "grey25"),
        legend.position = "bottom")
```

As an illustration of this change in efficiency, Figure 2 shows a model-derived estimate of the average number of times a child would need to hear a word before producing it (See SI: Changes in Word Efficiency), using frequency estimates from the CHILDES Corpus [@mcwhinney2020;@sanchez2019]. Across different word classes (e.g., nouns, verbs, closed class words), efficiency increases exponentially, such that early-learned words such as "ball" may be heard tens of thousands of times before they are produced, while later learned words such as "vitamin" might only be heard tens of times before being produced. 


## Estimating input 


Under a range of plausible assumptions about variation in input, all children show supra-linear scaling of growth, even accounting for accumulation dynamics.

As discussed above, one important question is how much of variance is due to variance in rate. We explore this by asking about the parameter $\pi$,

$$\pi = \frac{\sigma_\alpha^2}{\sigma_\alpha^2 + \sigma_r^2}$$

If $\pi$ > .5, then the majority of variation between individuals is not due to pure accumulation rate. 

In addition, 80-90% of between-child variation remains unexplained by input quantity. 
These conclusions are supported by analysis of longitudinal input data from BabyView (N=20) and SEEDLings (N=44) datasets, which contain child-specific estimates of input, and Peekbank (N=XYZ), which contains child-specific estimates of processing speed.

```{r fig-io-partition, fig.width=11, fig.height=4.5}
#| label: fig-io-partition
#| fig-cap: ""
# Two-panel: (A) observed vocabulary trajectories per IO study; (B) age-
# dependent variance decomposition under the additive gamma model. Panel
# B uses sigma_r, sigma_alpha, sigma_zeta, gamma at posterior median (see
# experiments.md §30 for the diagnostic / wide-delta refit).
obs <- io_summary$obs |>
  mutate(study = factor(study, levels = names(STUDY_PAL)))
p_obs <- ggplot(obs, aes(age, prop, color = study)) +
  geom_point(alpha = 0.35, size = 0.6) +
  geom_smooth(se = FALSE, method = "loess", span = 0.7, linewidth = 0.7,
              formula = y ~ x) +
  scale_color_manual(values = STUDY_PAL, name = "study") +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  labs(x = "age (months)", y = "vocabulary proportion produced",
       title = "(A) IO datasets: observed growth") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 10))

# Variance partition (additive form, mean of slope channel)
p   <- io_summary$params
a0  <- 15; ages <- seq(8, 30, length.out = 60); L <- log(ages / a0)
sr2 <- p$sigma_r^2; sa2 <- p$sigma_alpha^2
sz2 <- p$sigma_zeta_add^2; g <- p$gamma_add
parts <- tibble(age = rep(ages, 4),
                L   = rep(L,    4),
                component = factor(rep(c("Input: intercept",
                                         "Input: slope (γ)",
                                         "Efficiency: intercept",
                                         "Efficiency: slope"),
                                       each = length(ages)),
                  levels = c("Efficiency: slope", "Input: slope (γ)",
                             "Efficiency: intercept", "Input: intercept"))) |>
  mutate(var = case_when(
    component == "Input: intercept"      ~ sr2,
    component == "Input: slope (γ)"      ~ pmax(L^2 * g^2 * sr2 + 2 * L * g * sr2, 0),
    component == "Efficiency: intercept" ~ sa2,
    component == "Efficiency: slope"     ~ L^2 * sz2
  ))

p_partition <- ggplot(parts, aes(age, var, fill = component)) +
  geom_area(alpha = 0.95) +
  scale_fill_manual(values = c(
    "Input: intercept"      = "#B4D7E8",
    "Input: slope (γ)"      = "#1F77B4",
    "Efficiency: intercept" = "#003D7A",
    "Efficiency: slope"     = "#E8A0A8"), name = NULL) +
  labs(x = "age (months)",
       y = expression(Var[i](theta[it])),
       title = "(B) Variance partition (γ-additive, wide-delta)") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 10),
        legend.text = element_text(size = 8))

p_obs + p_partition + plot_layout(widths = c(1, 1))
```


Thus, modeling human learning requires positing other sources of variability and acceleration. Sources of variability could include from variation in interactional input quality, genetic variation between children, environmental richness not captured by input, etc. Sources of acceleration could include increases in processing efficiency, learning-to-learn including mutual exclusivity, shape bias, syntactic bootstrapping, as well as domain general changes in perception, processing speed, memory, attention or other aspects of cognition. The key to understanding children’s data efficiency may thus lie in better understanding when and how children’s learning accelerates with maturation and experience. 

We end by providing a quantitative comparison between within-child human learning and LLM scaling laws, showing that children’s early language learning is strikingly different from LLM scaling: it shows both steeper scaling and greater variance in growth rate.

```{r fig-llm-acceleration, fig.width=11, fig.height=4.5}
#| label: fig-llm-acceleration
#| fig-cap: ""
# Best-guess two-panel composite (will iterate). Each panel is an
# existing PNG composed via grid; replace with chunked plots once we
# decide which LLM-vs-children comparison to feature.
panel_paths <- list(
  A = here("outputs", "figs", "longitudinal",
           "chang_bergen_slope_comparison.png"),
  B = here("outputs", "figs", "schematic", "D1_scaling_disanalogy.png"))
p_to_grob <- function(path) {
  if (!file.exists(path)) return(grid::nullGrob())
  grid::rasterGrob(png::readPNG(path), interpolate = TRUE)
}
gridExtra::grid.arrange(
  p_to_grob(panel_paths$A),
  p_to_grob(panel_paths$B),
  ncol = 2
)
```



# Discussion

What we did

Longitudinal growth models efficiency and acceleration. Two different paramters. 

Where doe acceleration come from? Could come from a variety of sources, including general maturation of memory and attention, specific changes in word recognition or processing [@fernald1998;@frank2026], the increasing sophistication of inferential strategies for word learning [@smith2002;@markman1988;etc], or other sources. 

Role of input: Coffey 2026 variance decomposition

<!-- Children’s early word learning is alternatively portrayed as either a process of gradual accumulation or as an inferential process that results in rapid mapping. How can these two views be rectified?  -->

<!-- Children’s early language learning is often portrayed as a process of accumulation and gradual learning, with statistical learning, mutual exclusivity and input viewpoints highlighting the accumulation nature. In contrast, studies of processing and word mapping highlight the increases in efficiency that happen over the same  period. How can these two views be rectified? One clue comes from examining the two key signatures of early vocabulary growth: variability and acceleration. What is the shape of this acceleration? And what is the source of variability, from children’s input or otherwise?? We present a quantitative model that unify these views while explaining both the variability and acceleration of children’s early language growth. The resulting model provides a good quantitative explanation for longitude vocabulary growth across large population of children. It also shows that by their increasing efficiency, and their ability of children, learning fundamentally different ways than large language, models, which show classic accumulator dynamics. -->

Furthermore, how does the accumulation process relate to the rapid growth of vocabulary? On some models, accumulation by itself can produce an exponential “vocabulary explosion” while the child’s own learning abilities remain constant. On other accounts, however, the child’s own abilities change over development. These changes could be due to the accumulation of language allowing bootstrapping (e.g., via mutual excusivity or the shape bias), they could be through faster processing (Fernald rich get richer), or they could be via general maturation (e.g., better memory, faster processing). Or all of these. 


<!-- Blingualism  -->

<!-- International adoption -->

<!-- https://bergelsonlab.fas.harvard.edu/sites/g/files/omnuum3941/files/2025-01/meylan_bergelson_2021_ARL.pdf	 -->

# References

:::{.acknowledgement}
Claude Code was used for literature review, code generation, visualization, and simulation infrastructure. All writing is original to the author. 
:::

<!-- :::{.suppinfo} -->


<!-- ::: -->

