---
author:
  - name: Michael C. Frank
    affiliations: 
      - name: Department of Psychology, Stanford University

title: Efficiency and acceleration in the accumulation of children’s early vocabulary
title-short: Efficiency and acceleration 
keywords: Early word learning; Bayesian data analysis; Large language models
abstract: |
  Children learn hundreds of words over the first few years of their life. This process begins slowly but increases until only a few exposures to a word are necessary for a child to learn it. Yet there are also wide variations between children, and the distribution across children has not not been accurately characterized. Here we present a statistical model describing this process as accelerating accumulation, with children varying in both the slope and intercept of their accumulation. Using data from four languages, this model provides a good characterization of longitudinal growth and provides insights into the sources of age and socioeconomic variation. Using data on children's early language environment, the model also can be used to estimate how much variation between children is due to differences in language input, finding estimates that converge with recent meta-analyses. In contrast to children, large language models are not well-described by this model: they show standard accumulator dynamics and very limited variability across a number of dimensions. 

bibliography: standard_model.bib
format:
  pdf:
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
    linestretch: 1.05
    colorlinks: true
    keep-tex: true
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(here); library(dplyr); library(tidyr); library(readr)
  library(ggplot2); library(patchwork); library(latex2exp)
  library(knitr); library(scales); library(stringr); library(forcats)
  library(broom); library(png); library(grid); library(gridExtra)
})
source(here("paper", "_helpers.R"))

# load cached data — produced by `Rscript paper/build_cache.R`
# (Table 1 is built inline in its own chunk, not cached.)
CACHE <- here("paper", "cache")
ladder     <- readRDS(file.path(CACHE, "fig2_glmer_ladder.rds"))
demo_blups <- readRDS(file.path(CACHE, "blups_demographics.rds"))
io_summary <- readRDS(file.path(CACHE, "fig5_io_summary.rds"))

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE, cache=TRUE,
  fig.align = "center", out.width = "100%", dpi = 300
)
```

<!-- Children’s early word learning is alternatively portrayed as either a process of gradual accumulation or as an inferential process that results in rapid mapping. How can these two views be rectified?  -->

<!-- Children’s early language learning is often portrayed as a process of accumulation and gradual learning, with statistical learning, mutual exclusivity and input viewpoints highlighting the accumulation nature. In contrast, studies of processing and word mapping highlight the increases in efficiency that happen over the same  period. How can these two views be rectified? One clue comes from examining the two key signatures of early vocabulary growth: variability and acceleration. What is the shape of this acceleration? And what is the source of variability, from children’s input or otherwise?? We present a quantitative model that unify these views while explaining both the variability and acceleration of children’s early language growth. The resulting model provides a good quantitative explanation for longitude vocabulary growth across large population of children. It also shows that by their increasing efficiency, and their ability of children, learning fundamentally different ways than large language, models, which show classic accumulator dynamics. -->

Children’s early vocabulary grows very rapidly during early childhood, with large variation between children. The nature of this growth has theoretical, practical, and methodological importance.

From a theoretical perspective, growth mechanisms shed light on children’s fundamental learning mechanisms and how they change across childhood. From a practical perspective, variation in growth is an important precursor — and predictor — of academic achievement. From a methodological perspective, vocabulary also affords a unique psychometric opportunity to study human variation because, unlike nearly all important human psychological metrics, there is a true zero point - a time before which children know no words. Thus, absolute levels of variability can be calculated. 

One dominant viewpoint is that this growth can be characterized as a process of accumulation [@mcmurray2007; @mitchell2009; @hidaka2013; @mollica2017; @kachergis2022]. This pure accumulator view connects with a viewpoint in which the quantity of language that children hear has a strong causal role in the speed of their learning [@fernald2012]. 

<!-- It also connects with the idea that large language models – whose learning scales in predictable, accumulator-like ways (Kaplan, Chinchilla, scaling laws) – might be good models for children’s learning.  -->

But important parts of this proposal have not been fleshed out. For example, how much variation is due to input quantity vs. other sources? Meta-analysis of this relationship suggests it’s there but not that big [@coffey2026]. Could be quality as well, of course.

Furthermore, how does the accumulation process relate to the rapid growth of vocabulary? On some models, accumulation by itself can produce an exponential “vocabulary explosion” while the child’s own learning abilities remain constant. On other accounts, however, the child’s own abilities change over development. These changes could be due to the accumulation of language allowing bootstrapping (e.g., via mutual excusivity or the shape bias), they could be through faster processing (Fernald rich get richer), or they could be via general maturation (e.g., better memory, faster processing). Or all of these. 

We address these questions by creating Bayesian data analysis models that connect proposals about the nature of the accumulation process with large-scale data about the longitudinal growth of children’s vocabulary over time. Our key innovations are 1) that we model population variability between children, not just the central tendency and 2) that we directly connect quantitative estimates of language exposure to the rate of vocabulary growth.


We fit these models to data on children's language production. Our primary data source is the MacArthur-Bates Communicative Development Inventories [MB-CDI; @fenson1994;@marchman2021], a popular parent report instrument in which parents are asked to indicate which words their child produces and understand. Here we focus exclusively on production, which is used with older children and which tends to be a more reliable measure [@frank2021;others]. We make use of data from Wordbank [@frank2017], an open repository of data from the MB-CDI. Since our efficiency variation model is not identifiable in cross-sectional data, we focus on those datasets that include substantial amounts of longitudinal data, which come from American English (N=XYZ), Norwegian (N=XYZ), Quebec French (N=XYZ), and Japanese (N=XYZ). 

Critically, we make use of new data resources, first connecting between vocabulary growth and data-driven estimates of the general range of language input quantity available to children (study 1), then to estimates of specific children’s language input from BabyView and SEEDLings datasets (study 2), and finally to estimates of children’s processing speed from Peekbank (study 3). 

Overall, our analysis supports a view in which children’s vocabulary growth is not a pure process of accumulation but rather one that accelerates rapidly over time. Further, although variation in input rate accounts for significant variation in learning speed, it is limited to approximately 10% of total variation across datasets. 

Finally, we explore an application of the same framework to large language models. Making use of GPT-2 models trained on human developmental data, we show that – in contrast to children – these models do appear to function as pure accumulators. This analysis exposes a critical disanalogy between LLMs and human learners and suggests a potential route for understanding the vast “data gap” between children and large language models [@frank2023]. 

Here we ask whether these two signatures of vocabulary growth - variation and acceleration - are consistent with pure accumulator models, which are prevalent in the word learning literature and which characterize the scaling dynamics of large language models.


# Model 


```{r fig-schematic, fig.width=7, fig.height=4.5}
#| label: fig-schematic
#| fig-cap: ""
# Schematic walk through the model ladder. Four columns are model
# variants (A → D). Top row: latent ability theta_i(t). Bottom row: the
# expected VOCABULARY curve, summed over a distribution of word
# difficulties — V(t) = mean_j plogis(theta_i(t) - delta_j). This is the
# McMurray (2007) construction: even a pure accumulator (A, kappa=1)
# produces an accelerating vocabulary curve because theta sweeps through
# the difficulty distribution. The SAME Gaussian difficulty distribution
# is used in every panel (the same words), so the columns differ only in
# the latent dynamics (kappa) and between-child variance (sigma_alpha,
# sigma_kappa), not in the word set. Five quantiles (10/25/50/75/90) in
# the Wordbank palette match Fig 2.
ages <- seq(8, 30, length.out = 80)
A0   <- 8                                  # anchor age: log(t/A0) >= 0
L    <- log(ages / A0)
qs   <- c(0.1, 0.25, 0.5, 0.75, 0.9)
z    <- qnorm(qs)

XI0     <- -2.0                            # median intercept (log r + log alpha)
KAPPA_A <- 2.5                             # acceleration for B/C/D
# shared word-difficulty distribution (Gaussian, McMurray 2007)
DELTA   <- qnorm(ppoints(150), mean = 0, sd = 1.5)
vocab_of <- function(th) vapply(th, function(x) mean(plogis(x - DELTA)), numeric(1))

# population parameters chosen to be illustrative, not estimated.
variants <- tibble::tribble(
  ~name,                          ~kappa_pop, ~sigma_alpha, ~sigma_kappa, ~eq,
  "(A) pure accumulator",         1,           0,            0,            r"($\theta = \log(r) + \log(t)$)",
  "(B) + acceleration",           KAPPA_A,     0,            0,            r"($\theta = \log(r) + \kappa\,\log(t)$)",
  "(C) + efficiency variance",    KAPPA_A,     1.1,          0,            r"($\theta = \log(\alpha_i) + \log(r) + \kappa\,\log(t)$)",
  "(D) + per-child acceleration", KAPPA_A,     1.1,          0.7,          r"($\theta = \log(\alpha_i) + \log(r) + \kappa_i\,\log(t)$)"
)

# For each model x quantile x age: theta and the summed vocabulary curve
sim <- variants |>
  mutate(name = factor(name, levels = name), panel_order = seq_along(name)) |>
  rowwise() |>
  do({
    v <- .
    bind_rows(lapply(seq_along(qs), function(i) {
      th <- XI0 + z[i] * v$sigma_alpha +
            (v$kappa_pop + z[i] * v$sigma_kappa) * L
      tibble(
        name = v$name, panel_order = v$panel_order,
        qprob = qs[i], qprob_f = factor(as.character(qs[i]),
                                        levels = c("0.1","0.25","0.5","0.75","0.9")),
        age   = ages, theta = th, vocab = vocab_of(th), eq = v$eq
      )
    }))
  }) |>
  ungroup()

# annotation table — one equation per panel
eq_lab <- variants |>
  mutate(name = factor(name, levels = name)) |>
  select(name, eq)

# vertical line size: median bold, others thinner
lwidth <- function(qf) ifelse(as.character(qf) == "0.5", 0.95, 0.55)

p_theta <- ggplot(sim,
                  aes(age, theta, color = qprob_f, group = qprob_f)) +
  geom_line(aes(linewidth = lwidth(qprob_f))) +
  geom_label(data = eq_lab,
             aes(x = -Inf, y = Inf,
                 label = latex2exp::TeX(eq, output = "character")),
             inherit.aes = FALSE, parse = TRUE,
             hjust = -0.05, vjust = 1.4, size = 2.4, label.size = 0,
             fill = scales::alpha("white", 0.8)) +
  facet_wrap(~ name, nrow = 1) +
  scale_color_manual(values = WORDBANK_PALETTE, name = "Percentile",
                     labels = c("10th", "25th", "50th", "75th", "90th"),
                     drop = FALSE) +
  scale_linewidth_identity() +
  coord_cartesian(ylim = c(-3.5, 4.5)) +
  labs(x = NULL, y = expression(theta[i](t))) +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(face = "bold", size = 9),
        legend.position = "bottom",
        panel.grid.minor = element_blank())

p_vocab <- ggplot(sim,
                  aes(age, vocab, color = qprob_f, group = qprob_f)) +
  geom_line(aes(linewidth = lwidth(qprob_f))) +
  facet_wrap(~ name, nrow = 1) +
  scale_color_manual(values = WORDBANK_PALETTE, name = "Percentile",
                     labels = c("10th", "25th", "50th", "75th", "90th"),
                     drop = FALSE) +
  scale_linewidth_identity() +
  scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1)) +
  labs(x = "age (months)", y = "expected vocabulary proportion") +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_blank(),  # suppress strip on lower row
        legend.position = "bottom",
        panel.grid.minor = element_blank())

(p_theta / p_vocab) + plot_layout(guides = "collect", heights = c(1, 1)) &
  theme(legend.position = "bottom")
```

We present a nested set of accumulator-type models, many of which have been presented in the literature. As we show below, each refinement of the model contributes to its ability to describe the variability between children. We start with the Rasch model from psychometrics [@rasch1960;@bond2013], which assumes that each child $i$ has a learning ability $\theta_i$ and each word $j$ has a difficulty $\delta_j$. The probability that child $i$ produces word $j$ is

$$P(w_{i,j}=1 \mid \theta_i, \delta_j) =\frac{\exp(\theta_i - \delta_j)}{1 + \exp(\theta_i - \delta_j)}$$
  
  
Following @kachergis2022, we can write our first model a simple Rasch-style accumulator model ("M1: pure accumulator"), specifying that $\theta_i$ is not a single parameter but instead a function of accumulation over time. In the simplest form: 
  
  $$\theta_i  = \log(r_i) + \log(t)$$
  
  \noindent where $r_i$ is the child's rate of language input (in tokens per month) and $t$ is the child's (age in months); note that we use log age to simulate linear accumulation because $\exp(\theta) = r \cdot t$. This form instantiates the basic hypothesis that the only thing that changes over time is more input, and the only source of variation between children is variation in rate. 

We compare this model to an accelerating accumulator model, in which the amount learned from each exposure increases with time by a multiplicative factor $kappa$ (which manifests as an exponent on time in the full Rasch model).

$$\theta_i  = \log(r_i) + \kappa ~ \log(t)$$
Neither of these models has any way to capture variation across children beyond variation in input rate. Thus, we augment the basic accelerating accumulator in two ways, rewriting it as the the sum of two individually variable quantities:

$$\theta_i = \xi_i + \zeta_i ~ \log(t)$$
The intercept term $\xi_i = \log(\alpha_i) + \log(\r_i)$ then assumes that there is both rate-related and non-rate variation in children's efficiency. Similarly, the slope term $\zeta_i = \kappa_i + \gamma_i \log r_i$ decomposes acceleration into rate-related and non-rate-related acceleration. 

This final model instantiates the idea that children vary in their base rate of efficiency and their acceleration, instantiating  a longitudinal growth model in which children vary on their slope and intercept. As we will show, this model provides a good description of variation in children's longitudinal growth trajectories. @fig-schematic shows schematic predictions from a pure accumulator model (A), an accelerating accumulator (B) in comparison to models including child-level variation in efficiency (C) and both efficiency and acceleration (D). Each model's predictions are represented as ability scores ($\theta$) across age and then also projected into proportion correct scores on a discrete set of vocabulary items, representing an observed accuracy score that is subject to compression at floor and ceiling. 

A variety of previous work has focused on estimating the difficulty of individual words  [@goodman2008;@hidaka2013;@roy2015;@braginsky2019;@kachergis2022]. These analyses reveal that there is predictable word-level difficulty information but that only a modest portion of the variation between words is carried by frequency information. Words can be hard for a child to acquire because they vary also in the complexity and concreteness of their meanings, their syntactic category and syntactic context, their phonology, and many other factors. In the current work, we estimate word difficulty as a free parameter, and focus instead on variation between children. 

Our analysis generally depends on the availability of longitudinal data at the item level. The different models described above are not distinguishable from the cross-sectional, aggregate data that are typically considered in many analyses of early vocabulary. First, aggregate data do not allow for the estimation of individual word difficulties, thus the distribution of difficulties remains latent and different difficulty distributions can be posited to explain the accelerating shape of vocabulary growth [@mcmurray2007; @mitchell2009]. Second, cross-sectional data do not allow for accurate estimation of whether variation is due to differences in efficiency (intercept) or acceleration (slope) (see Supplemental Figure 1). 

# Results

## Children's longitudinal growth reflects differences in both efficiency and acceleration

We first retrieved longitudinal data on children's vocabulary from Wordbank [@frank2017], an open archive of data from the MacArthur-Bates Communicative Development Inventory and its international adaptations. @tbl-datasets shows the four longitudinal datasets in Wordbank that contained more than $>100$ children with multiple observations.  

```{r tbl-datasets}
#| label: tbl-datasets
#| tbl-cap: "Characteristics of the datasets."
# Wordbank rows are pulled from outputs/glmer_ladder/00_language_survey.csv;
# IO rows are hardcoded from each bundle's actual extracts. Citation column
# is a [@CITE] placeholder for the author to fill in.
# Table 1 is built inline (not cached) so the numbers and names are easy
# to edit here. Wordbank longitudinal rows are derived from the language
# survey CSV (outputs/glmer_ladder/00_language_survey.csv); input-observed
# (IO) rows are hardcoded from each dataset's extract. The Citation column
# holds [@CITE] placeholders for the author to fill in.

# --- Wordbank longitudinal rows (from the survey) ---
survey <- read_csv(
  here("outputs", "glmer_ladder", "00_language_survey.csv"),
  show_col_types = FALSE)

wb_rows <- survey |>
  filter(language %in% LANG_LABELS[PAPER_LANGS]) |>
  # keep the paper's language order
  mutate(language = factor(language, levels = LANG_LABELS[PAPER_LANGS])) |>
  arrange(language) |>
  transmute(
    dataset        = as.character(language),
    citation       = "[@CITE]",
    n_kids         = long_kids,                 # longitudinal kids only
    admins_per_kid = round(admins / total_kids, 1),
    age_range      = paste0(age_min, "–", age_max, " mo"),
    input_info     = "—"
  )

# --- Input-observed (IO) rows (hardcoded from each dataset's extract) ---
io_rows <- tibble::tribble(
                  ~dataset,    ~citation, ~n_kids, ~admins_per_kid, ~age_range, ~input_info,
  "BabyView",  "[@CITE]", 22L,     5.0,  "6–32 mo",  "Head-cam (n = 5,739 videos)",
  "SEEDLingS", "[@CITE]", 44L,     12.0, "6–17 mo",  "LENA day (n = 525 recordings)",
  "AM2018",    "[@CITE]", 66L,     3.0,  "16–24 mo", "LENA (n = 126 recordings)",
  "FMW2013",   "[@CITE]", 51L,     3.0,  "18–30 mo", "LENA (n = 51 recordings)"
)

table1 <- bind_rows(wb_rows, io_rows)


table1 |>
  mutate(n_kids = format(n_kids, big.mark = ","),
         admins_per_kid = sprintf("%.1f", admins_per_kid)) |>
  knitr::kable(
    col.names = c("Dataset", "Citation",
                  "N", "Admins (mean)",
                  "Ages", "Recordings"),
    align = "lllrlll") |>
  kableExtra::kable_styling(font_size = 8)
```

Taking advantage of the equivalence of Rasch models to generalized linear mixed effects models [@deboeck2011], we fit the set of models shown in Figure @fig-schematic as a series of logistic regressions. Each model predicted the probability of an individual child producing a particular word as a function of age (either unmodified in model A or multiplied by a constant in model B) and a random effect of word difficulty. Successive models added  random intercepts (model C) and slopes (model D) by child. Model D, which incorporated variation in acceleration and efficiency, was the best fitting model across all four datasets, with AIC and BIC differences shown in @tbl-aic. @fig-glmer-ladder shows the fit of the four model variants; the increasing spread of children's trajectories with age is especially noticeable for low-vocabulary older children [who may be classified as "late talkers"; @fernald2012; @rescorla2011].

Following the literature, we primarily investigate linear accumulator models, which show power-law growth of ability over time because $\theta$ is a function of $\log(t)$. However, it is also possible that ability grows linearly in time, which would lead to exponential increases in the probability of word production. We fit linear-time models, but found that they were dispreferred statistically, though differences were relatively small within the restricted age range for which we have data ($\delta$AIC `RENDER AIC RANGE FOR LOG AGE VS LINEAR VARIANTS`; see Supplemental Figure 2 for model fits). 

```{r fig-glmer-ladder, fig.width=8, fig.height=5}
#| label: fig-glmer-ladder
#| fig-cap: "Individual children's longitudinal vocabulary growth trajectories across age in four languages. The same data are plotted in each column of panels, with overlaid models of increasing complexity (colored lines). Transparency is higher for English (American) and Norwegian panels to avoid overplotting."
# Replicates glmer_ladder/04b_plot.R style: empirical kid trajectories
# (grey lines + points) overlaid with the per-kid quantile predictions
# from each model. C_log and D_log get the full 10/25/50/75/90 fan
# (Wordbank palette); A and B_log have no per-kid variation so the
# quantiles collapse — we show only the median for them.
qtiles <- ladder$qtiles |>
  filter(model %in% c("A", "B_log", "C_log", "D_log")) |>
  mutate(qprob_f = factor(qprob, levels = c(0.1, 0.25, 0.5, 0.75, 0.9),
                          labels = c("0.1", "0.25", "0.5", "0.75", "0.9")))
emp <- ladder$emp |>
  mutate(child_lang = paste(lang_slug, child_id, sep = "::"))

model_lab <- c("A"     = "(A) accumulator",
               "B_log" = "(B) + pop. acceleration",
               "C_log" = "(C) + per-child efficiency",
               "D_log" = "(D) + per-child acceleration")
qtiles$model_label <- factor(model_lab[qtiles$model], levels = model_lab)
lang_lab <- ladder$lang_labels
emp$lang_label    <- factor(lang_lab[emp$lang_slug],    levels = lang_lab)
qtiles$lang_label <- factor(lang_lab[qtiles$lang_slug], levels = lang_lab)
# replicate empirical data per model for facetting (grey background)
emp_facet <- bind_rows(lapply(names(model_lab), function(m)
  emp |> mutate(model_label = factor(model_lab[m], levels = model_lab))))

# split quantiles by model: full fan for C_log/D_log, median only for A/B_log
qt_fan    <- filter(qtiles, model %in% c("C_log", "D_log"))
qt_median <- filter(qtiles, !model %in% c("C_log", "D_log"),
                    qprob_f == "0.5")

ggplot() +
  geom_line(data = filter(emp_facet, 
                          language %in% c("English (American)", "Norwegian")),
            aes(age, vocab, group = child_lang),
            color = "grey25", alpha = 0.03, linewidth = 0.25) +
  geom_line(data = filter(emp_facet, 
                          !(language %in% c("English (American)", "Norwegian"))),
            aes(age, vocab, group = child_lang),
            color = "grey25", alpha = 0.25, linewidth = 0.25) +
  # geom_point(data = emp_facet, aes(age, vocab),
  #            color = "grey25", alpha = 0.20, size = 0.25) +
  geom_line(data = qt_fan,
            aes(age, vocab, color = qprob_f, group = qprob_f,
                linewidth = ifelse(as.character(qprob_f) == "0.5",
                                   0.95, 0.55))) +
  geom_line(data = qt_median,
            aes(age, vocab, color = qprob_f, group = qprob_f),
            linewidth = 0.95) +
  facet_grid(lang_label ~ model_label, scales = "free_y", space = "fixed",
             switch = "y") +
  scale_color_manual(values = WORDBANK_PALETTE, name = "Percentile",
                     labels = c("10th", "25th", "50th", "75th", "90th"),
                     drop = FALSE) +
  scale_linewidth_identity() +
  labs(x = "Age (months)", y = "Productive vocabulary") +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(size = 8, face = "bold"),
        strip.placement = "outside",
        panel.grid.minor = element_blank(),
        legend.position = "bottom",
        panel.spacing = unit(0.4, "lines"))
```

The best-fitting modified accumulator model included large developmental changes in efficiency across all languages ($\beta = XYZ - XYZ$). This finding suggests that we can robustly reject pure accumulation models of early word learning. Instead, children get vastly more efficient over development. 


This change in efficiency could come from a variety of sources, including general maturation of memory and attention, specific changes in word recognition or processing [@fernald1998;@frank2026], the increasing sophistication of inferential strategies for word learning [@smith2002;@markman1988;etc], or other sources. 

## Modeling demographic variation 


```{r fig-demographics, fig.width=9, fig.height=5.5}
#| label: fig-demographics
#| fig-cap: ""
# Per-kid BLUPs (xi = intercept, zeta = log-age slope) from each language's
# D_log fit, joined with Wordbank demographics. For each (language, outcome)
# we regress the z-scored random effect on sex + maternal education.
#
# Maternal education enters as CONTINUOUS years (Wordbank's harmonized
# categories mapped to approximate years of schooling), standardized
# within-language. This (a) puts the differently-grained category sets on
# one scale (Norwegian has 4 levels, English/French 7), (b) uses 1 df
# instead of k-1, and (c) yields one slope per language we can pool.
# Birth order is dropped (sparse outside Norwegian). Sex is kept for all
# four languages; Japanese has NO maternal-education data, so it
# contributes to the sex term only.
ED_YEARS <- c("None" = 0, "Primary" = 6, "Some Secondary" = 9,
              "Secondary" = 12, "Some College" = 14, "College" = 16,
              "Some Graduate" = 18, "Graduate" = 20)

fit_one <- function(d, outcome) {
  d <- d |>
    mutate(y = as.numeric(scale(.data[[outcome]])),
           ed_years = unname(ED_YEARS[as.character(caregiver_ed)])) |>
    filter(!is.na(y), !is.na(sex))
  if (nrow(d) < 30) return(NULL)
  has_ed <- sum(!is.na(d$ed_years)) >= 30
  if (has_ed) {
    dd <- filter(d, !is.na(ed_years))
    m  <- lm(y ~ sex + scale(ed_years), data = dd)
  } else {
    dd <- d
    m  <- lm(y ~ sex, data = dd)   # Japanese: sex only
  }
  broom::tidy(m, conf.int = TRUE) |> mutate(outcome = outcome, n = nrow(dd))
}
combos <- expand.grid(
  lang_slug = unique(demo_blups$lang_slug),
  outcome   = c("xi", "zeta"),
  stringsAsFactors = FALSE
)
coefs <- bind_rows(lapply(seq_len(nrow(combos)), function(i) {
  d <- filter(demo_blups, lang_slug == combos$lang_slug[i])
  res <- fit_one(d, combos$outcome[i])
  if (is.null(res)) NULL else mutate(res, lang_slug = combos$lang_slug[i])
})) |>
  filter(term != "(Intercept)")

# Random-effects meta-analysis per (outcome, term) across the languages
# that contribute that term (4 for sex, 3 for education).
meta_summary <- coefs |>
  group_by(outcome, term) |>
  group_modify(~ {
    if (nrow(.x) < 2) return(tibble())
    res <- tryCatch(
      metafor::rma(yi = .x$estimate, sei = .x$std.error,
                   method = "REML", control = list(stepadj = 0.5)),
      error = function(e) NULL)
    if (is.null(res)) return(tibble())
    tibble(estimate = as.numeric(res$beta),
           conf.low = res$ci.lb, conf.high = res$ci.ub,
           std.error = res$se, k = res$k)
  }) |>
  ungroup() |>
  mutate(lang_slug = "_meta_")

coefs_all <- bind_rows(coefs, meta_summary) |>
  mutate(
    term_label = recode(term,
      "sexMale"         = "Male\n(vs Female)",
      "scale(ed_years)" = "Maternal education\n(per SD, years)") |>
      factor(levels = rev(c("Male\n(vs Female)",
                            "Maternal education\n(per SD, years)"))),
    outcome = factor(outcome, levels = c("xi", "zeta"),
                     labels = c("Efficiency~(xi)",
                                "'Change in efficiency'~(zeta)")),
    lang_label = factor(c(LANG_LABELS, "_meta_" = "Meta-analytic")[lang_slug],
                        levels = c(LANG_LABELS[ladder$langs],
                                   "Meta-analytic")))
plot_pal <- c(setNames(unname(LANG_PAL[ladder$langs]),
                       unname(LANG_LABELS[ladder$langs])),
              "Meta-analytic" = "black")

coefs_all <- coefs_all |>
  mutate(is_meta = lang_label == "Meta-analytic")
ggplot(coefs_all,
       aes(estimate, term_label, color = lang_label,
           group = lang_label, size = is_meta)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey55") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high),
                  position = position_dodge(width = 0.70),
                  fatten = 3) +
  facet_wrap(~ outcome, ncol = 2, scales = "free_x",
             labeller = label_parsed) +
  scale_color_manual(values = plot_pal, name = NULL, drop = FALSE) +
  scale_size_manual(values = c(`FALSE` = 0.35, `TRUE` = 0.65),
                    guide = "none") +
  guides(color = guide_legend(nrow = 1)) +
  labs(x = "regression coefficient (z-scored outcome)", y = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold", size = 11),
        axis.text.y = element_text(size = 9))
```




## Changing efficiency 

```{r fig-efficiency, out.width="100%"}
#| label: fig-efficiency
#| fig-cap: ""
knitr::include_graphics(
  here("outputs", "figs", "longitudinal", "exposure_to_learn_EN.png"))
```

As an illustration of this change in efficiency, Figure 2 shows a model-derived estimate of the average number of times a child would need to hear a word before producing it (See SI: Changes in Word Efficiency), using frequency estimates from the CHILDES Corpus [@mcwhinney2020;@sanchez2019]. Across different word classes (e.g., nouns, verbs, closed class words), efficiency increases exponentially, such that early-learned words such as "ball" may be heard tens of thousands of times before they are produced, while later learned words such as "vitamin" might only be heard tens of times before being produced. 


## Estimating input 


Under a range of plausible assumptions about variation in input, all children show supra-linear scaling of growth, even accounting for accumulation dynamics.

As discussed above, one important question is how much of variance is due to variance in rate. We explore this by asking about the parameter $\pi$,

$$\pi = \frac{\sigma_\alpha^2}{\sigma_\alpha^2 + \sigma_r^2}$$

If $\pi$ > .5, then the majority of variation between individuals is not due to pure accumulation rate. 

In addition, 80-90% of between-child variation remains unexplained by input quantity. 
These conclusions are supported by analysis of longitudinal input data from BabyView (N=20) and SEEDLings (N=44) datasets, which contain child-specific estimates of input, and Peekbank (N=XYZ), which contains child-specific estimates of processing speed.

```{r fig-io-partition, fig.width=11, fig.height=4.5}
#| label: fig-io-partition
#| fig-cap: ""
# Two-panel: (A) observed vocabulary trajectories per IO study; (B) age-
# dependent variance decomposition under the additive gamma model. Panel
# B uses sigma_r, sigma_alpha, sigma_zeta, gamma at posterior median (see
# experiments.md §30 for the diagnostic / wide-delta refit).
obs <- io_summary$obs |>
  mutate(study = factor(study, levels = names(STUDY_PAL)))
p_obs <- ggplot(obs, aes(age, prop, color = study)) +
  geom_point(alpha = 0.35, size = 0.6) +
  geom_smooth(se = FALSE, method = "loess", span = 0.7, linewidth = 0.7,
              formula = y ~ x) +
  scale_color_manual(values = STUDY_PAL, name = "study") +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  labs(x = "age (months)", y = "vocabulary proportion produced",
       title = "(A) IO datasets: observed growth") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 10))

# Variance partition (additive form, mean of slope channel)
p   <- io_summary$params
a0  <- 15; ages <- seq(8, 30, length.out = 60); L <- log(ages / a0)
sr2 <- p$sigma_r^2; sa2 <- p$sigma_alpha^2
sz2 <- p$sigma_zeta_add^2; g <- p$gamma_add
parts <- tibble(age = rep(ages, 4),
                L   = rep(L,    4),
                component = factor(rep(c("Input: intercept",
                                         "Input: slope (γ)",
                                         "Efficiency: intercept",
                                         "Efficiency: slope"),
                                       each = length(ages)),
                  levels = c("Efficiency: slope", "Input: slope (γ)",
                             "Efficiency: intercept", "Input: intercept"))) |>
  mutate(var = case_when(
    component == "Input: intercept"      ~ sr2,
    component == "Input: slope (γ)"      ~ pmax(L^2 * g^2 * sr2 + 2 * L * g * sr2, 0),
    component == "Efficiency: intercept" ~ sa2,
    component == "Efficiency: slope"     ~ L^2 * sz2
  ))

p_partition <- ggplot(parts, aes(age, var, fill = component)) +
  geom_area(alpha = 0.95) +
  scale_fill_manual(values = c(
    "Input: intercept"      = "#B4D7E8",
    "Input: slope (γ)"      = "#1F77B4",
    "Efficiency: intercept" = "#003D7A",
    "Efficiency: slope"     = "#E8A0A8"), name = NULL) +
  labs(x = "age (months)",
       y = expression(Var[i](theta[it])),
       title = "(B) Variance partition (γ-additive, wide-delta)") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 10),
        legend.text = element_text(size = 8))

p_obs + p_partition + plot_layout(widths = c(1, 1))
```


Thus, modeling human learning requires positing other sources of variability and acceleration. Sources of variability could include from variation in interactional input quality, genetic variation between children, environmental richness not captured by input, etc. Sources of acceleration could include increases in processing efficiency, learning-to-learn including mutual exclusivity, shape bias, syntactic bootstrapping, as well as domain general changes in perception, processing speed, memory, attention or other aspects of cognition. The key to understanding children’s data efficiency may thus lie in better understanding when and how children’s learning accelerates with maturation and experience. 

We end by providing a quantitative comparison between within-child human learning and LLM scaling laws, showing that children’s early language learning is strikingly different from LLM scaling: it shows both steeper scaling and greater variance in growth rate.

```{r fig-llm-acceleration, fig.width=11, fig.height=4.5}
#| label: fig-llm-acceleration
#| fig-cap: ""
# Best-guess two-panel composite (will iterate). Each panel is an
# existing PNG composed via grid; replace with chunked plots once we
# decide which LLM-vs-children comparison to feature.
panel_paths <- list(
  A = here("outputs", "figs", "longitudinal",
           "chang_bergen_slope_comparison.png"),
  B = here("outputs", "figs", "schematic", "D1_scaling_disanalogy.png"))
p_to_grob <- function(path) {
  if (!file.exists(path)) return(grid::nullGrob())
  grid::rasterGrob(png::readPNG(path), interpolate = TRUE)
}
gridExtra::grid.arrange(
  p_to_grob(panel_paths$A),
  p_to_grob(panel_paths$B),
  ncol = 2
)
```



# Discussion

Coffey 2026 variance decomposition

Blingualism 

International adoption

https://bergelsonlab.fas.harvard.edu/sites/g/files/omnuum3941/files/2025-01/meylan_bergelson_2021_ARL.pdf	

:::{.acknowledgement}
Claude Code was used for literature review, code generation, visualization, and simulation infrastructure. All writing is original to the author. 
:::

<!-- :::{.suppinfo} -->


<!-- ::: -->

<!-- ## References -->