---
author:
  - name: Michael C. Frank
    affiliations: 
      - name: Department of Psychology, Stanford University

title: Efficiency and acceleration in the accumulation of children’s early vocabulary
title-short: Efficiency and acceleration 
keywords: Early word learning; Bayesian data analysis; Large language models
abstract: |
  \noindent Children typically learn hundreds of words over the first few years of their life, in a process that begins slowly but quickly picks up speed. Yet children vary widely in their rate of word learning, and this variation has important practical and theoretical consequences. Prior models of word learning describe vocabulary growth as a process of accumulation of evidence over time. Here we show that the process is best characterized as *accelerating* accumulation, with children varying in both their overall learning efficiency (intercept) and their acceleration rate (slope). This model provides a lever for understanding different loci of variation. Applied to data from 24 languages, it suggests that the female advantage in early vocabulary results from differences in overall learning efficiency, while socioeconomic differences may compound through effects on acceleration rate. Using dense observations of children's early language environment and word recognition speed, the model also suggests that differences in processing speed load on efficiency and do not compound, while differences in home language show some compounding effects on acceleration. In contrast to children, large language models -- even those trained on child-directed speech -- are not well-described by this model: they show standard accumulator dynamics with no acceleration and very limited variability. In sum, our model provides a quantitative characterization of key signatures of early vocabulary growth, providing a target for future mechanistic learning models.

bibliography: standard_model.bib
# Dataset citations appear as plain text in Table 1 (kable can't render
# [@key] markers); list their bib keys here so they still appear in the
# References. Add @marchman / @smith dataset keys once they exist in the bib.
nocite: |
  @thal2013, @simonsen2014, @hagihara2023, @long2024, @egan2025, @adams2018, @fernald2013
format:
  pdf:
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
    linestretch: 1.05
    colorlinks: true
    keep-tex: true
    # pdflscape: needed by kableExtra::landscape() (Table 1). Declared
    # explicitly because kableExtra's automatic dependency injection is
    # unreliable in Quarto (dropped on cached-chunk replay), which makes
    # the compile fail with "Environment landscape undefined."
    include-in-header:
      text: |
        \usepackage{pdflscape}
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(here); library(dplyr); library(tidyr); library(readr)
  library(ggplot2); library(patchwork); library(latex2exp)
  library(knitr); library(scales); library(stringr); library(forcats)
  library(broom); library(png); library(grid); library(gridExtra)
  library(ggrepel)
})
source(here("paper", "_helpers.R"))

XYZ <- "VALUE NOT SPECIFIED"

# load cached data — produced by `Rscript paper/build_cache.R`
# (Table 1 is built inline in its own chunk, not cached.)
CACHE <- here("paper", "cache")
ladder     <- readRDS(file.path(CACHE, "fig2_glmer_ladder.rds"))
demo_blups <- readRDS(file.path(CACHE, "blups_demographics.rds"))
io_summary <- readRDS(file.path(CACHE, "fig5_io_summary.rds"))
input3     <- readRDS(file.path(CACHE, "fig3_input.rds"))
aic        <- readRDS(file.path(CACHE, "aic_summary.rds"))
exposure   <- readRDS(file.path(CACHE, "fig4_exposure.rds"))
llm_slopes <- readRDS(file.path(CACHE, "fig6_llm_slopes.rds"))

# Inline range: how much worse the best (D) model becomes when age enters
# linearly rather than as log-age, across the four languages (dAIC of
# D_lin relative to the best model, D_log).
.dlin <- aic |> filter(model == "D_lin") |> pull(dAIC)
aic_loglin_range <- sprintf("%s--%s",
  format(round(min(.dlin)), big.mark = ","),
  format(round(max(.dlin)), big.mark = ","))

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE, cache=TRUE,
  fig.align = "center", out.width = "100%", dpi = 300
)
```


Children’s early vocabulary grows very rapidly during early childhood, with large variation between children [@fenson1994;@fenson2007;@frank2021]. The growth has theoretical, practical, and methodological importance. From a theoretical perspective, growth mechanisms shed light on children’s fundamental learning mechanisms and how they change across childhood [@mcmurray2007;@bloom2002;@frank2021]. From a practical perspective, early vocabulary is an important precursor -- and predictor -- of academic achievement [@marchman2008;@catts2002;@muter2004]. From a methodological perspective, vocabulary also affords a unique psychometric opportunity to study human variation because, unlike nearly all important human psychological metrics, there is a true zero point -- a time before which children know no words. Thus, absolute levels of variability can be calculated. 

One dominant viewpoint is that children's vocabulary growth can be characterized as a process of accumulation [@mcmurray2007; @mitchell2009; @hidaka2013; @mollica2017; @kachergis2022]. Each distinct word (type) can be conceptualized as a "bucket," and every token of language input then is a "drop" that falls into its respective bucket. When each bucket is full, the word is learned [@mcmurray2007;@kachergis2022]. These models have been used to recover interpretable differences in "bucket size," often based on word properties including concreteness and syntactic category [@hidaka2013;@kachergis2022]. They also provide a graded explanation for the widely-observed "vocabulary spurt" phenomenon -- in which children's rate of vocabulary learning appears to increase dramatically  [@frank2021;@ganger2004;@gomezdiaz2024]. 

The accumulator view connects with a viewpoint in which the quantity of language that children hear plays a strong causal role in the speed of their learning [@fernald2012]. Indeed, there is systematic evidence that variation in children's input -- both its *quantity* and its *quality* --  is related to variation in vocabulary [@hart1999;@fernald2012;@weisleder2013;@anderson2021;@rowe2012;@hirsh2015;@coffey2026]. While input quantity is relatively easy to measure, measuring what makes input high quality has been challenging. One influential framework suggests that high quality input is interactionally rich, linguistically appropriate for the level of the child, and conceptually complex [@rowe2020].  

A naïve interpretation of the accumulator view would suggest that substantial  variance in vocabulary growth should be related to input -- either via quantity (number of drops) or quality (size of each drop). Two recent meta-analyses complicate this view, however. @anderson2021 showed that only a modest proportion of the variance in vocabulary outcomes is related to input quantity ($r^2 = .11$) and quality ($r^2 = 04$). A newer meta-analysis of quantity confirms and extends the proportion of variance due to quantity to 4--7% of total variation in vocabulary [@coffey2026]. Thus -- as we might have supposed based on other research on human variation -- not all variation in vocabulary between children can be attributed to input differences. 

Further, work on children's word recognition suggests that children become faster and more accurate at matching words to their referent across early childhood [@bergelson2020;@fernald1998;@fernald2006;@peter2019;@frank2026]. Much of this work relies on the "looking while listening" (LWL) paradigm, in which children see two pictures (e.g. a book and a car) and their eye movements are recorded as they hear a sentence that refers to one of the pictures (e.g., "look at the book"). Children's reaction time to shift their eyes to the target (the book) provides an index of their speed of word recognition that decreases logarithmically across early childhood  [@fernald2008;@frank2026] and relates to their vocabulary and other language outcomes [@fernald2006;@marchman2008;@fernald2012;@fernald2013;@peter2019;@frank2026]. One hypothesis proposed in this work is that greater language input leads to more rapid processing, in a virtuous cycle in which faster processing in turn leads to greater word learning  [@weisleder2013]. In the accumulator metaphor, one way to describe this proposal would be that individual childrens' buckets can be bigger or smaller based on their prior input and their processing efficiency, but to our knowledge this suggestion has not been fleshed out. 

As this discussion indicates, despite the promise of accumulator models, they have largely not been used to understand variation between individuals. Instead, the majority of applications have focused on variation between words [@hidaka2013;@kachergis2022]. Further, the rapidity of children's early word learning stands as a rebuke to a naive accumulator view. Sometimes exposure to a small number of examples of a word in the lab is enough for a child to form a lasting  [@markson1997;@woodward1994]. In some cases, it seems only a small number of high-quality instances is sufficient for learning [@trueswell2013;@mollica2017], but if so, why do some words take much longer to learn?

The goal of the current work is to address these puzzles through systematic investigation and extension of accumulator models. All accumulator models seek to explain the same pattern: that the overall size of children's vocabulary increases supra-linearly. Where they differ is in what this growth means for the underlying learning process. Following previous work, we start with the idea that children's *internal learning ability* (not just their observed supra-linear vocabulary growth) accelerates across development [@hidaka2013;@mayor2010]. We tehn extend a prior psychometric accumulator model [@kachergis2022]: our key addition is to describe variation between individual children by modeling both their individual learning efficiency and the acceleration of their learning -- following the bucket metaphor and allowing children to have different size buckets. This extension defines a longitudinal growth model that can be fit directly to large-scale datasets about children's actual vocabularies.

To constrain our model, we use data from the MacArthur-Bates Communicative Development Inventories [MB-CDI; @fenson1994;@marchman2021], a popular parent report instrument in which parents are asked to indicate which words their child produces and understands. Here we focus exclusively on production, which is used with older children and which tends to be a more reliable measure [@frank2021]. We make use of longitudinal data from Wordbank [@frank2017], an open repository of data from the MB-CDI. 

To address questions about the relationship between language input, language processing, and growth, we fit a Bayesian version of our growth model. We make use of new data resources to estimate the magnitude of linkages between input  and vocabulary growth. First, we estimate plausible population values based on independent estimates of population variation in language input [@coffey2024]. In addition, we use measures of individual children’s language input from egocentric videos [@long2024] and day-long home recordings [@egan2025;@adams2018;@fernald2013] to estimate the input-uptake relationship more precisely for a subsample of children, and measures of LWL word recognition from Peekbank [@zettersten2023], a large scale repository of LWL data.

Overall, our analysis supports a view in which children’s vocabulary growth is not a pure process of accumulation but rather one that accelerates rapidly over time. We end by comparing the fit of our models on human data to their fit on large-language models. Making use of GPT-2 models trained on human developmental data, we show that -- in contrast to children -- these models *do* appear to function as pure accumulators. This analysis exposes a critical disanalogy between LLMs and human learners and suggests a potential route for understanding the vast "data gap" between children and large language models [@frank2023;@warstadt2023]. 

# Model 


```{r fig-schematic, fig.width=8, fig.height=6}
#| label: fig-schematic
#| fig-cap: ""
# CONSOLIDATED model figure (vocab space). [MCF: caption on next pass.]
#   (A) Top: illustrative vocabulary buildup M1->M4 — the expected
#       vocabulary curve V(t) = mean_j plogis(theta_i(t) - delta_j), summed
#       over a shared Gaussian word-difficulty distribution (McMurray 2007:
#       even a pure accumulator accelerates because theta sweeps through the
#       difficulty distribution). Same words in every panel. The theta-space
#       version of these panels now lives in the Supplement.
#   (B) Bottom: the best-fitting model (M4 / D_log) fit to the real data
#       across the five datasets — empirical grey trajectories + the
#       10/25/50/75/90 quantile fan. The full 7-model ladder is in the
#       Supplement.
# [Prior fig-glmer-ladder caption, kept for reuse: "Individual children's
#  longitudinal vocabulary growth trajectories across age in four languages.
#  The same data are plotted in each column of panels, with overlaid models
#  of increasing complexity. Transparency is higher for English (American)
#  and Norwegian panels to avoid overplotting."]

## ---- (A) illustrative vocab-space buildup (schematic) ----
ages <- seq(8, 30, length.out = 80)
A0   <- 8                                  # anchor age: log(t/A0) >= 0
L    <- log(ages / A0)
qs   <- c(0.1, 0.25, 0.5, 0.75, 0.9)
z    <- qnorm(qs)
XI0     <- -2.0                            # median intercept (log r + log alpha)
KAPPA_A <- 2.5                             # acceleration for M2-M4
DELTA   <- qnorm(ppoints(150), mean = 0, sd = 1.5)   # shared difficulties
vocab_of <- function(th) vapply(th, function(x) mean(plogis(x - DELTA)), numeric(1))

variants <- tibble::tribble(
  ~name,                       ~kappa_pop, ~sigma_alpha, ~sigma_kappa,
  "(M1) accumulator",          1,           0,            0,
  "(M2) + acceleration",       KAPPA_A,     0,            0,
  "(M3) + efficiency var.",    KAPPA_A,     1.1,          0,
  "(M4) + per-child accel.",   KAPPA_A,     1.1,          0.7
)
sim <- variants |>
  mutate(name = factor(name, levels = name)) |>
  rowwise() |>
  do({
    v <- .
    bind_rows(lapply(seq_along(qs), function(i) {
      th <- XI0 + z[i] * v$sigma_alpha + (v$kappa_pop + z[i] * v$sigma_kappa) * L
      tibble(name = v$name,
             qprob_f = factor(as.character(qs[i]),
                              levels = c("0.1","0.25","0.5","0.75","0.9")),
             age = ages, vocab = vocab_of(th))
    }))
  }) |>
  ungroup()
lwidth <- function(qf) ifelse(as.character(qf) == "0.5", 0.95, 0.55)

p_schematic <- ggplot(sim, aes(age, vocab, color = qprob_f, group = qprob_f)) +
  geom_line(aes(linewidth = lwidth(qprob_f))) +
  facet_wrap(~ name, nrow = 1) +
  scale_color_manual(values = WORDBANK_PALETTE, name = "Percentile",
                     labels = c("10th","25th","50th","75th","90th"), drop = FALSE) +
  scale_linewidth_identity() +
  scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1)) +
  labs(x = NULL, y = "expected vocab.\nproportion") +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(face = "bold", size = 8),
        legend.position = "bottom", panel.grid.minor = element_blank())

## ---- (B) best model (M4 / D_log) fit to real data, 5 datasets ----
qt_mb <- ladder$qtiles |>
  filter(model == "D_log") |>
  mutate(qprob_f = factor(qprob, levels = c(0.1,0.25,0.5,0.75,0.9),
                          labels = c("0.1","0.25","0.5","0.75","0.9")))
emp <- ladder$emp |> mutate(child_lang = paste(lang_slug, child_id, sep = "::"))
lang_lab <- ladder$lang_labels
emp$lang_label   <- factor(lang_lab[emp$lang_slug],   levels = lang_lab)
qt_mb$lang_label <- factor(lang_lab[qt_mb$lang_slug], levels = lang_lab)

p_mbest <- ggplot() +
  geom_line(data = filter(emp, language %in% c("English (American)", "Norwegian")),
            aes(age, vocab, group = child_lang),
            color = "grey25", alpha = 0.03, linewidth = 0.25) +
  geom_line(data = filter(emp, !(language %in% c("English (American)", "Norwegian"))),
            aes(age, vocab, group = child_lang),
            color = "grey25", alpha = 0.25, linewidth = 0.25) +
  geom_line(data = qt_mb,
            aes(age, vocab, color = qprob_f, group = qprob_f,
                linewidth = ifelse(as.character(qprob_f) == "0.5", 0.95, 0.55))) +
  facet_wrap(~ lang_label, nrow = 1, scales = "free_y") +
  scale_color_manual(values = WORDBANK_PALETTE, name = "Percentile",
                     labels = c("10th","25th","50th","75th","90th"), drop = FALSE) +
  scale_linewidth_identity() +
  labs(x = "Age (months)", y = "Productive vocabulary") +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(face = "bold", size = 8),
        legend.position = "bottom", panel.grid.minor = element_blank(),
        panel.spacing = unit(0.3, "lines"))

(p_schematic / p_mbest) +
  plot_layout(guides = "collect", heights = c(1, 1.1)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom")
```

To begin our analysis, we present a nested set of accumulator-type models, many of which have been presented in the literature (see SI: Comparison with Prior Models). As we show below, each refinement of the model contributes to its ability to describe the variability between children. We start with the Rasch model from psychometrics [@rasch1960;@bond2013], which assumes that each child $i$ has a learning ability $\theta_i$ and each word $j$ has a difficulty $\delta_j$.^[A variety of previous work has focused on estimating the properties of individual words  [@goodman2008;@hidaka2013;@roy2015;@braginsky2019;@kachergis2022;@kachergis2022b]. These analyses reveal that there is predictable word-level difficulty information but that only a modest portion of the variation between words is carried by frequency information. Words can be hard for a child to acquire because they vary also in the complexity and concreteness of their meanings, their syntactic category and syntactic context, their phonology, and many other factors. In the current work, we estimate word difficulty as a free parameter, and focus instead on variation between children.] The probability that child $i$ produces word $j$ is

$$P(w_{i,j}=1 \mid \theta_i, \delta_j) =\frac{\exp(\theta_i - \delta_j)}{1 + \exp(\theta_i - \delta_j)}$$
  
  
Following @kachergis2022, we can write our first model as a Rasch-style accumulator model (Model 1), specifying that $\theta_i$ is not a single parameter but instead a function of accumulation over time. In the simplest form: 
  
  $$\text{M1: }\theta_i(t)  = \log(r_i) + \log(t/a_0)$$
  
\noindent where $r_i$ is the child's rate of language input (in tokens per month) and $t$ is the child's age (in months, anchored at $a_0$); note that we use log age to simulate linear accumulation because $\exp(\theta) = r \cdot t$. This form instantiates the basic hypothesis that the only thing that changes over time is more input, and the only source of variation between children is variation in rate. 

We compare this model to an *accelerating accumulator* model (Model 2), in which the amount learned from each exposure increases with time by an exponent $\kappa$.

$$\text{M2: }\theta_i(t)  = \log(r_i) + \kappa ~ \log(t/a_0)$$
\noindent When $\kappa>1$, each successive month yields more accumulation than the last.

Neither of these models has any way to capture variation across children beyond variation in input rate. Thus, we augment the basic accelerating accumulator, rewriting it as the sum of either one or two individually variable quantities:

$$\text{M3: }\theta_i(t) = \xi_i + \kappa ~ \log(t/a_0)$$

$$\text{M4: }\theta_i(t) = \xi_i + \kappa_i ~ \log(t/a_0)$$

The intercept term in both models $\xi_i = \log(r_i) + \log(\alpha_i)$ can then decompose individual variation into both input rate-related and other variation in children's efficiency. Similarly, in M4, the slope term $\kappa_i = \kappa + \gamma \log r_i + \zeta_i$ adds *individual variation in acceleration*, decomposing acceleration into a population term, input-related acceleration, and a child‑specific residual acceleration $\zeta_i$ (the part not explained by input). The coefficient $\gamma$ then captures the strength of the relationship between input and acceleration. 

Model 4 instantiates the idea that children vary in their base rate of efficiency and their acceleration, creating variant of a standard longitudinal growth model in which children vary on their slope and intercept [@grimm2016]. Note that, while we define the relationship of $\xi_i$ and $\kappa_i$ to a child's input $r_i$, we can nevertheless fit the simplest form of the model by just allowing these parameters to vary as random child-level slopes and intercepts. 

As we will show, Model 4 provides a good description of variation in children's longitudinal growth trajectories. @fig-schematic (top) shows schematic predictions from a pure accumulator model (M1), an accelerating accumulator (M2) in comparison to models including child-level variation in efficiency (M3) and both efficiency and acceleration (M4). Each model's predictions are represented as ability scores ($\theta$) across age and then also projected into proportion correct scores on a discrete set of vocabulary items, representing an observed accuracy score that is subject to compression at floor and ceiling.

Our analysis generally depends on the availability of longitudinal data at the item level. The different models described above are not distinguishable from the cross-sectional, aggregate data that are typically considered in many analyses of early vocabulary. First, aggregate data do not allow for the estimation of individual word difficulties, thus the distribution of difficulties remains latent and different difficulty distributions can be posited to explain the accelerating shape of vocabulary growth [@mcmurray2007; @mitchell2009]. Second, cross-sectional data do not allow for identification of whether individual children's variation is due to differences in efficiency (intercept) or acceleration (slope) (see SI: Nonidentifiability in Cross-Sectional Data). 

# Results

## Children's longitudinal growth reflects differences in both efficiency and acceleration

We first retrieved longitudinal data on monolingual, typically-developing children's vocabulary from Wordbank [@frank2017], an open archive of data from the MacArthur-Bates Communicative Development Inventory and its international adaptations. SI: Datasets shows the datasets used in each study in the paper; in this first analyses, we made use of five longitudinal datasets from Wordbank (three from American English and one each from Norwegian and Japanese, $N_{total} = 3135$).  

Taking advantage of the equivalence of Rasch models to generalized linear mixed effects models [@deboeck2011], we fit the set of models shown in @fig-schematic as a series of multi-level logistic regressions. Each model predicted the probability of an individual child producing a particular word as a function of age (either unmodified in M1 or multiplied by a constant in M2) and a random effect of word difficulty. Successive models add random intercepts (M3) and slopes (M4) by child. This initial set of fits simply attempts to describe the distribution of vocabulary growth curves. 

Model 4, which incorporated variation in acceleration and efficiency, was the best fitting model across all datasets with differences in AIC $>1000$ in all cases (see SI: AIC and BIC for Longitudinal Models). @fig-schematic (bottom) shows the fit of the four model variants; the increasing spread of children's trajectories with age is especially noticeable for low-vocabulary older children [who may be classified as "late talkers"; @fernald2012; @rescorla2011].^[Following the literature, we primarily investigate linear accumulator models, which show power-law growth of ability over time because $\theta$ is a function of $\log(t)$. However, it is also possible that ability grows linearly in time, which would lead to exponential increases in the probability of word production. We fit linear-time models, but found that they were dispreferred statistically in all but one case, though differences were relatively small within the restricted age range for which we have data (See SI: Full Cross-linguistic Model Ladder).]


## Efficiency and acceleration relate to separate demographic predictors


```{r fig-demographics, fig.width=7.5, fig.height=13}
#| label: fig-demographics
#| fig-cap: "Regression coefficients predicting individual children's efficiency and acceleration as a function of gender and maternal education. Panels A and B show data from individual languages (both longitudinal and cross-sectional); blue regions show meta-analytic estimates from a random-effects meta analysis. Error bars show 95\\% confidence intervals. Panel C shows a meta-analytic summary for cross-sectional and longitudinal datasets."

# cross_sectional_demographics/composite_figure.R from that directory's
# committed cache (cache/fits.rds <- 00_build.R): sex and maternal-ed effects
# on efficiency and acceleration, per language with a meta-analysis (Panel C).
# [MCF: the caption and the two surrounding paragraphs describe the OLD
#  longitudinal-only figure and need rewriting for the composite.]
source(here("cross_sectional_demographics", "composite_figure.R"))
xsec_fits <- readRDS(here("cross_sectional_demographics", "cache", "fits.rds"))

# Inline beta/p values in the text below read from meta_summary, which is now
# the CROSS-SECTIONAL meta-analytic main effects plotted in Panel C, ordered to
# match the inline indices: [1] maternal-ed efficiency, [2] sex efficiency,
# [3] maternal-ed acceleration, [4] sex acceleration.
meta_summary <- panelC_meta(xsec_fits)
meta_summary <- meta_summary[c(
  which(meta_summary$predictor == "matEd" & meta_summary$component == "efficiency"),
  which(meta_summary$predictor == "sex"   & meta_summary$component == "efficiency"),
  which(meta_summary$predictor == "matEd" & meta_summary$component == "acceleration"),
  which(meta_summary$predictor == "sex"   & meta_summary$component == "acceleration")), ]

make_demographics_composite(xsec_fits)
```

Children's vocabulary is known to be related to a number of sociodemographic predictors, including sex and maternal education (which is interpreted as a proxy for socioeconomic status). In particular, female children show consistently larger vocabularies than male children across dozens of languages and countries [@eriksson2012;@frank2021]. Similarly, children of more educated mothers tend to show larger vocabularies, though this effect is somewhat more heterogeneous across countries and languages [@frank2021].

Our model offers a summary of children's growth in terms of both their efficiency (intercept) and their acceleration (slope). This decomposition allows us to ask whether these two demographics effects -- sex and maternal education -- are seen in efficiency, acceleration, or both. For each child in our analysis, we extracted the best linear unbiased prediction (BLUP) from our models. We then regressed intercept and slope terms against each demographic predictor separately for each language. Although we cannot fit individual children's growth trajectories in cross-sectional data, we can also in principle apply the same model to recover population parameters based on cross-sectional data, so we did so for an additional group of datasets for which sex and maternal education information were available. 

@fig-demographics shows the per-language coefficients from this analysis, along with a cross-language meta-analytic estimate for both the cross-sectional and longitudinal data (which largely agree). Examining the cross-sectional data since this analysis includes more datasets, the female advantage is apparent across all datasets, but this effect manifests itself more substantially efficiency  ($\beta = `r round(meta_summary$estimate[2],2)`$, $p = `r papaja::apa_p(meta_summary$p)[2]`$) than acceleration ($\beta = `r round(meta_summary$estimate[4],2)`$, $p = `r papaja::apa_p(meta_summary$p)[4]`$). In contrast, maternal education effects were present in both measures, but stronger in acceleration ( efficiency: $\beta = `r round(meta_summary$estimate[1],2)`$, $p = `r papaja::apa_p(meta_summary$p)[1]`$; acceleration: $\beta = `r round(meta_summary$estimate[3],2)`$, $p = `r papaja::apa_p(meta_summary$p)[3]`$). One interpretation of these findings is that sex effects reflect a small constant offset in average ability. In contrast, socioeconomic differences -- as reflected in maternal education -- might continue to compound across development [@fernald2013;@hart2003].

## Population input-related variation corresponds to meta-analytic estimates

In our prior model fits, we fit population variation in efficiency and acceleration without considering the connection to language input. We next consider whether this relationship can be estimated at a population level. We use markov chain monte carlo to fit a version of model 4 with linkages to input rate. 

We first estimate these linkages at the population level by assuming a known level of population variation in language input rate: $\sigma_r$.
We can then examine the variance due to input variation vs. population by fitting the parameter $\pi_\alpha = \frac{\sigma_\alpha^2}{\sigma_\alpha^2 + \sigma_r^2}$, which defines the proportion of population variation in efficiency due to input variation. Importantly, this parameter is directly comparable to the meta-analytic estimates reported earlier [@anderson2021;@coffey2026]. This model -- which does not include individual child-level input data -- can not distinguish input contributions to efficiency vs. acceleration, but it provides a large dataset to begin analyzing the question. 

Estimating children's input range is notoriously tricky, since the rate of language that children hear varies across cultures, families, activities, and measurement occasions [@coffey2024]. To anchor our model using a range of population values for $\sigma_r$, we surveyed the literature for plausible estimates (see SI: Population Input Range Estimation). In our main simulations we set $\sigma_r=.53$, a value drawn from @sperry2018, who estimated a mean of 419k tokens per month ($\pm$1 SD: 329k -- 957k). We provide sensitivity analysis for alternative values based on an analytic computation of the relationship between $\sigma_r$ and $\pi_\alpha$ (see SI: $\sigma_r$ Sensitivity Analysis).

@fig-io-partition shows the results of this analysis for both English and Norwegian longitudinal data (combined N=`r `XYZ`). $\pi_\alpha$ was `r XYZ` (CI: `r XYZ`) for English and `r XYZ` (CI: `r XYZ`) for Norwegian, both largely consistent with the previous meta-analyses. Sensitivity analysis showed that only implausibly large input variation would produce substantially higher values. 


## Observed input predicts both efficiency and acceleration


```{r fig-io-partition, fig.width=11, fig.height=8}
#| label: fig-io-partition
#| fig-cap: ""
# [MCF: caption on next pass.] Fig 3 -- input & processing on the efficiency
# vs acceleration channels. TOP row = the *effect* (per-child fans), in
# narrative order io-imputed -> io -> proc: (A) English + (B) Norwegian
# model-implied input (input is imputed there, so the curves are the sigma_r
# slice -- xi = mu_r + sigma_r*z_q -- NOT a data split), (C) observed io input,
# (D) processing (RT). BOTTOM row = model quantities: (E) factor x channel
# variance partition (1-pi_alpha / 1-pi_zeta), (F) sigma_r sensitivity.
# Built by paper/build_input_cache.R -> fig3_input (=input3).
ip  <- input3
pct <- scales::percent_format(accuracy = 1)

## TOP ROW (the *effect*): per-child trajectories + model curves, in narrative
## order io-imputed -> io -> proc. gcol = the per-child grouping column (EN/NO
## spaghetti is keyed on child_id; io/proc on ii).
fan_panel <- function(curves, spag, title, gcol = "ii") ggplot() +
  geom_line(data = spag, aes(age, vocab, group = .data[[gcol]]), color = "grey75",
            alpha = 0.30, linewidth = 0.22) +
  geom_line(data = curves, aes(age, vocab, color = q), linewidth = 1.05) +
  scale_color_viridis_d(end = 0.9, name = "input / processing percentile") +
  scale_y_continuous(limits = c(0, 1), labels = pct) +
  labs(x = "age (months)", y = "proportion produced", title = title) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 9))
pEN <- fan_panel(ip$panelEN_curves, ip$panelEN_spag, "(A) English — model-implied input", "child_id")
pNO <- fan_panel(ip$panelNO_curves, ip$panelNO_spag, "(B) Norwegian — model-implied input", "child_id")
pIO <- fan_panel(ip$panelD_curves, ip$panelD_spag, "(C) Observed input")
pPR <- fan_panel(ip$panelE_curves, ip$panelE_spag, "(D) Processing speed")

## BOTTOM ROW (the *quantification*).
## (E) factor x channel variance partition: input and processing shares of
## EFFICIENCY (xi) and ACCELERATION (kappa) variance. Headline -- input modest
## on both; processing loads on efficiency but ~0 on acceleration.
FACTOR_PAL <- c("Input" = "#009E73", "Processing" = "#CC79A7")
mr <- ip$meta
pPart <- ggplot(ip$panel_partition, aes(channel, med, color = factor)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = mr$lo, ymax = mr$hi,
           fill = "grey85", alpha = 0.55) +
  geom_hline(yintercept = mr$mid, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(ymin = lo, ymax = hi),
                  position = position_dodge(width = 0.45), fatten = 3, linewidth = 0.9) +
  scale_color_manual(values = FACTOR_PAL, name = NULL) +
  scale_y_continuous(labels = pct, limits = c(0, NA)) +
  labs(x = NULL, y = "share of between-child variance",
       title = "(E) Variance partition by channel") +
  theme_minimal(base_size = 9) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 9))

## (F) robustness of the input-efficiency share to the external sigma_r pin.
MODEL_PAL <- c("English (D)" = "#E69F00", "Norwegian (D)" = "#56B4E9")
pSens <- ggplot(ip$panelB_curve, aes(sigma_r, input_share, color = model)) +
  annotate("rect", xmin = ip$sr_range[1], xmax = ip$sr_range[2],
           ymin = 0, ymax = 0.28, fill = "grey85", alpha = 0.5) +
  geom_vline(xintercept = ip$sr_main, linetype = "dashed", color = "grey40") +
  geom_line(linewidth = 0.9) +
  geom_point(data = ip$panelB_anchors, aes(sigma_r, input_share, color = model),
             shape = 21, fill = "white", size = 2.2, stroke = 0.9) +
  scale_color_manual(values = MODEL_PAL, name = NULL) +
  scale_y_continuous(limits = c(0, 0.28), labels = pct) +
  labs(x = expression(sigma[r] ~ "(pinned input SD)"),
       y = "input share of efficiency", title = "(F) Sensitivity to input SD") +
  theme_minimal(base_size = 9) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 9))

top <- (pEN | pNO | pIO | pPR) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
bot <- (pPart | pSens)
top / bot + plot_layout(heights = c(1, 1.05))
```

While the population estimates of input-sensitivity given above match our expectations from meta-analyses, they do not offer a direct estimate of input couplings within individual children, and they do not allow us to estimate the separate contributions of language input to efficiency and acceleration. For this, we turn to a set of four longitudinal datasets that include dense home recordings. Three include day-long audio recordings using LENA (Language Environment Analysis) recorders, which offer an automated adult word count estimate that has been validated by transcription studies [@cristia2021]. The last includes short video recordings of egocentric video recorded in children's homes, with word counts estimated from automated transcripts [@long2024]. Each of these metrics provides a noisy proxy for total home input, but nevertheless allows us to estimate models that include individual input.
We thus fit the same model linking input estimates to variation in efficiency and acceleration, but this time with observations of $r_i$ for each child, allowing us to fit the full version of Model 4 described above, in which language input could contribute to both efficiency and acceleration. @fig-io-partition (panel C) shows a small separation of trajectories by input quartile, but the key question of interest is how this variation partitions across efficiency and acceleration.

Results for $\pi_alpha$ and $\pi_zeta$ are shown in @fig-io-partition. Overall, we observe small but highly reliable effects of language input on both efficiency and acceleration. These effects are also in the range of the meta-analyses, but they suggest that input-related differences -- though small -- could compound through their effect on acceleration. 

## Processing speed is strongly associated with efficiency 

Two of the datasets we included also include longitudinal reaction time data from the looking-while-listening paradigm (LWL), archived in the Peekbank repository [@zettersten2023].

We fit a modification of our model to this subset of the data (N=XYZ). Efficiency and acceleration terms included *both* input and LWL reaction time as predictors. 



## Growth in ability leads to rapid changes in learning times for individual words

```{r fig-efficiency, fig.width=7, fig.height=4}
#| label: fig-efficiency
#| fig-cap: ""
# Per-word "exposures to 50% production" for a typical English child,
# from the M_best fit + CHILDES frequencies (data in fig4_exposure.rds;
# see paper/build_cache.R section 5 and model/scripts/exposure_to_learn.R).
# Each point is a word; x = predicted age at 50% production, y = cumulative
# exposures of that word by that age (log scale). Lines: per-class lm fits.
items <- exposure$items
meta  <- exposure$meta

# label the extreme word per class on log-exposure (balanced across classes)
n_to_label <- 3
  
labels_df <- items |>
  mutate(log_N = log10(N_word)) |>
  group_by(lexical_class) |>
  arrange(log_N) |>
  slice(c(1:n_to_label, (n() - n_to_label):n())) |>
  ungroup()

ggplot(items, aes(t_50, N_word, colour = lexical_class)) +
  geom_point(alpha = 0.55, size = 1.1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7, formula = y ~ x) +
  geom_label_repel(data = labels_df, aes(label = item),
                   size = 2.8, alpha = 0.95, show.legend = FALSE,
                   box.padding = 0.4, max.overlaps = Inf,
                   min.segment.length = 0) +
  scale_y_log10(labels = scales::label_number(big.mark = ","),
                breaks = c(100,1000,10000,100000), 
                limits = c(100, 1000000)) +
  scale_x_continuous(limits = c(10, 40)) +
  scale_colour_brewer(palette = "Dark2", name = "Class") +
  labs(x = "Age (months) when 50% of children produce word",
       y = "Cumulative exposures at AoA") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(size = 9, colour = "grey25"),
        legend.position = "bottom")
```

As an illustration of the changes in efficiency and acceleration that we observe in our data, we computed the average proportion of exposures necessary for a child to learn a word across ages. @fig-efficiency shows a model-derived estimate of the average number of times a word is heard before more than 50% of children produce it (See SI: Changes in Word Efficiency), using frequency estimates from the CHILDES Corpus [@macwhinney2000;@sanchez2019]. 

Across different word classes (e.g., nouns, verbs, closed class words), efficiency increases exponentially, such that early-learned words such as "book" may be heard tens of thousands of times before they are produced, while later learned words such as "vitamin" might only be heard hundreds times before being produced. Indeed, this analysis likely understates the degree of change in efficiency as it assumes that each exposure to a word is equally meaningful. In fact, individual exposures vary widely in how much information they give. Sometimes words are overheard rather than directed to the child, or lack physical context to allow for clear inference [@effective_learning_instances_erika;@trueswell2014;@mollica2017]. Thus, in line with both intuition and experiments, the total number of exposures required on average to learn a word might indeed be only a small handful by age three [@markson1997;@bloom2002].

## Large language models do not show acceleration in their vocabulary growth


```{r fig-llm-acceleration, fig.width=7, fig.height=4}
#| label: fig-llm-acceleration
#| fig-cap: "Comparison of acceleration slopes (kappa) estimated in our experiments for children in English and Norwegian, vs. LLM slopes per model estimated by Chang & Bergen (2022) and in our experiments."
# Children's per-child kappa_i vs. LLM per-word sigmoid slopes (Chang &
# Bergen 2022), placed on a common "logit per natural-log experience"
# axis. Data cached by paper/build_cache.R section 6 (fig6_llm_slopes.rds):
# `sl` is the per-instance long frame; `sm` the per-population summary
# (median, 95% interval, n) available for inline numbers.
sl <- llm_slopes$slopes
sm <- llm_slopes$summary

slope_pal <- c("Children (English)"   = "#c41e37",
               "Children (Norwegian)" = "#fb9a99",
               "BERT"   = "#1f78b4", "GPT-2"  = "#33a02c",
               "BiLSTM" = "#a6cee3", "LSTM"   = "#b2df8a")

ggplot(sl, aes(slope_natural, fill = label, colour = label)) +
  geom_density(alpha = 0.35, linewidth = 0.5) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey55",
             linewidth = 0.4) +
  annotate("text", x = 1.4, y = 0.5, label = "kappa = 1\n(unit accumulator)",
           hjust = 0, vjust = 0.5, size = 3.0, colour = "grey45") +
  scale_fill_manual(values = slope_pal, name = NULL) +
  scale_colour_manual(values = slope_pal, name = NULL) +
  coord_cartesian(xlim = c(0, 22)) +
  labs(x = "slope on log(experience): logit per natural-log unit",
       y = "density") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

Small-scale language models trained on developmentally-realistic corpora have become a useful tool for understanding early language development [@huebner2022;@futtrell2026;@frank2025]. For our final experiments, we were interested in whether large language models show the same patterns of developmental change in efficiency or acceleration. We made use of work by @chang2022, who extracted per-word acquisition curves from language models. They computed logistic fits to the surprisal (negative log probability) of individual words across training, estimating 

We can compare 

@evanston2023 investigated 

We end by providing a quantitative comparison between within-child human learning and LLM scaling laws, showing that children’s early language learning is strikingly different from LLM scaling: it shows both steeper scaling and greater variance in growth rate.

# Discussion

What we did

Longitudinal growth models efficiency and acceleration. Two different paramters. 

Where does acceleration come from? Could come from a variety of sources, including general maturation of memory and attention, specific changes in word recognition or processing [@fernald1998;@frank2026], the increasing sophistication of inferential strategies for word learning [@smith2002;@markman1988;etc], or other sources. 

@mitchell2009 prove that acceleration does not result from simply leverage of vocabulary -- 

Role of input: Coffey 2026 variance decomposition


Furthermore, how does the accumulation process relate to the rapid growth of vocabulary? On some models, accumulation by itself can produce an exponential “vocabulary explosion” while the child’s own learning abilities remain constant. On other accounts, however, the child’s own abilities change over development. These changes could be due to the accumulation of language allowing bootstrapping (e.g., via mutual excusivity or the shape bias), they could be through faster processing (Fernald rich get richer), or they could be via general maturation (e.g., better memory, faster processing). Or all of these. 

Thus, modeling human learning requires positing other sources of variability and acceleration. Sources of variability could include from variation in interactional input quality, genetic variation between children, environmental richness not captured by input, etc. Sources of acceleration could include increases in processing efficiency, learning-to-learn including mutual exclusivity, shape bias, syntactic bootstrapping, as well as domain general changes in perception, processing speed, memory, attention or other aspects of cognition. The key to understanding children’s data efficiency may thus lie in better understanding when and how children’s learning accelerates with maturation and experience. 


Continuous variation or clusters? [@pinstock2026;@singletary2025] We think it's just continuous variation (Supplement)

Network models and their predictions e.g. [@steyvers2005;@hills2009;@fourtassi2020]

<!-- Blingualism  -->

<!-- International adoption -->

<!-- https://bergelsonlab.fas.harvard.edu/sites/g/files/omnuum3941/files/2025-01/meylan_bergelson_2021_ARL.pdf	 -->


# Methods

## Data and Code

Data were retrieved from Wordbank using the `wordbankr` package [@frank2017], childes-db VER using `childesr` [@sanchez2019], and peekbank version 2026.1 using `peekbankr` [@zettersten2023]. BabyView data are from release 2025.2, specifically those used by @tan2026. SEEDLings data are downloaded from the materials provided by @egan2025.

Reproducible code for this paper is available at http://github.com/langcog/standard-model-2. 

## Models

### Frequentist models 

Initial models were fit with `glmer` [@glmer], with the following specifications:

* M1: `produces ~ offset(log_age) + (1 | item)` 
* M2: `produces ~ 1 + log_age + (1 | item)` 
* M3: `produces ~ 1 + log_age + (1 | item) + (1 | child)` 
* M4: `produces ~ 1 + log_age + (1 | item) + (log_age | child)` 

Linear variants reported in the supplement used linear age rather than log age. 

### Bayesian models

Bayesian models were fit using Stan [@stan].

Priors

Inference

### Input

### GPT-2 models

We trained XYZ

# References

:::{.acknowledgement}
Claude Code was used for literature review, code generation, visualization, and simulation infrastructure. All writing is original to the author. 
:::

<!-- :::{.suppinfo} -->


<!-- ::: -->

{{< supplemental.qmd >}}