---
author:
  - name: Michael C. Frank
    affiliations: 
      - name: Department of Psychology, Stanford University

title: Children, but not language models, learn words via accelerating accumulation 
#Children but not language models learn vocabulary via accelerating accumulation
#Efficiency and acceleration in the accumulation of children’s early vocabulary
title-short: Accelerating accumulator models
keywords: Early word learning; Bayesian data analysis; Large language models
abstract: |
  \noindent Children learn hundreds of words over the first few years of their life, in a process that begins slowly but quickly picks up speed.  Prior models of word learning describe vocabulary growth as a process of accumulation of evidence over time. Here we show that the process is best characterized as *accelerating* accumulation, with children varying in both the overall efficiency with which they accumulate vocabulary and their acceleration rate. In contrast to children, language models -- even those trained on child-directed speech -- are not well-described by this model: they show standard accumulator dynamics with no acceleration and very limited variability. Children learn using orders of magnitude less training data than language models do, and our findings suggest a potential explanation. Children, but not language models, "learn to learn", leveraging the knowledge they have to accelerate future learning. 

bibliography: standard_model.bib
# Dataset citations appear as plain text in Table 1 (kable can't render
# [@key] markers); list their bib keys here so they still appear in the
# References. Add @marchman / @smith dataset keys once they exist in the bib.
nocite: |
  @thal2013, @simonsen2014, @hagihara2023
format:
  pdf:
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
    linestretch: 1.05
    colorlinks: true
    keep-tex: true
    # pdflscape: needed by kableExtra::landscape() (Table 1). Declared
    # explicitly because kableExtra's automatic dependency injection is
    # unreliable in Quarto (dropped on cached-chunk replay), which makes
    # the compile fail with "Environment landscape undefined."
    include-in-header:
      text: |
        \usepackage{pdflscape}
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(here); library(dplyr); library(tidyr); library(readr)
  library(ggplot2); library(patchwork); library(latex2exp)
  library(knitr); library(scales); library(stringr); library(forcats)
  library(broom); library(png); library(grid); library(gridExtra)
  library(ggrepel)
})
source(here("paper", "_helpers.R"))

# Cached data for the three main-text figures — produced by
# `Rscript paper/build_cache_short.R`.
CACHE <- here("paper", "cache")
fig1_fan   <- readRDS(file.path(CACHE, "fig1_fan.rds"))        # Fig 1: M0-M3 schematic + per-dataset M3 fan
fig2       <- readRDS(file.path(CACHE, "fig2_efficiency.rds")) # Fig 2: per-word exposures-to-learn (M3)
llm_slopes <- readRDS(file.path(CACHE, "fig6_llm_slopes.rds")) # Fig 3: child vs LLM acceleration

# Inline text values (raw numbers in si_inline.rds, built by build_cache_short
# section 7); formatted here to the precision they render at in the text.
inl <- readRDS(file.path(CACHE, "si_inline.rds"))
.k1  <- function(x) sprintf("%.1f", x)                        # kappa: one decimal
.pct <- function(x) if (x >= 10) sprintf("%.0f", x) else sprintf("%.1f", x)
age_lo   <- inl$age_lo; age_hi <- inl$age_hi                  # integer months
loo_min  <- format(signif(inl$loo_min, 2), big.mark = ",")   # 2 sig figs
kap_lo   <- .k1(inl$kappa_lo); kap_lo_q5 <- .k1(inl$kappa_lo_q5); kap_lo_q95 <- .k1(inl$kappa_lo_q95)
kap_hi   <- .k1(inl$kappa_hi); kap_hi_q5 <- .k1(inl$kappa_hi_q5); kap_hi_q95 <- .k1(inl$kappa_hi_q95)
en_kappa <- .k1(inl$en_kappa); no_kappa <- .k1(inl$no_kappa)
en_sd    <- .k1(inl$en_sd);    no_sd    <- .k1(inl$no_sd)
qc_mar   <- .pct(inl$qc_marchman); qc_nor <- .pct(inl$qc_norwegian)

knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE, cache=TRUE,
  fig.align = "center", out.width = "100%", dpi = 300
)
```


Children’s early vocabulary grows rapidly during early childhood [@frank2021;@kidd2020;@fenson1994], and this "explosion" of expressive vocabulary is a signature human cognitive achievement [@bloom2002;@spelke2003]. Yet children's vocabulary also varies widely between individuals [@frank2021],
and the relative size of early vocabulary is a predictor of later academic success [@marchman2008;@catts2002;@muter2004]. Providing quantitative characterization of this process is an important scientific challenge [@frank2009;@mcmurray2012;@trueswell2013;@kachergis2012;@jones2017;@goodman2008;@hidaka2013;@roy2015;@braginsky2019;@kachergis2022;@kachergis2022b]

Until recently, language acquisition could only be studied in human learners. Language models (LMs) show increasing linguistic sophistication, however, allowing them to become a model system for the study of human language [@futrell2025;@frank2025]. The most successful LMs differ from human learners in an important way, however: they are trained on vastly more data than human learners receive during childhood. While even open-source models are routinely trained on trillions of words of data [@gao2020], children hear on the order of 100k -- 1M words of input per month, with 36M words as an upper bound by age 3 [@frank2023;@warstadt2023]. Why do models require so much more training data to achieve even basic linguistic competence [@hu2024]? 

Here we study this "data gap" by comparing children and LMs on their vocabulary growth. The number of words a child produces is a concrete indicator, and children's early vocabulary is very tightly coupled to their grammatical development and other aspects of their early language [@bates1994;@marchman1994;@frank2021]. Further, total vocabulary for young children can be approximated effectively using parent report forms, such as the MacArthur Bates Communicative Development Inventories (MB-CDI) [@marchman2021]. CDIs are a widely-used, reliable, and valid method for getting a snapshot of children's early vocabulary  [@fenson1994;@fenson2007;@marchman2021;@frank2021]. We take advantage of Wordbank, an open access database of cross-linguistic parent report data on the vocabulary of children ages 8 -- 36 months [@frank2017].

We begin by developing a psychometric model of vocabulary growth in children -- an "accelerating accumulator" model. This model shows good fit to both individual growth trajectories and their variation across children. Next, we develop methods for measuring the acceleration of vocabulary growth in language models. Making use of transformer language models trained on human developmental data, we show that -- in contrast to children -- these models function as pure accumulators, with no acceleration and very limited variability. This analysis exposes a critical disanalogy between LMs and human learners. 

# An accelerating accumulator model of vocabulary growth

Recent models describe children's vocabulary growth as a process of accumulation [@mcmurray2007; @mitchell2009; @hidaka2013; @mollica2017; @kachergis2022]. Each distinct word (type) can be conceptualized as a "bucket," and every token of language input then is a "drop" that falls into its respective bucket. When each bucket is full, the word is learned [@mcmurray2007;@kachergis2022]. These models have been used to recover interpretable differences in "bucket size," often based on word properties including concreteness and syntactic category [@hidaka2013;@kachergis2022]. They also provide a graded explanation for the widely-observed "vocabulary spurt" phenomenon -- in which children's rate of vocabulary learning appears to increase dramatically [@frank2021;@ganger2004;@gomezdiaz2024]. 

We begin with the model of @kachergis2022, an accumulator over ratio-scale variables with real units: words (tokens) per hour, total words (types) in the child's vocabulary, and reaction times in word processing. The absolute magnitude of these variables -- not just their variability -- can be measured and compared across studies and relations can be estimated and compared between them. Further, the presence of a true zero for each means that they can be log transformed, which is critical for capturing the shape of their growth over development [@frank2026;@kail1991]. These features distinguish our modeling approach from the wide variety of models of early word learning that typically establish either schematic or variability-based linkages to developmental data [@frank2009;@mcmurray2012;@trueswell2013;@kachergis2012]. 

We start with the Rasch model from psychometrics [@rasch1960;@bond2013], which assumes that each child $i$ has a learning ability $\theta_i$ and each word $j$ has a difficulty $\delta_j$.^[A variety of previous work has focused on estimating the properties of individual words  [@goodman2008;@hidaka2013;@roy2015;@braginsky2019;@kachergis2022;@kachergis2022b]. These analyses reveal that there is predictable word-level difficulty information but that only a modest portion of the variation between words is carried by frequency information. Words can be hard for a child to acquire because they vary also in the complexity and concreteness of their meanings, their syntactic category and syntactic context, their phonology, and many other factors. In the current work, we estimate word difficulty as a free parameter, and focus instead on variation between children.] The probability that child $i$ produces word $j$ is: 

$$P(w_{i,j}=1 \mid \theta_i, \delta_j) =\frac{\exp(\theta_i - \delta_j)}{1 + \exp(\theta_i - \delta_j)}$$
  
We then describe the child's ability as

  $$\theta_i(t) = \xi_i + \kappa_i ~ \log(t/a_0) + \log(H)$$
\noindent which is contains three terms, an intercept $\xi_i$, an acceleration term, $\kappa_i ~ \log(t/a_0)$ that describes the rate of accumulation growing exponentially in the log of the child's age, and a constant offset $H$ for the number of hours over which accumulation occurs (See SI: Generative Model). This is our "accelerating accumulator" model, in which accumulating language ``counts for more'' later in development. Children vary on both their initial rate of accumulation $\xi_i$ and their acceleration $\kappa_i$, creating variant of a standard longitudinal growth model in which children vary on their slope and intercept [@grimm2016]. Our model is closest to the formulations of @hidaka2013 and @kachergis2022, but has not yet been described in the literature (see SI: Comparison with Prior Models).

Under this model, we can use the mean value of $\kappa$ to distinguish pure accumulator behavior ($\kappa \approx 1$) and accelerating accumulator behavior ($\kappa > 1$). Note that it is critical to distinguish between supra-linear growth in children's actual vocabulary (in terms of count) -- a pattern that is robustly observed across children worldwide [@frank2021;@fenson1994] -- and the idea that children's *internal learning ability* accelerates across development, which is the substantive claim we test here.

# Fitting the accumulator model 

Our analysis generally depends on the availability of longitudinal data at the individual word level. Aggregate data do not allow for the estimation of word difficulties $\delta$, thus their distribution remains latent; in this regime different difficulty distributions can be posited to explain the accelerating shape of vocabulary growth [@mcmurray2007; @mitchell2009]. In addition, cross-sectional data do not allow for identification of whether individual children's variation is due to differences in efficiency (intercept) or acceleration (slope) (see SI: Non-identifiability in Cross-Sectional Data). 

Thus, to constrain our model, we use data from CDIs archived in Wordbank. Here we focus exclusively on production, which is used with older children and which tends to be a more reliable measure [@frank2021]. We downloaded data from Wordbank, focusing on five longitudinal datasets of monolingual, typically-developing children ages `r age_lo`--`r age_hi` (three from American English and one each from Norwegian and Japanese, $N_{total} = sum(fig1$meta$n_kids)$; see SI: Datasets). 

We fit the accelerating accumulator model to each of these datasets using Bayesian inference in Stan [@stancite]. To test whether each model component was necessary, we additionally fit three model ablations, removing individual variation in acceleration (removing $\kappa_i$), accumulation (removing $\xi_i$), and finally acceleration entirely (removing $\kappa$). 
The accelerating accumulator provided the best fit to children's growth trajectories (@fig-schematic, see SI: Full Model Comparison Results), with substantial differences in cross-validated predicted probability between the accelerating accumulator and the next best model (LOO ELPD $>`r loo_min`$ for all five datasets). For all datasets, estimates of acceleration were high, robustly rejecting the pure accumulator at $\kappa$=1: $\kappa = $ `r kap_lo` [`r kap_lo_q5`, `r kap_lo_q95`] -- `r kap_hi` [`r kap_hi_q5`, `r kap_hi_q95`]. 

We considered linear growth in ability with the logarithm of age. Pure linear growth is also possible, but linear time models were dispreferred statistically (see SI: Linear vs. Logistic Age). Further, the increasing spread of children's trajectories with age is especially noticeable for low-vocabulary older children [who may be classified as "late talkers"; @fernald2012; @rescorla2011], but variation was unimodal across all samples (see SI: Characterizing Variation in Efficiency and Acceleration) [@pinstock2026;@singletary2025]. 


```{r fig-schematic, fig.width=8, fig.height=6}
#| label: fig-schematic
#| fig-cap: "(A) Schematic predictions of all four models for vocabulary growth across age. Pure accumulator (M1) and accelerating accumulator (M2) models predict no variability across children, while M3 predicts variation and M4 predicts an increasing spread across chidren. (B) Best fitting model (M4) for five datasets. Gray lines show individual children's trajectories; colored lines show bootstrapped predictions at standard quantiles. Transparency is higher for panels with denser data to avoid overplotting."

## Fig 1 precomputed in paper/build_cache_short.R (fig1_fan.rds):
## (A) illustrative M0-M3 schematic; (B) per-dataset Bayesian M3 model-implied
## fan over empirical trajectories. Norwegian appears once its M3 fit lands.
pal <- unname(WORDBANK_PALETTE)

p_schematic <- ggplot(fig1_fan$schematic, aes(age, vocab, color = qf, group = qf)) +
  geom_line(aes(linewidth = ifelse(qf == "50th", 0.95, 0.55))) +
  facet_wrap(~ name, nrow = 1) +
  scale_color_manual(values = pal, name = "Percentile", drop = FALSE) +
  scale_linewidth_identity() +
  scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1)) +
  labs(x = NULL, y = "expected vocab.\nproportion") +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(face = "bold", size = 8),
        legend.position = "bottom", panel.grid.minor = element_blank())

p_fits <- ggplot() +
  geom_line(data = fig1_fan$spag,
            aes(age, prop, group = interaction(lang, child)),
            color = "grey25", alpha = 0.12, linewidth = 0.22) +
  geom_line(data = fig1_fan$fan,
            aes(age, vocab, color = qf, group = qf,
                linewidth = ifelse(qf == "50th", 0.95, 0.55))) +
  facet_wrap(~ lang, nrow = 1) +
  scale_color_manual(values = pal, name = "Percentile", drop = FALSE) +
  scale_linewidth_identity() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Age (months)", y = "proportion of items produced") +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(face = "bold", size = 8),
        legend.position = "bottom", panel.grid.minor = element_blank(),
        panel.spacing = unit(0.3, "lines"))

(p_schematic / p_fits) +
  plot_layout(guides = "collect", heights = c(1, 1.1)) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom")
```


# Growth in ability leads to rapid changes in learning times for individual words

```{r fig-efficiency, fig.width=7, fig.height=5}
#| label: fig-efficiency
#| fig-cap: "Estimated cumulative exposures to a word plotted by the age at which 50% of the children in our three English-language datasets were estimated to produce the word. Colors show different word classes, and lines show simple linear fits."

items <- fig2$items
k <- 8
.vis <- items |> filter(N_word >= 50, N_word <= 1e6, t_50 >= 10, t_50 <= 42)
labels_df <- bind_rows(
  slice_min(.vis, t_50,   n = k),   # far left  (earliest)
  slice_max(.vis, t_50,   n = k),   # far right (latest)
  slice_min(.vis, N_word, n = k),   # bottom    (fewest exposures)
  slice_max(.vis, N_word, n = k)) |># top       (most exposures)
  distinct(item, .keep_all = TRUE)
ggplot(items, aes(t_50, N_word, colour = lexical_class)) +
  geom_point(alpha = 0.5, size = 1.1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8, formula = y ~ x) +
  ggrepel::geom_text_repel(data = labels_df,
                           aes(label = sub("[ ]*\\(.*\\)$", "", word)), size = 2.7,
                           max.overlaps = Inf, min.segment.length = 0,
                           segment.size = 0.2, show.legend = FALSE) +
  scale_y_log10(labels = scales::label_number(big.mark = ","),
                breaks = c(100, 1000, 10000, 100000),
                limits = c(50, 1000000)) +
  scale_x_continuous(limits = c(10, 42)) +
  scale_colour_brewer(palette = "Dark2", name = NULL,
                      labels = c("Nouns", "Action Words", "Descriptive Words", "Function Words")) +
  labs(x = "Age (months) when 50% of children produce word",
       y = "Cumulative exposures at age of acquisition") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

The accelerating accumulator model provides a quantitative description of a very salient fact about language acquisition: the number of exposures necessary to learn a word decreases dramatically over childhood. While a toddler may have heard a word like "book" thousands of times before they say it, a few years later a preschooler may hear "ibex" a handful of times on a trip to the zoo and repeat it the next day. To quantify this phenomenon, we computed the average proportion of exposures necessary for a child to learn a word across ages (focusing here on English, the language for which we had the most data). 

We computed each word's age of acquisition (AoA) $A$ from our fitted models, with AoA defined as the point at which 50% of children produce the word -- in other words, the age at which child ability and word difficulty balance: $A(j) = a_0 \exp\!\big((\delta_j - \log H - \mu_\xi)/\kappa\big)$. Then for each word, we made an estimate of the cumulative word frequency for an individual child based on frequency estimates from the CHILDES archive, a large set of transcripts of speech to children [@macwhinney2000;@sanchez2019]  (See SI: Changes in Word Efficiency).

@fig-efficiency shows the relationship between cumulative exposures and age of acquisition. Across different word classes (e.g., nouns, verbs, and function words words), the number of examples needed for learning drops exponentially with development.^[Indeed, this analysis likely understates the degree of change in efficiency as it assumes that each exposure to a word is equally meaningful. In fact, individual exposures vary widely in how much information they give. Sometimes words are overheard rather than directed to the child, or lack physical context to allow for clear inference [@cartmill2013;@trueswell2013;@mollica2017].] Thus, in line with both intuition and experiments, the total number of exposures required on average to learn a new noun might indeed be only a small handful by age three [@markson1997;@bloom2002].

# Large language models show no acceleration in their vocabulary growth

```{r fig-llm-acceleration, fig.width=7, fig.height=4}
#| label: fig-llm-acceleration
#| fig-cap: "The density of acceleration slopes for language models (by word) and children (by child), extracted from English and Norwegian models."

# Children's per-child kappa_i vs. LLM per-word sigmoid slopes (Chang &
# Bergen 2022), placed on a common "logit per natural-log experience"
# axis. Data cached by paper/build_cache.R section 6 (fig6_llm_slopes.rds):
# `sl` is the per-instance long frame; `sm` the per-population summary
# (median, 95% interval, n) available for inline numbers.
sl <- llm_slopes$slopes
sm <- llm_slopes$summary

# Children: by-child kappa_i (the acceleration signature). LMs: by-word
# sigmoid slope, three populations -- Chang & Bergen (2022): 4 architectures
# (LSTM, BiLSTM, GPT-2, BERT) on a BookCorpus+WikiText-103 corpus, training
# axis; our CHILDES GPT-2 on the training axis; and our CHILDES ladder on the
# DISTINCT-INPUT (development) axis. All three cluster at the kappa = 1
# unit-accumulator boundary; children sit ~10x higher.
slope_pal <- c("Children (English)"              = "#c41e37",
               "Children (Norwegian)"            = "#fb9a99",
               "LMs: C&B 2022 (4 architectures)" = "#8856a7",
               "LMs: CHILDES (training)"         = "#2c7fb8",
               "LMs: CHILDES (development)"      = "#41ab5d")

ggplot(sl, aes(slope_natural, fill = group, colour = group)) +
  geom_density(alpha = 0.32, linewidth = 0.5) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey55",
             linewidth = 0.4) +
  annotate("text", x = 1.6, y = 0.92, hjust = 0, vjust = 1, size = 3.0,
           colour = "grey45",
           label = "kappa = 1\nunit accumulator\n(no acceleration)") +
  # annotate("segment", x = 2.7, xend = 1.15, y = 1.2, yend = 1.2,
  #          arrow = arrow(length = unit(0.12, "cm")), colour = "#2c7fb8") +
  # annotate("text", x = 2.8, y = 1.2, hjust = 0, vjust = 0.5, size = 2.8,
  #          colour = "#2c7fb8", label = "all LMs cluster here") +
  scale_fill_manual(values = slope_pal, name = NULL) +
  scale_colour_manual(values = slope_pal, name = NULL) +
  coord_cartesian(xlim = c(0, 22)) +
  labs(x = "kappa: slope on log(experience) (logit per natural-log unit)",
       y = "density") +
  theme_minimal(base_size = 11) +
  theme(legend.position = c(0.70, 0.60),
        legend.background = element_rect(fill = "white", colour = NA))
```

Small-scale language models trained on developmentally-realistic corpora have become a useful tool for understanding early language development [@huebner2021;@futrell2025;@frank2025;@evanson2023;@misra2024]. One major disanalogy between humans and language models, however, is the massive scale of training data necessary for language models to achieve high levels of performance on linguistic tasks [@warstadt2023;@frank2023;@hu2024]. In our final set of analyses, we hypothesized that this data gap might be related to a lack of acceleration in the learning process. 

We begin by building a bridge between the well-known "scaling law" for large language models [@hoffmann2022;@kaplan2020] and the model we describe here for vocabulary growth. The form of one popular scaling law is that the loss $L$ of a neural network language model on a dataset size $D$ is $L = E + \frac{B}{D^\beta}$ (holding the size of the the network constant), where $B$ and $\beta$ are the free parameters of a power law and $E$ is the entropy of the text being modeled. @michaud2023 propose that this power law scaling behavior can emerge from a set of independently learned "quanta" that are themselves power-law distributed, much like the words of natural language [@zipf1949]. 

Following this idea, we explore the emergence of individual words in neural language models. If their emergence indeed follows the kind of power-law dynamics described by @michaud2024, then it would follow that their accumulation *decelerates* rather than accelerating (as we have shown in children). In particular, each further order of magnitude of data should produce a smaller gain in word knowledge. This finding is in contrast to children, who we have shown grow linearly in the logarithm of age. Since children receive relatively constant levels of input across development 

To trace the emergence of individual words in language models, we follow @chang2022, who fit sigmoidal curves for individual words on the trajectory of surprisal values across training. These sigmoids providing a direct analogue to the logistic curves in the Rasch model we used with children. The slope of the sigmoid for each word is directly analogous to our acceleration parameter $\kappa$ as it captures how fast evidence for that word accumulates with training data (see SI: Computing Acceleration in LMs).

First, we extracted slopes from fits provided by @chang2022, in which four different language model architectures (LSTM, BiLSTM, GPT-2, and BERT) were fit to standard corpora (BookCorpus and WikiText-103). Next, to compare to models fit to developmental data, we fit GPT-2 to the full set of English transcripts in CHILDES [@feng2024]. We used GPT-2 for this analysis [@radford2019] as it has become the preferred model system for such investigations [@feng2024;@misra2024;@evanson2023;@choshen2022]. In both @chang2022's models and ours, $\kappa$ clustered around 1 (see @fig-llm-acceleration, A): knowledge about individual words accumulated gradually. In contrast, across all five of our samples, children's $\kappa$ values were both high and highly variable. 

However, this analysis cannot show acceleration directly because learning is shown across training passes through the same dataset. To isolate developmental change, we trained a separate set of GPT-2 models each to convergence on increasingly larger subsamples of the same CHILDES data, allowing us to estimate surprisal as a function of actual amount of input data -- a more direct developmental analogue. This analysis revealed strong support for the scaling laws analysis above -- when we plotted $\kappa$ values on a log-log plot, they were precisely linear, showing the predicted power law structure (@fig-llm-acceleration, B; see SI: Relation to Scaling Laws). 

In sum, neural scaling laws suggest that language models decelerate in their acquisition of individual quanta of knowledge -- including words. Our fitted, child-scale models confirm this regularity. This deceleration stands in fundamental contrast to the acceleration we robustly observe in human learners.  

# Discussion

How does children's vocabulary grow over early childhood? We proposed that this process is characterized as accelerating accumulation, in which variation between children is produced by differences in both the efficiency of the accumulation process and the rate of acceleration. This cognitive process leads to a simple longitudinal growth model that provides a good description of cross-linguistic data. In contrast to children, language models follow neural scaling laws, meaning that they decelerate across training rather than accelerating. 

Where does the acceleration in children's learning come from? One possibility is that it is driven by developmental maturation of general abilities such as learning, memory, and attention -- such changes could mean that children's "buckets" require fewer drops to fill them. 

Another possibility is that children's increasingly sophisticated inferential abilities allow them to extract progressively more information from the linguistic signal [@smith2002;@markman1988;@bohn2019;meylan2022]. 

@mitchell2009 studied how these sorts of "leverage" mechanisms can change the slope of acceleration in accumulator models. Network growth models provide one promising avenue for understanding how leverage might play out in specific patterns of vocabulary growth at the level of word content [@hills2009;@fourtassi2020].

# Methods

## Data and Code

Data were retrieved from Wordbank using the `wordbankr` package [@frank2017], childes-db VER using `childesr` [@sanchez2019].

Reproducible code for this paper is available at http://github.com/langcog/standard-model-2. 

All longitudinal datasets with three or more observations per child were included. We excluded a small number of children from the English (Marchman) and Norwegian datasets,(`r qc_mar`% and `r qc_nor`&, respectively, due to suspicious observations in which there was a decline in vocabulary of more than 20% in some longitudinal observations. 

## Bayesian models

Bayesian models were fit using Stan [@stan].


### Neural network language models

Following @feng2024, we fit \texttt{GPT-2-small} (124M parameters). Training was performed on NVIDIA A40 GPUs.

# Acknowledgements

Some of the computing for this project was performed on the Sherlock and Marlowe clusters. We would like to thank Stanford University and Stanford Research Computing for providing computational resources and support that contributed to these research results. Funds for computational experiments were also provided by a gift from Meta to support research on children's home language environment. Claude Code was used for literature review, code generation, visualization, and simulation infrastructure. 

# References

::: {#refs}
:::


<!-- :::{.suppinfo} -->


<!-- ::: -->

```{=latex}
\newpage
\begin{center}
{\LARGE\bfseries Supplemental Information}
\end{center}
```

{{< include supplemental.qmd >}}