```{r si-setup, include=FALSE}
suppressPackageStartupMessages({
  library(here); library(dplyr); library(tidyr); library(ggplot2)
  library(patchwork); library(latex2exp); library(knitr); library(kableExtra)
})
source(here("paper", "_helpers.R"))
si_blups  <- readRDS(here("paper", "cache", "si_blups.rds"))   # per-child BLUPs (M3): xi, kappa
si_loo    <- readRDS(here("paper", "cache", "si_loo.rds"))     # LOO ladder (M0-M3)
si_loglin <- readRDS(here("paper", "cache", "si_loglin.rds"))  # log- vs linear-age (M3)
fig2      <- readRDS(here("paper", "cache", "fig2_efficiency.rds"))  # word efficiency + delta validation
si_ds     <- readRDS(here("paper", "cache", "si_datasets.rds"))      # datasets table (5 M3 bundles)
si_gamlss <- readRDS(here("paper", "cache", "si_gamlss.rds"))        # parametric vs GAMLSS overlay + per-child kappa
si_set    <- readRDS(here("paper", "cache", "si_settings.rds"))      # robustness: 2+/3+/mega x 5 datasets
si_pooled <- attr(si_set, "pooled")                                 # converged mega hyperparameters (or NULL)
llm_slopes <- readRDS(here("paper", "cache", "fig6_llm_slopes.rds")) # kappa slopes incl. C&B per-arch (SI: other architectures)
si_2pl    <- readRDS(here("paper", "cache", "si_2pl.rds"))           # 1PL vs 2PL: discrimination, kappa, AoA, LOO
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", out.width = "100%", dpi = 300
)
```

# Comparison with Prior Models

Our accelerating accumulator model has not been described in the literature in the exact form we use, but it is closely related to other models. Here we describe these models as special cases of our model. We write the latent ability of child $i$ on word $j$ at age $t$ as

$$\eta_{i,j}(t) = \log r_i + \log \alpha_i + \kappa_i \log t - \delta_j,$$
where $r_i$ and $\alpha_i$ decompose the term $\xi_i$ in our model: $r_i$ is the input rate for the child and $alpha_i$ is their learning efficiency. Several prior accumulator models can then be seen as special cases in which one or more of these components is fixed:

* @mcmurray2007 is a pure accumulator in which $\kappa=1$ and children do not differ from one another ($\sigma_{\alpha}=\sigma_{\kappa}=0$), so both input rate and efficiency reduce to a constant, corresponding to our M0. 
* @hidaka2013 introduce a model in which ability grows as $t^{D+1}$, creating acceleration. This exponent is exactly our $\kappa$, and this model is equivalent to our M1.
* @kachergis2022 present our M2: they allow $\alpha$ to vary across individuals, but not $\kappa$. 

In addition, several other models in the literature can be described as variants of this framework:

* @mitchell2009 augment the pure M0 accumulator with a "leverage" term (in which words learned earlier can facilitate the learning of later words). Rhis term shifts the slope the learning function. While it does not map cleanly onto our M1, the dynamics are similar. 
* @mollica2017 define a pure accumulator model with $kappa=1$, but with an additional parameter relating to the speed of per-word accumulation. This model maps onto a combination of our M0 with the a 2-parameter logistic variant that we fit below.

```md
[draft for promotion — MCF to review. Proposed replacement for the @mitchell2009 bullet
 above, which currently reads: "...Rhis term shifts the slope the learning function. While
 it does not map cleanly onto our M1, the dynamics are similar."

 Verified against papers/mitchell_mcmurray_2009.pdf. Their abstract: "leverage can change
 both the shape and timing of the acceleration, but ... it cannot create acceleration if it
 did not exist in the corresponding model without leveraging", and their Zipf analysis
 "confirm[s] that leveraging does not create acceleration". Their Fig 1A plots "Words
 Known" from the MCDI long form, so their acceleration is in OBSERVED VOCABULARY COUNT,
 not latent ability -- exactly the distinction the main text draws. The current bullet
 states the opposite of their central finding. (Also fixes the "Rhis" typo.)]

* @mitchell2009 augment the pure M0 accumulator with a "leverage" term, in which words
learned earlier facilitate the learning of later words. Their central result is a negative
one: leverage changes the shape and timing of vocabulary growth, but it cannot create
acceleration that is not already present in the distribution of word difficulties. In our
coordinates it therefore alters when words are acquired without changing
$\kappa_{\text{pop}}$, and so it does not correspond to our M1. Note also that their
acceleration is defined over observed vocabulary counts rather than over latent ability.
```


# Non-identifiability in Cross-Sectional Data

Since acceleration is a property of change over time. A single cross-sectional observation from an individual child cannot identify the child's acceleration. As @fig-cross-sectional-confound shows, a single observation from a child could be accounted for by variation in either $\xi_i$ (efficiency) or $\kappa_i$ (acceleration). 

```{r fig-cross-sectional-confound, fig.width=9, fig.height=4}
#| label: fig-cross-sectional-confound
#| fig-cap: "Non-identifiability of growth models in cross-sectional data. Panels show the same simulated data, reflecting the possibility of stable between-child intercept differences (A) vs. varying growth slopes (B)."
# Why longitudinal data are needed to identify per-child acceleration.
#
# A single cross-sectional snapshot — one ability estimate per child at
# their observed age — is consistent with qualitatively different
# individual-trajectory structures. We draw ONE set of cross-sectional
# points (identical in both panels) and overlay two generative stories
# that each pass through every point exactly:
#   (A) stable between-child differences: every child shares the
#       population slope and differs only in intercept (parallel lines);
#   (B) diverging trajectories: every child starts from a common early
#       ability and differs in slope (a fan).
# With one observation per child the two are indistinguishable — only
# repeated measurement of the same child reveals which holds.

set.seed(42)
N         <- 28
a0        <- 6                       # reference age; log(t/a0) > 0 over 8-30
kappa_pop <- 2.5                     # population acceleration (slope on log-age)
sigma_xi  <- 1.0                     # between-child intercept spread

ages_obs <- runif(N, 8, 30)
L_obs    <- log(ages_obs / a0)
# cross-sectional cloud: ability rises with log-age + between-child noise
theta_obs <- kappa_pop * L_obs + rnorm(N, 0, sigma_xi)
pts <- data.frame(age = ages_obs, theta = theta_obs)

age_grid <- seq(8, 30, length.out = 60)
L_grid   <- log(age_grid / a0)

# (A) parallel: each child same slope kappa_pop, intercept set to hit point
parallel <- bind_rows(lapply(seq_len(N), function(i) {
  xi <- theta_obs[i] - kappa_pop * L_obs[i]
  data.frame(child = i, age = age_grid, theta = xi + kappa_pop * L_grid)
}))

# (B) fanning: common early origin theta0, slope set to hit point
theta0 <- min(theta_obs) - 1.5       # origin below the cloud so all slopes > 0
fanning <- bind_rows(lapply(seq_len(N), function(i) {
  kappa_i <- (theta_obs[i] - theta0) / L_obs[i]
  data.frame(child = i, age = age_grid, theta = theta0 + kappa_i * L_grid)
}))

yr <- range(theta_obs) + c(-1, 1)
mk <- function(traj, ttl) {
  ggplot() +
    geom_line(data = traj, aes(age, theta, group = child),
              color = "#2b8cbe", alpha = 0.35, linewidth = 0.3) +
    geom_line(data = data.frame(age = age_grid,
                                theta = kappa_pop * L_grid +
                                        (mean(theta_obs) - kappa_pop * mean(L_obs))),
              aes(age, theta), linetype = "dashed",
              color = "grey45", linewidth = 0.4) +
    geom_point(data = pts, aes(age, theta), size = 1.3, color = "grey15") +
    coord_cartesian(ylim = yr) +
    labs(x = "age (months)", y = expression(ability~theta), title = ttl) +
    theme_minimal(base_size = 10) +
    theme(plot.title = element_text(size = 10, face = "bold"),
          panel.grid.minor = element_blank())
}

mk(parallel, "(A) stable between-child differences (intercepts vary)") +
mk(fanning,  "(B) diverging trajectories (slopes vary)")
```

# Datasets

@tbl-datasets gives the characteristics of each dataset used. In the main text we focus on children with three or more administrations (for more precise estimates of $\kappa$), but below we also confirm the robustness of our claims including children wtih two or more administrations. 

```{r tbl-datasets}
#| label: tbl-datasets
#| tbl-cap: "Characteristics of all datasets used in the paper. N and ages reflect the analysis samples (post-exclusion). Adm/Chi gives the median administrations per child."

# Built by paper/build_cache_short.R (section 6, si_datasets.rds) from the five
# longitudinal M3 bundles -- the single source now that the io/proc/cross-sectional
# analyses are cut (replaces the retired build_table1.R / table1_datasets.csv).
# Citation column is plain text (kable can't render [@key]); bib keys live in the
# YAML `nocite:`.
# N column now reports BOTH the main-text analysis sample (>=3 administrations, the
# subset where per-child acceleration is identified) and the full longitudinal sample
# (>=2 administrations), as "n_3+ / n_2+". This makes the exclusion legible: it is a
# modeling requirement, not a small dataset (Marchman is 162 of 2,136). n_kids_all
# comes from the 2+ bundles (build_cache_short.R, si_datasets.rds).
# [MCF: (a) Thal/Smith/Marchman citation years are 20XX placeholders (@smith/
#  @marchman keys still missing from the bib). (b) the caption still describes the OLD
#  multi-analysis table AND now needs to define the "N (3+/2+)" column -- rewrite for
#  the 5 longitudinal datasets. (c) the both-N column addresses the earlier
#  N-reconciliation note: the 3+ analysis total is 1,841; the 2+ total is 4,922.]
si_ds |>
  transmute(citation, language,
            n = sprintf("%s / %s", format(n_kids, big.mark = ","),
                        format(n_kids_all, big.mark = ",")),
            n_admins = format(n_admins, big.mark = ","),
            mean_age = sprintf("%.1f", mean_age),
            min_age = as.character(min_age), max_age = as.character(max_age),
            med_admins = as.character(med_admins)) |>
  knitr::kable(
    col.names = c("Citation", "Language", "N (3+/2+)", "Admins",
                  "Mean", "Min", "Max", "Adm/Chi"),
    align = "llrrrrrr") |>
  kableExtra::add_header_above(c(" " = 4, "Age (months)" = 3, " " = 1)) |>
  kableExtra::kable_styling(font_size = 9, latex_options = "HOLD_position")
```

@fig-qc shows the full spaghetti plots for each dataset, highlighting in red those observations that failed our quality check and were excluded. We speculate that these observations are due to either data entry issues or faulty correspondences between observations. Such issues occur frequently in archival datasets and affect only a small portion of our data. 

```{r fig-qc}
#| label: fig-qc
#| fig-width: 11
#| fig-height: 3.6
#| fig-cap: "Data-quality exclusions. Red lines show children excluded from the analysis entirely, with red dots marking flagged administrations; grey lines are a random sample of retained children, shown for reference."
#
# What the colours mean (checked against studies/bayes_long/qc_exclusion_report.R):
#   red  = `bad`, i.e. children with FEWER THAN TWO administrations surviving the
#          filter, so they leave the analysis altogether (62 of 4,984 = 1.2%);
#   grey = a random SAMPLE of retained children (N_SPAG per dataset), not all of them.
# So this figure shows the dropped children, NOT all 140 removed administrations --
# administrations flagged inside a child who still survives are not drawn.
TR <- readRDS(here("paper", "cache", "qc_spaghetti_data.rds"))
ggplot() +
  geom_line(data = subset(TR, !bad), aes(age, v, group = ckey), color = "grey70", alpha = 0.20, linewidth = 0.25) +
  geom_line(data = subset(TR,  bad), aes(age, v, group = ckey), color = "#d7191c", alpha = 0.70, linewidth = 0.40) +
  geom_point(data = subset(TR, bad & !keep), aes(age, v), color = "#d7191c", size = 0.7) +
  facet_wrap(~dataset, nrow = 1) + scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Age (months)", y = "Proportion of items produced") +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(face = "bold"), panel.grid.minor = element_blank())
```

# Full Model Comparison Results

@tbl-loo gives the full leave-one-out model comparison results across models for both the dataset with 3+ administrations/child and the larger dataset with 2+ administrations/child. 

```{r tbl-loo}
#| label: tbl-loo
#| tbl-cap: "Bayesian leave-one-out (LOO) model comparison. Each cell is the difference in expected log predictive density (ELPD) relative to the best model (M3), with the standard error of the difference in parentheses; more negative is worse. Every step of the ladder improves fit in every dataset, and the full model M3 (per-child efficiency *and* acceleration) is preferred throughout. Built from the per-fit LOO objects (paper/build_cache_short.R)."
short <- c("English (Thal)" = "Thal", "English (Smith)" = "Smith",
           "English (Marchman)" = "Marchman", "Norwegian" = "Norwegian",
           "Japanese" = "Japanese")
mod_desc <- c(M0 = "M0. accumulator ($\\kappa=1$)",
              M1 = "M1. + acceleration",
              M2 = "M2. + efficiency var.",
              M3 = "M3. + acceleration var.")
lang_order <- levels(si_loo$lang)
cell <- function(m, e, s) ifelse(m == "M3", "0",
  sprintf("%s (%s)", format(round(e), big.mark = ","), format(round(s))))
# Both thresholds are shown: the >=3-admin ladder (main text, where per-child slopes
# are identified) and the full >=2-admin ladder. They are stacked with pack_rows --
# 4 model rows per threshold, 5 dataset columns -- rather than doubling the columns.
# M3 wins every rung at both thresholds. (Raw ELPD scales with n_obs, so the 2+ gaps
# are larger mainly because the 2+ samples are larger; the per-observation comparison
# lives in the settings table.)
# [MCF: the caption should note the table now stacks both admin thresholds.]
tab <- si_loo |>
  mutate(txt = cell(as.character(model), elpd_diff, se_diff)) |>
  arrange(threshold, model) |>
  select(threshold, model, lang, txt) |>
  tidyr::pivot_wider(names_from = lang, values_from = txt) |>
  mutate(Model = mod_desc[as.character(model)])
n3 <- sum(tab$threshold == "3+")
knitr::kable(tab |> select(Model, all_of(lang_order)), format = "latex",
             align = c("l", rep("r", length(lang_order))), escape = FALSE,
             col.names = c("Model", unname(short[lang_order]))) |>
  kableExtra::add_header_above(
    c(" " = 1, "$\\\\Delta$ELPD vs M3 (SE)" = length(lang_order)), escape = FALSE) |>
  kableExtra::pack_rows("$\\geq$3 administrations (main text)", 1, n3, escape = FALSE) |>
  kableExtra::pack_rows("$\\geq$2 administrations (full longitudinal)",
                        n3 + 1, nrow(tab), escape = FALSE) |>
  kableExtra::kable_styling(latex_options = "scale_down", font_size = 8)
```

# Comparison to a Non-Parametric Quantile Model

To test whether the accelerating accumulator appropriately fits the distribution of growth curves across children, we compared against a strong baseline: the flexible non-parametric GAMLSS family that is used to produce normative curves for the CDI [@marchman2021]. Specifically, for each dataset we fit a GAMLSS beta-regression of the proportion of words produced by the child's age. We used a a penalized spline function on the mean and on the scales.

@fig-gamlss-overlay shows that across all five datasets the M3 fit (blue) tracks both the non-parametric GAMLSS fit (orange) and the empirical quantiles (circles). Thus, the the accelerating accumulator
is a faithful, low-dimensional compression of the data's quantile structure.

```{r fig-gamlss-overlay, fig.width=13, fig.height=3.5}
#| label: fig-gamlss-overlay
#| fig-cap: "M3 fits (blue, drawn from fitted per-child acceleration) vs. non-parametric GAMLSS beta-regression (orange), with empirical 10th/50th/90th percentiles (open circles). Lines are the 10/50/90 percentiles of each model. Color denotes model."

ll <- si_gamlss$overlay$lines |> dplyr::filter(model != "M3 (MVN draw)") |>
  tidyr::pivot_longer(c(q10, q50, q90), names_to = "pct", values_to = "v") |>
  dplyr::mutate(dataset = factor(dataset, levels = si_gamlss$levels))
empd <- si_gamlss$overlay$emp |> dplyr::mutate(dataset = factor(dataset, levels = si_gamlss$levels))
ggplot() +
  geom_point(data = empd, aes(age, v), shape = 1, size = 1.2, stroke = .6, color = "grey25") +
  geom_line(data = ll, aes(age, v, color = model, group = interaction(model, pct)), linewidth = .7) +
  facet_wrap(~dataset, nrow = 1) +
  scale_color_manual(values = c("M3 (fitted-κ)" = "#2c7fb8", "GAMLSS (auto-λ)" = "#d95f02"), name = NULL) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Age (months)", y = "Proportion of items produced") +
  theme_minimal(base_size = 9) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(), strip.text = element_text(face = "bold"))
```

# Robustness Across Model and Data Settings

In the main text, we report fits to children 3+ administrations with separate models fit to each dataset, but our parameter estimates were substantially similar across two other settings: first, separate fits to children with 2+ administrations, and second, a full hierarchical model fit to children with 3+ administrations. @tbl-settings shows LOO fits and the robustness of the $\kappa$ estimates across these different settings. Including the
two-administration children inflated the acceleration variance $\sigma_b$ with
measurement error rather than genuine heterogeneity, leading to slightly noisier fits to the distribution, but population $\kappa$ was extremely similar. The pooled model reproduced the parameters of the independent fits closely, but convergence was slow and we do not report intervals from this model.

```md
## How the pooled model shares information across datasets  [draft for promotion — MCF to review]

[Requested description of the hierarchical/pooling structure. Numbers are cache-derived
 (si_settings.rds); the structure is m_pool.stan. Suggested placement: directly after the
 paragraph above, before @tbl-settings.]

The pooled model fits all five datasets jointly, but it does not simply concatenate them.
Children and words remain nested within their own dataset, and each dataset keeps its own
mean efficiency and mean acceleration, so no dataset is pushed toward another's
population values. What is shared is the *scale* of between-child variation. The
dataset-level standard deviations of efficiency and acceleration are themselves given a
distribution, $\log \sigma_{a[d]} \sim \mathcal N(m_a, s_a)$ and
$\log \sigma_{b[d]} \sim \mathcal N(m_b, s_b)$, so a dataset with few children has its
estimated spread drawn toward the typical value across datasets, while a dataset with
many children is left essentially unconstrained. Pooling on the log scale keeps the
standard deviations positive and makes the shared quantity a typical *ratio* rather than a
typical difference. The efficiency-acceleration correlation and the item-difficulty
distribution are estimated separately within each dataset and are not pooled, since these
concern different instruments and different languages.

This structure yields one quantity the separate fits cannot provide: a typical
between-child acceleration standard deviation across samples,
$\exp(m_b) = 5.13$ (90% CI $[3.90, 6.63]$), whose interval reflects variation between
datasets rather than only within them. In practice the shrinkage it induces is very
small. Every per-dataset $\sigma_b$ moves by at most 1% relative to its independent fit
(3.19 to 3.22, 5.14 to 5.13, 6.60 to 6.59, 5.65 to 5.70, 5.45 to 5.40). This is itself
informative: partial pooling only pulls estimates together when they disagree relative to
their uncertainty, so the near-absence of shrinkage indicates the five samples already
agree about how much children differ in acceleration.
```


```{r tbl-settings}
#| label: tbl-settings
#| tbl-cap: "For each dataset and setting: number of children (N), population acceleration ($\\kappa$), between-child acceleration SD ($\\sigma_b$), and the per-observation LOO advantage of the full model M3 over M2 ($\\Delta$ELPD per item response)."
# Per-observation gap is the only cross-setting-comparable LOO quantity (raw ELPD
# scales with n_obs). Mega rows carry no M2 rung -> gap is "---".
# The asterisk that used to flag pooled rows with slow per-dataset scale mixing
# (max_rhat > 1.1) has been removed per MCF; max_rhat is still carried in
# si_settings.rds if it is ever wanted back.
sset <- si_set |>
  mutate(
    Setting = dplyr::recode(as.character(setting),
      "2+ separate" = "$\\geq$2 separate",
      "3+ separate" = "$\\geq$3 separate (main text)",
      "mega (3+)"   = "pooled ($\\geq$3)"),
    N   = format(n_kids, big.mark = ","),
    kap = sprintf("%.1f", kappa),
    sb  = sprintf("%.2f", sigma_b),
    gap = ifelse(is.na(gap_per_obs), "---", sprintf("%.4f", gap_per_obs)))
grp <- table(droplevels(si_set$dataset))                 # rows per dataset, factor order
knitr::kable(sset |> select(Setting, N, kap, sb, gap), format = "latex",
             col.names = c("Setting", "N", "$\\kappa$", "$\\sigma_b$",
                           "$\\Delta$ELPD/obs (M2$\\to$M3)"),
             align = c("l", "r", "r", "r", "r"), escape = FALSE) |>
  kableExtra::pack_rows(index = setNames(as.integer(grp), names(grp)), escape = FALSE) |>
  kableExtra::kable_styling(font_size = 9, latex_options = "HOLD_position")
```

# Linear vs. Logistic Age

We additionally fit M3 variants with linear, rather than logarithmic, growth in acceleration over time. @tbl-loglin shows that the linear model was dispreferred across all five datasets. 

```{r tbl-loglin}
#| label: tbl-loglin
#| tbl-cap: "Log-age vs linear-age accumulation. LOO ELPD advantage of the log-age model (M3) over its linear-age counterpart, with SE; positive favors log-age."
short <- c("English (Thal)" = "Thal", "English (Smith)" = "Smith",
           "English (Marchman)" = "Marchman", "Norwegian" = "Norwegian",
           "Japanese" = "Japanese")
tab <- si_loglin |>
  mutate(Dataset = unname(short[as.character(lang)]),
         adv = sprintf("%s (%s)", format(round(-elpd_diff), big.mark = ","),
                       format(round(se_diff)))) |>
  select(Dataset, adv)
knitr::kable(tab, format = "latex", align = c("l", "r"), escape = FALSE,
             col.names = c("Dataset", "$\\Delta$ELPD (log $-$ linear)")) |>
  kableExtra::kable_styling(font_size = 9, latex_options = "HOLD_position")
```

# 2PL Model Comparison

In the main analysis we use a Rasch (1PL) item model, in which every word has a difficulty but all words have equal discrimination between children. @kachergis2022b found that a 2PL model, which adds a per-word discrimination $\lambda_j$, was preferred over the 1PL for CDI data. We therefore refit M3 with a per-word discrimination parameter.

@tbl-2pl shows the results of this comparison. The 2PL does genuinely fit better, with substantial differences in LOO ELPD for every language. However, model results are otherwise unchanged. Estimates of $\kappa$ increase somewhat, but they are not directly comparable across item models. In normal IRT models, ability scores are constrained to a standard normal distribution, but in our model, abilities grow over time so there is no single standard scale, and the addition of the discrimination parameter can change the overall scale. For this reason, we report $\log \kappa$, since a pure change of units would be a constant shift. In addition, across individual words, we find that the predicted age of acquisition (the age at which the word is predicted to have a 50% probability of being known) is almost identical, suggesting that the models are highly comparable. 

Critically, $\delta$ (difficulty) and $lambda$ (discrimination) have a very low correlation. If harder words were substantially more discriminating, this correlation could inflate estimates of $\kappa$ in the 1PL model; in practice we find little support for this hypothesis.

```{r tbl-2pl}
#| label: tbl-2pl
#| tbl-cap: "Rasch (1PL) versus two-parameter logistic (2PL) item models, M3 refit on the same data. Discrimination: SD of $\\log \\lambda_j$, its 10th-90th percentile range on the natural scale, and its correlation with item difficulty (90% interval). Acceleration: population $\\kappa$ under each item model and the difference in $\\log \\kappa$. Age of acquisition: correlation and median absolute difference in per-word $t_{50}$, which is invariant to the discrimination scale. LOO: expected log predictive density of the 1PL relative to the 2PL, with SE; negative favours the 2PL."
## Everything comes from paper/cache/si_2pl.rds (studies/bayes_long/03_compare_1pl_2pl.R).
## The table renders whatever datasets have BOTH fits; rerun that script to fill in any
## still pending. [MCF: Norwegian's 2PL was still running when this was written.]
##
## Column choices worth knowing:
##  - d log kappa, not the kappa ratio: a pure change of units is one multiplicative
##    factor, i.e. a CONSTANT additive shift in log kappa. Constancy across datasets is
##    therefore the diagnostic for "units" vs "the 2PL revises the growth story".
##  - t_50 rather than kappa for the substantive comparison: solving theta(t) = delta_j
##    gives log(t_50/a0) = (delta_j - log_H - mu_xi)/kappa, in which lambda_j does not
##    appear at all, and which is invariant under the lambda rescaling. It is also the
##    quantity Fig 2 is built from.
##  - LOO sign: elpd_diff is the loser's deficit, and the 2PL wins everywhere so far,
##    so the printed value is the 1PL's deficit (negative).
s2 <- si_2pl$summary
tab2 <- s2 |>
  transmute(
    Dataset  = as.character(lang),
    sig_lam  = sprintf("%.2f", sigma_lambda),
    lam_rng  = sprintf("%.2f--%.2f", lambda_p10, lambda_p90),
    cor_ld   = sprintf("%+.2f [%+.2f, %+.2f]", cor_lambda_delta,
                       cor_lambda_delta_q5, cor_lambda_delta_q95),
    k1       = sprintf("%.1f", kappa_1pl),
    k2       = sprintf("%.1f", kappa_2pl),
    dlk      = sprintf("%+.3f", d_log_kappa),
    aoa_r    = sprintf("%.3f", aoa_cor),
    aoa_d    = sprintf("%.2f", aoa_mad),
    loo      = ifelse(is.na(loo_elpd_diff), "---",
                      sprintf("%s (%s)", format(round(loo_elpd_diff), big.mark = ","),
                              format(round(loo_se_diff)))))
## format="latex" is REQUIRED whenever a kable is piped into kableExtra here. Under
## Quarto, knitr::opts_knit$get("out.format") is "markdown", so kable() emits a pipe
## table and kable_styling() converts it to LaTeX -- escaping as it goes and discarding
## escape=FALSE. Without it, math in col.names/cells renders as the literal source
## ("\$\textbackslash{}kappa\$"). Setting the format explicitly makes escape=FALSE stick.

knitr::kable(tab2, format = "latex", align = c("l", rep("r", 9)), escape = FALSE,
             col.names = c("Dataset",
                           "SD $\\log\\lambda$", "$\\lambda$ p10--p90", "$r(\\lambda,\\delta)$",
                           "1PL", "2PL", "$\\Delta\\log\\kappa$",
                           "$r(t_{50})$", "med.\\ $|\\Delta t_{50}|$",
                           "$\\Delta$ELPD (SE)")) |>
  kableExtra::add_header_above(c(" " = 1, "Discrimination" = 3, "Acceleration $\\\\kappa$" = 3,
                                 "Age of acquisition" = 2, "LOO" = 1), escape = FALSE) |>
  kableExtra::kable_styling(font_size = 7, latex_options = c("HOLD_position", "scale_down"))
```

# Variation in Acceleration

```{r blup-prep, include=FALSE}
# Per-child BLUP distributions (efficiency xi, acceleration kappa) from the
# Bayesian M3 fits (per-child exports), used for the dip-test table and histogram
# below. We z-score WITHIN dataset (datasets differ in scale) and drop a handful
# of extreme BLUPs (|z| > 4) that come from sparsely-observed children rather
# than distinct subgroups.
bz <- si_blups |>
  group_by(lang) |>
  mutate(xz = as.numeric(scale(xi)), zz = as.numeric(scale(kappa))) |>
  ungroup() |>
  filter(is.finite(xz), is.finite(zz), abs(xz) <= 4, abs(zz) <= 4) |>
  rename(study = lang)
```

To explore the distribution of individual children's fitted parameters, we show the best linear unbiased predictor (BLUP) for each child (@fig-blup-hist) as well as the random effects of the fitted models (@tbl-ranef). While the distribution appears generally normal, we do not give a strong interpretation as this shape may in part be due to the normal form of the model's random effect structure. However, across all datasets, acceleration appeared right-skewed. 

```{r fig-blup-hist, fig.width=7, fig.height=6.5}
#| label: fig-blup-hist
#| fig-cap: "Histograms of per-child BLUP estimates, z-scored within each dataset, for both efficiency and acceleration parameters. Dashed lines indicate the standard normal reference distribution."

hist_df <- bz |>
  select(study, xz, zz) |>
  pivot_longer(c(xz, zz), names_to = "param", values_to = "z") |>
  mutate(param = factor(param, levels = c("xz", "zz"),
                        labels = c("Efficiency~(xi)", "Acceleration~(kappa)")))
ggplot(hist_df, aes(z)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = "#2b8cbe", colour = "white", linewidth = 0.1) +
  stat_function(fun = dnorm, colour = "grey30", linetype = "dashed",
                linewidth = 0.4) +
  facet_grid(study ~ param, scales = "free_y",
             labeller = labeller(param = label_parsed)) +
  labs(x = "per-child BLUP (z-scored within dataset)", y = "density") +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(size = 8, face = "bold"),
        panel.grid.minor = element_blank())
```


```{r tbl-ranef}
#| label: tbl-ranef
#| tbl-cap: "Random-effect parameters for M3 fits, by dataset: population acceleration $\\kappa$, between-child SD in efficiency ($\\sigma_a$) and acceleration ($\\sigma_b$), their correlation ($\\rho$), and the percentage of children estimated above the pure-accumulator value $\\kappa = 1$. Brackets give 90% posterior intervals."

ranef <- readRDS(here("paper", "cache", "si_ranef.rds"))
ci <- function(m, lo, hi, d = 2)
  sprintf(paste0("%.", d, "f [%.", d, "f, %.", d, "f]"), m, lo, hi)
ranef |>
  transmute(Dataset = as.character(lang), N = format(n_kids, big.mark = ","),
            kappa   = ci(kappa,  kappa_q5,  kappa_q95,  1),
            sigma_a = ci(sigma_a, sigma_a_q5, sigma_a_q95),
            sigma_b = ci(sigma_b, sigma_b_q5, sigma_b_q95),
            rho     = sprintf("%+.2f [%+.2f, %+.2f]", rho, rho_q5, rho_q95),
            gt1     = sprintf("%.1f", pct_gt1)) |>
  ## format="latex" required whenever a kable is piped into kableExtra here: under
  ## Quarto knitr's out.format is "markdown", so kable() emits a pipe table and
  ## kable_styling() converts it to LaTeX, escaping as it goes and discarding
  ## escape=FALSE. This chunk lost the argument when it moved from the main text.
  knitr::kable(format = "latex", align = c("l", "r", "r", "r", "r", "r", "r"), escape = FALSE,
               col.names = c("Dataset", "N", "$\\kappa$", "$\\sigma_a$", "$\\sigma_b$",
                             "$\\rho$", "\\% $\\kappa_i>1$")) |>
  kableExtra::kable_styling(font_size = 8, latex_options = c("HOLD_position", "scale_down"))
```

# Input Estimation

We make use of estimates of children's language input in both @fig-efficiency and @fig-llm-acceleration, panel C. We calculate average language input per month by estimating words per hour and multiply by an estimated number of waking hours per month. Word tokens per hour estimates come from 42 children with manual transcript data
in two prior studies with manual transcripts [@sperry2019;@hart1995; estimates used also in @kachergis2022]. Across these studies, the pooled geometric mean is $\exp(7.338) \approx 1{,}537$ word tokens per hour
of child-directed (any-adult) speech, with between-child standard deviation of
log rate $\sigma_r = 0.534$. We multiply these estimates by an assumed 12 waking exposure-hours per day to get a monthly average number of hours ($12 \times 30.4 \approx 365$), giving 
$1{,}537 \times 365 \approx 561\text{K}$ tokens per month. A child's average cumulative input at age $t$ months is then $t \times 561\text{K}$ under a constant-rate assumption.

In @fig-llm-acceleration, panel C, we give uncertainty bands on representative children based on $\pm 1$ SD of the
between-child log input rate. A child one standard deviation above (below) the
median rate reaches the same age having heard $\times 1.71$ ($\div 1.71$) as
many tokens. This results in a horizontal shift of the child's curve along the log-token axis.

# Computing Acceleration in LMs

To compare acceleration in LMs and children, we link the $\kappa$ parameter from our accelerating accumulator model to the word-by-word learning performance of LMs. We first focus on the slope for individual words rather than individual children. Starting with the M3, differentiating $\eta_{ij}(t)$ (the probability of producing a particular word) with respect to $\log t$ gives $\kappa_i$:
the logit-probability that a child produces a word rises linearly in
log-age with slope equal to the child's scaling exponent. Thus, the per-child
sigmoid slope *is* $\kappa_i$, by construction.

Next, to estimate the slope of acquisition of individual words, @chang2022 compute surprisal for each word across LM training checkpoints and then fit a four-parameter logistic curve expressing how surprisal decreasess. Expressed as the fraction of a word's surprisal reduction achieved so far, their fit is a sigmoid in
$\log_{10}(\text{training steps})$ with slope $1/\text{scale}_w$.
Dividing by $\ln 10$ converts this to natural-log units — the same units
as $\kappa_i$ — so the two slopes can be read off the same axis.

# Acceleration in Other Architectures

The acceleration results we report in the main text are not specific to our choice of the GPT-2 architecture. We re-analyze the published fits of @chang2022. @fig-si-cb-arch shows a similar distribution of per-word $\kappa$ values for the four different architectures reported in that study, which were trained on standard LM corpora rather than child-directed input. 

```{r fig-si-cb-arch, fig.width=6.5, fig.height=4}
#| label: fig-si-cb-arch
#| fig-cap: "Per-word acquisition slopes for the four architectures studied in Chang & Bergen (2022) -- BERT, GPT-2, BiLSTM, LSTM -- trained on BookCorpus + WikiText-103. CHILDES-trained GPT-2 models (training axis) are overlaid in grey."

# Data: build_cache.R §6 (fig6_llm_slopes.rds: slopes_cb_arch + slopes).
cb   <- llm_slopes$slopes_cb_arch
ours <- llm_slopes$slopes |>
  dplyr::filter(group == "LMs: CHILDES (training)") |>
  dplyr::transmute(arch = "GPT-2 / CHILDES (this work)", slope_natural)
arch_pal <- c("BERT"="#7b3294","GPT-2"="#c2a5cf","BiLSTM"="#a6dba0","LSTM"="#008837",
              "GPT-2 / CHILDES (this work)"="grey35")
ggplot(dplyr::bind_rows(cb, ours), aes(slope_natural, colour=arch, fill=arch)) +
  geom_density(alpha=0.15, linewidth=0.6) +
  geom_vline(xintercept=1, linetype="dashed", colour="grey55", linewidth=0.4) +
  annotate("text", x=1.15, y=Inf, hjust=0, vjust=1.5, size=2.8, colour="grey45",
           label="kappa = 1 (unit accumulator)") +
  scale_colour_manual(values=arch_pal, name=NULL) +
  scale_fill_manual(values=arch_pal, name=NULL) +
  coord_cartesian(xlim=c(0,4)) +
  labs(x="per-word kappa: sigmoid slope on log(experience)", y="density") +
  theme_minimal(base_size=10) +
  theme(legend.position=c(0.78,0.72), legend.text=element_text(size=7.5),
        legend.background=element_rect(fill="white", colour=NA))
```

