```{r si-setup, include=FALSE}
suppressPackageStartupMessages({
  library(here); library(dplyr); library(tidyr); library(ggplot2)
  library(patchwork); library(latex2exp); library(knitr); library(kableExtra)
  library(diptest)
})
source(here("paper", "_helpers.R"))
blups     <- readRDS(here("paper", "cache", "blups_demographics.rds"))
si_loo    <- readRDS(here("paper", "cache", "si_loo.rds"))     # LOO ladder (M0-M3)
si_loglin <- readRDS(here("paper", "cache", "si_loglin.rds"))  # log- vs linear-age (M3)
fig2      <- readRDS(here("paper", "cache", "fig2_efficiency.rds"))  # word efficiency + delta validation
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", out.width = "100%", dpi = 300
)
```

# Generative Model

*[Pending prose for the generative accumulator model. The figure below shows the latent-ability ($\theta$) buildup across the model rungs M0-M3.]*

```{r fig-schematic-theta, fig.width=8, fig.height=2.6}
#| label: fig-schematic-theta
#| fig-cap: ""
# [MCF: caption on next pass.] Latent-ability (theta) version of the
# main-text model buildup (M0->M3). The main text shows only the vocab-
# space projection of these curves; here is theta_i(t) itself, before it
# is pushed through the word-difficulty distribution. Illustrative
# parameters, not estimated.
ages <- seq(8, 30, length.out = 80)
A0 <- 8; L <- log(ages / A0)
qs <- c(0.1, 0.25, 0.5, 0.75, 0.9); z <- qnorm(qs)
XI0 <- -2.0; KAPPA_A <- 2.5
variants <- tibble::tribble(
  ~name,                       ~kappa_pop, ~sigma_alpha, ~sigma_kappa, ~eq,
  "(M0) accumulator",          1,           0,            0,            r"($\theta = \log(r) + \log(t)$)",
  "(M1) + acceleration",       KAPPA_A,     0,            0,            r"($\theta = \log(r) + \kappa\,\log(t)$)",
  "(M2) + efficiency var.",    KAPPA_A,     1.1,          0,            r"($\theta = \xi_i + \kappa\,\log(t)$)",
  "(M3) + per-child accel.",   KAPPA_A,     1.1,          0.7,          r"($\theta = \xi_i + \kappa_i\,\log(t)$)"
)
sim <- variants |>
  mutate(name = factor(name, levels = name)) |>
  rowwise() |>
  do({ v <- .
    bind_rows(lapply(seq_along(qs), function(i) {
      tibble(name = v$name,
             qprob_f = factor(as.character(qs[i]),
                              levels = c("0.1","0.25","0.5","0.75","0.9")),
             age = ages,
             theta = XI0 + z[i]*v$sigma_alpha + (v$kappa_pop + z[i]*v$sigma_kappa)*L,
             eq = v$eq)
    })) }) |>
  ungroup()
eq_lab <- variants |> mutate(name = factor(name, levels = name)) |> select(name, eq)
lwidth <- function(qf) ifelse(as.character(qf) == "0.5", 0.95, 0.55)
ggplot(sim, aes(age, theta, color = qprob_f, group = qprob_f)) +
  geom_line(aes(linewidth = lwidth(qprob_f))) +
  geom_label(data = eq_lab, aes(x = -Inf, y = Inf,
             label = latex2exp::TeX(eq, output = "character")),
             inherit.aes = FALSE, parse = TRUE, hjust = -0.05, vjust = 1.4,
             size = 2.4, label.size = 0, fill = scales::alpha("white", 0.8)) +
  facet_wrap(~ name, nrow = 1) +
  scale_color_manual(values = WORDBANK_PALETTE, name = "Percentile",
                     labels = c("10th","25th","50th","75th","90th"), drop = FALSE) +
  scale_linewidth_identity() +
  coord_cartesian(ylim = c(-3.5, 4.5)) +
  labs(x = "age (months)", y = expression(theta[i](t))) +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(face = "bold", size = 8),
        legend.position = "bottom", panel.grid.minor = element_blank())
```

# Comparison with Prior Models

```md
## Relation to prior accumulator models  [draft for promotion — MCF to review]

Our model is one member of a family of accumulator models of early
vocabulary, and the family is easiest to compare in a single set of
coordinates. Writing the latent ability of child $i$ on word $j$ at
age $t$ as

$$\eta_{ij}(t) = \log r_i + \log \alpha_i + \kappa_i \log t - \delta_j,$$

every prior model in this tradition is a special case that fixes one or
more of three quantities: the between-child variance in efficiency
($\sigma_\alpha$), the population scaling exponent ($\kappa_{\text{pop}}$),
and — critically — the between-child variance in that exponent
($\sigma_\kappa$). Mapping each prior model into these coordinates makes
the comparison structural rather than notational.

**McMurray [-@mcmurray2007]** is a pure accumulator: every word fills at a
single unit rate ($\kappa = 1$) and children do not differ
($\sigma_\alpha = \sigma_\kappa = 0$). Its central insight is that the
*vocabulary* curve accelerates even when the underlying rate does not,
because ability sweeps through a distribution of word difficulties — the
construction reproduced in panel (A) of our schematic. Acceleration here
is a population-level property of the difficulty distribution, not a
change in the child. This corresponds to our M1.

**Mitchell & McMurray [-@mitchell2009]** add a leverage term: already-known
words speed the acquisition of new ones. They prove that this feedback
shifts *when* words are acquired but cannot manufacture acceleration that
is not already present in the difficulty distribution. In our coordinates
it moves the timing of acquisition without changing $\kappa$, and — being
non-linear in accumulated vocabulary — it is the one prior model that does
not embed cleanly in the linear predictor above.

**Hidaka [-@hidaka2013]** introduces a rate-change process in which ability
grows as $t^{D+1}$; his exponent $D$ is exactly $\kappa_{\text{pop}} - 1$
in our notation, so $D > 0$ is population-level acceleration
($\kappa_{\text{pop}} > 1$). This is our M2. The acceleration is shared
across children ($\sigma_\kappa = 0$). Hidaka relates $D$ to word
imageability — i.e. a *per-word* scaling exponent — which is an
interesting direction we do not pursue here.

**Mollica & Piantadosi [-@mollica2017]** model each word as a Gamma
waiting-time process in which acquisition occurs after roughly ten
"effective learning instances." In our coordinates this is again a pure
accumulator ($\kappa = 1$); the per-word variation they capture lives in
the Gamma shape parameter, which corresponds to a per-word discrimination
(a 2PL extension) rather than to per-child structure.

**Kachergis et al. [-@kachergis2022]** are the first to free
$\sigma_\alpha$: they cast vocabulary as a Rasch model in which each child
has their own latent ability, unifying the accumulator and item-response
traditions. This is our M3. But the scaling exponent remains shared across
children ($\sigma_\kappa = 0$), and the model is proposed as a framework
rather than fit to longitudinal individual data.

Reading down this list, the literature progressively frees parameters —
McMurray fixes all three, Hidaka frees $\kappa_{\text{pop}}$, Kachergis
frees $\sigma_\alpha$ — but every model holds $\sigma_\kappa = 0$.

| Model | $\sigma_\alpha$ | $\kappa_{\text{pop}}$ | $\sigma_\kappa$ | per-word $\delta_j$ |
|---|:--:|:--:|:--:|:--:|
| McMurray (2007)            | 0 | 1     | 0 | free |
| Mitchell & McMurray (2009) | 0 | 1     | 0 | free + leverage |
| Hidaka (2013), rate-change | 0 | $D+1$ | 0 | free |
| Mollica & Piantadosi (2017)| 0 | 1     | 0 | free |
| Kachergis et al. (2022)    | free | free | 0 | free |
| **This work (M4)**         | **free** | **free** | **free** | **free** |

Our contribution is to free $\sigma_\kappa$: children differ not only in
how efficient they are but in how steeply their learning accelerates. This
is the load-bearing addition — it is what lets the model fit the widening
fan of trajectories with age (M4 vs M3 in the main text). It is also a
falsifiable claim rather than a parameter convenience: if $\sigma_\kappa$
were near zero the model would collapse onto Kachergis et al., but the
estimated per-child slope variance is well above zero and large relative
to the efficiency variance $\sigma_\alpha$ (main text, Figure 2 and
Results).
```

# Non-identifiability in Cross-Sectional Data

```{r fig-cross-sectional-confound, fig.width=9, fig.height=4}
#| label: fig-cross-sectional-confound
#| fig-cap: "Non-identifiability of growth models in cross-sectional data. Panels show the same simulated data, reflecting the possibility of stable between-child intercept differences (A) vs. varying growth slopes (B)."
# Why longitudinal data are needed to identify per-child acceleration.
#
# A single cross-sectional snapshot — one ability estimate per child at
# their observed age — is consistent with qualitatively different
# individual-trajectory structures. We draw ONE set of cross-sectional
# points (identical in both panels) and overlay two generative stories
# that each pass through every point exactly:
#   (A) stable between-child differences: every child shares the
#       population slope and differs only in intercept (parallel lines);
#   (B) diverging trajectories: every child starts from a common early
#       ability and differs in slope (a fan).
# With one observation per child the two are indistinguishable — only
# repeated measurement of the same child reveals which holds.

set.seed(42)
N         <- 28
a0        <- 6                       # reference age; log(t/a0) > 0 over 8-30
kappa_pop <- 2.5                     # population acceleration (slope on log-age)
sigma_xi  <- 1.0                     # between-child intercept spread

ages_obs <- runif(N, 8, 30)
L_obs    <- log(ages_obs / a0)
# cross-sectional cloud: ability rises with log-age + between-child noise
theta_obs <- kappa_pop * L_obs + rnorm(N, 0, sigma_xi)
pts <- data.frame(age = ages_obs, theta = theta_obs)

age_grid <- seq(8, 30, length.out = 60)
L_grid   <- log(age_grid / a0)

# (A) parallel: each child same slope kappa_pop, intercept set to hit point
parallel <- bind_rows(lapply(seq_len(N), function(i) {
  xi <- theta_obs[i] - kappa_pop * L_obs[i]
  data.frame(child = i, age = age_grid, theta = xi + kappa_pop * L_grid)
}))

# (B) fanning: common early origin theta0, slope set to hit point
theta0 <- min(theta_obs) - 1.5       # origin below the cloud so all slopes > 0
fanning <- bind_rows(lapply(seq_len(N), function(i) {
  kappa_i <- (theta_obs[i] - theta0) / L_obs[i]
  data.frame(child = i, age = age_grid, theta = theta0 + kappa_i * L_grid)
}))

yr <- range(theta_obs) + c(-1, 1)
mk <- function(traj, ttl) {
  ggplot() +
    geom_line(data = traj, aes(age, theta, group = child),
              color = "#2b8cbe", alpha = 0.35, linewidth = 0.3) +
    geom_line(data = data.frame(age = age_grid,
                                theta = kappa_pop * L_grid +
                                        (mean(theta_obs) - kappa_pop * mean(L_obs))),
              aes(age, theta), linetype = "dashed",
              color = "grey45", linewidth = 0.4) +
    geom_point(data = pts, aes(age, theta), size = 1.3, color = "grey15") +
    coord_cartesian(ylim = yr) +
    labs(x = "age (months)", y = expression(ability~theta), title = ttl) +
    theme_minimal(base_size = 10) +
    theme(plot.title = element_text(size = 10, face = "bold"),
          panel.grid.minor = element_blank())
}

mk(parallel, "(A) stable between-child differences (intercepts vary)") +
mk(fanning,  "(B) diverging trajectories (slopes vary)")
```

# Datasets

```{r tbl-datasets}
#| label: tbl-datasets
#| tbl-cap: "Characteristics of all datasets used in the paper. N and ages reflect the analysis samples (post-exclusion). Cross-sectional rows pool all contributed Wordbank datasets for a language. For AM2018 and FMW2013, which enter both the input and processing analyses with slightly different inclusion criteria, N is the union of the two subsamples."
# Reads paper/cache/table1_datasets.csv, built by paper/build_table1.R from
# the analysis bundles (glmer_ladder bundles; io_pooled + proc_dp bundles;
# cross_sectional_demographics frames). Re-run that script when subsamples
# change; the CSV is committed so rendering needs no fit bundles.
#
# The Citation column is PLAIN TEXT because kable does not render [@key]
# markers; the matching bib keys are listed in `nocite:` in the YAML header
# so the references still appear in the bibliography.
# [MCF: (a) confirm Thal/Smith/Marchman citations + years (@smith/@marchman
#  keys still missing from the bib). (b) cross-sectional rows currently cite
#  Wordbank (@frank2017) -- swap in per-language CDI-adaptation citations if
#  you prefer. (c) caption wording is a first pass. (d) the intro sentence
#  above still says "four longitudinal datasets" -- needs updating for the
#  by-study table.]
table1 <- read.csv(file.path(CACHE, "table1_datasets.csv")) |>
  mutate(across(c(input, longitudinal), ~ ifelse(is.na(.x), "", .x)),
         n = format(n, big.mark = ","),
         n_admins = format(n_admins, big.mark = ","),
         mean_age = sprintf("%.1f", mean_age),
         across(c(n_recordings, n_lwl),
                ~ ifelse(is.na(.x), "", format(.x, big.mark = ","))))

grp_n <- table(factor(table1$group, levels = unique(table1$group)))
table1 |>
  select(-group) |>
  knitr::kable(
    col.names = c("Citation", "Language", "N", "Admins",
                  "Mean", "Min", "Max", "Long.", "Input", "Recordings", "LWL sessions"),
    align = "llrrrrrclrr") |>
  kableExtra::add_header_above(c(" " = 4, "Age (months)" = 3, " " = 4)) |>
  kableExtra::pack_rows(index = grp_n) |>
  # Quarto forces longtable in PDF (scale_down is a no-op there), so the
  # oversize table goes on a landscape page until it moves to the supplement.
  kableExtra::kable_styling(font_size = 8, latex_options = "repeat_header") |>
  kableExtra::landscape()
```

# Full Model Comparison Results

```{r tbl-loo}
#| label: tbl-loo
#| tbl-cap: "Bayesian leave-one-out (LOO) model comparison. Each cell is the difference in expected log predictive density (ELPD) relative to the best model (M3), with the standard error of the difference in parentheses; more negative is worse. Every step of the ladder improves fit in every dataset, and the full model M3 (per-child efficiency *and* acceleration) is preferred throughout. Built from the per-fit LOO objects (paper/build_cache_short.R)."
short <- c("English (Thal)" = "Thal", "English (Smith)" = "Smith",
           "English (Marchman)" = "Marchman", "Norwegian" = "Norwegian",
           "Japanese" = "Japanese")
mod_desc <- c(M0 = "M0. accumulator ($\\kappa=1$)",
              M1 = "M1. + acceleration",
              M2 = "M2. + efficiency var.",
              M3 = "M3. + acceleration var.")
lang_order <- levels(si_loo$lang)
cell <- function(m, e, s) ifelse(m == "M3", "0",
  sprintf("%s (%s)", format(round(e), big.mark = ","), format(round(s))))
tab <- si_loo |>
  mutate(txt = cell(as.character(model), elpd_diff, se_diff)) |>
  select(model, lang, txt) |>
  tidyr::pivot_wider(names_from = lang, values_from = txt) |>
  arrange(model) |>
  mutate(Model = mod_desc[as.character(model)]) |>
  select(Model, all_of(lang_order))
knitr::kable(tab, align = c("l", rep("r", length(lang_order))), escape = FALSE,
             col.names = c("Model", unname(short[lang_order]))) |>
  kableExtra::add_header_above(
    c(" " = 1, "$\\\\Delta$ELPD vs M3 (SE)" = length(lang_order)), escape = FALSE) |>
  kableExtra::kable_styling(latex_options = "scale_down", font_size = 8)
```

# Linear vs. Logistic Age

```{r tbl-loglin}
#| label: tbl-loglin
#| tbl-cap: "Log-age vs linear-age accumulation. LOO ELPD advantage of the log-age model (M3) over its linear-age counterpart, with SE; positive favors log-age. Log-age is preferred in every dataset tested. (Norwegian's linear-age fit is pending.)"
short <- c("English (Thal)" = "Thal", "English (Smith)" = "Smith",
           "English (Marchman)" = "Marchman", "Norwegian" = "Norwegian",
           "Japanese" = "Japanese")
tab <- si_loglin |>
  mutate(Dataset = unname(short[as.character(lang)]),
         adv = sprintf("%s (%s)", format(round(-elpd_diff), big.mark = ","),
                       format(round(se_diff)))) |>
  select(Dataset, adv)
knitr::kable(tab, align = c("l", "r"), escape = FALSE,
             col.names = c("Dataset", "$\\Delta$ELPD (log $-$ linear)")) |>
  kableExtra::kable_styling(font_size = 9, latex_options = "hold_position")
```

# Characterizing Variation in Efficiency and Acceleration

```{r blup-prep, include=FALSE}
# Per-child BLUP distributions (efficiency xi, acceleration zeta) from the
# D_log (M4) fits, used for the dip-test table and histogram below. We
# z-score WITHIN dataset (datasets differ in scale) and drop a handful of
# extreme BLUPs (|z| > 4) that come from sparsely-observed children rather
# than distinct subgroups.
study_lab <- c(thal = "English (Thal)", smith = "English (Smith)",
               marchman = "English (Marchman)", norwegian = "Norwegian",
               japanese = "Japanese")
bz <- blups |>
  group_by(lang_slug) |>
  mutate(xz = as.numeric(scale(xi)), zz = as.numeric(scale(zeta))) |>
  ungroup() |>
  filter(is.finite(xz), is.finite(zz), abs(xz) <= 4, abs(zz) <= 4) |>
  mutate(study = factor(study_lab[lang_slug], levels = unname(study_lab)))
```

```{r tbl-blup-dip}
#| label: tbl-blup-dip
#| tbl-cap: ""
# Hartigan's dip test of unimodality on the per-child BLUPs, by dataset and
# pooled. Large p is consistent with a single mode — i.e. continuous
# between-child variation rather than discrete latent classes. Computed on
# the within-dataset z-scored values from blup-prep.
dip_row <- function(d, label) {
  tibble::tibble(Dataset = label, N = nrow(d),
                 eff = diptest::dip.test(d$xz)$p.value,
                 acc = diptest::dip.test(d$zz)$p.value)
}
diptab <- bind_rows(
  lapply(names(study_lab), function(s)
    dip_row(filter(bz, lang_slug == s), study_lab[[s]])),
  dip_row(bz, "Pooled"))
pfmt <- function(p) ifelse(p >= 0.995, ">0.99", sprintf("%.2f", p))
diptab |>
  mutate(eff = pfmt(eff), acc = pfmt(acc)) |>
  knitr::kable(align = c("l", "r", "r", "r"), escape = FALSE,
               col.names = c("Dataset", "N",
                             "Efficiency ($\\xi$)", "Acceleration ($\\zeta$)")) |>
  kableExtra::kable_styling(font_size = 9, latex_options = "hold_position")
```

```{r fig-blup-hist, fig.width=7, fig.height=6.5}
#| label: fig-blup-hist
#| fig-cap: ""
# Histograms of the per-child BLUPs (within-dataset z-scored): efficiency
# (xi) and acceleration (zeta), one row per dataset. Dashed curve is the
# standard-normal reference. Distributions are single-moded (cf. dip-test
# table), consistent with continuous between-child variation.
hist_df <- bz |>
  select(study, xz, zz) |>
  pivot_longer(c(xz, zz), names_to = "param", values_to = "z") |>
  mutate(param = factor(param, levels = c("xz", "zz"),
                        labels = c("Efficiency~(xi)", "Acceleration~(zeta)")))
ggplot(hist_df, aes(z)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = "#2b8cbe", colour = "white", linewidth = 0.1) +
  stat_function(fun = dnorm, colour = "grey30", linetype = "dashed",
                linewidth = 0.4) +
  facet_grid(study ~ param, scales = "free_y",
             labeller = labeller(param = label_parsed)) +
  labs(x = "per-child BLUP (z-scored within dataset)", y = "density") +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(size = 8, face = "bold"),
        panel.grid.minor = element_blank())
```

# Changes in Word Efficiency

```md
## Word-level efficiency: from difficulty to exposures  [draft for promotion — MCF to review]

The main-text efficiency figure translates the fitted model into a concrete
developmental claim: the number of exposures a word requires before it is learned
falls by orders of magnitude across early childhood. Two quantities underlie each
point.

**Age of acquisition.** For a word with fitted difficulty $\delta_j$, the age at
which a typical child (population intercept $\mu_\xi$, scaling $\kappa$) reaches
50% production probability is
$$t_{50}(j) = a_0 \exp\!\big((\delta_j - \log H - \mu_\xi)/\kappa\big),$$
read directly off the accumulator's linear predictor. Word difficulty here is the
*model's* $\delta_j$, not corpus frequency: high-frequency function words such as
"in" and "if" are acquired late, so frequency is a poor proxy for difficulty.

**Cumulative exposures.** The number of times a word has been heard by $t_{50}$ is
the input rate times the time-to-learn times the word's share of tokens,
$N_j = R \cdot t_{50}(j) \cdot p_j$, with $p_j$ the word's CHILDES token frequency
and $R$ the typical input rate. The descriptive model does not itself estimate an
input rate, so we anchor $R$ to the population input-rate estimates (see
Population Input Range Estimation): the figure uses the central value and the text
reports the range across sources.

**Validation.** Because $\delta_j$ drives the age axis, it must track how hard
words actually are for children. Fitted difficulty is strongly *negatively*
correlated with each word's empirical production rate in every dataset (easy words
are produced by more children; @fig-delta-validation) — the frequency-based check
that this replaces gives near-zero correlations and false-alarms.
```

```{r fig-delta-validation, fig.width=8, fig.height=2.4}
#| label: fig-delta-validation
#| fig-cap: ""
# Model item difficulty (delta_j) vs empirical per-item production rate, per
# dataset. The strong negative relationship confirms delta_j tracks how hard
# words are for children (validated against production, not frequency).
.rlab <- fig2$meta |> mutate(txt = sprintf("r = %.2f", cor_dj_prod),
                             lang = factor(lang, levels = levels(fig2$items$lang)))
ggplot(fig2$items, aes(emp_prod, delta_j)) +
  geom_point(alpha = 0.30, size = 0.6, colour = "#2b8cbe") +
  geom_smooth(method = "lm", se = FALSE, colour = "grey20", linewidth = 0.6, formula = y ~ x) +
  geom_text(data = .rlab, aes(x = Inf, y = Inf, label = txt), inherit.aes = FALSE,
            hjust = 1.1, vjust = 1.5, size = 2.6, colour = "grey25") +
  facet_wrap(~ lang, nrow = 1) +
  labs(x = "empirical production rate", y = expression("fitted difficulty  " * delta[j])) +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(face = "bold", size = 8),
        panel.grid.minor = element_blank())
```

# Computing Acceleration in LMs

```md
## Comparing children and language models on one slope axis  [draft for promotion — MCF to review]

The comparison to language models in Figure 6 places children and models
on a single, literally shared axis: the slope of the acquisition sigmoid
in log-experience. Both populations have sigmoidal per-word acquisition
curves, and the slope of that sigmoid is directly comparable.

On the child side, our linear predictor makes this slope exact.
Differentiating $\eta_{ij}(t)$ with respect to $\log t$ gives $\kappa_i$:
the logit-probability that a child produces a word rises linearly in
log-age with slope equal to that child's scaling exponent. The per-child
sigmoid slope *is* $\kappa_i$, by construction.

On the model side, @chang2022 fit a four-parameter logistic to each
(model, word) pair, describing how the model's surprisal for a word falls
over training. Re-expressed as the fraction of a word's surprisal
reduction achieved so far, their fit is a sigmoid in
$\log_{10}(\text{training steps})$ with slope $1/\text{scale}_w$.
Dividing by $\ln 10$ converts this to natural-log units — the same units
as $\kappa_i$ — so the two slopes can be read off the same axis.

The result is a striking separation (Figure 6). Language models cluster
near a slope of one — the unit-accumulator boundary, where each log-unit
of additional experience yields a fixed increment of learning. Children
sit in a broad distribution roughly an order of magnitude higher. On the
shared axis, children's word learning accelerates with experience in a way
that these models' does not.

Three caveats bound this comparison. First, the variability is asymmetric:
children's slopes vary across *children*, whereas a single model is a
deterministic learner, so model slopes vary across *words* within a run
rather than across replicate learners. Second, the steep child slope is
the *per-child* quantity; a researcher fitting a single sigmoid to
cross-sectional Wordbank data per word would recover a shallower slope,
because marginalizing over the population attenuates it. Third, the models
in @chang2022 were trained on adult written text rather than
child-directed speech, so part of the gap could reflect the input
distribution rather than the learner. Models trained directly on CHILDES
(Feng et al., forthcoming) provide the matched-input test: if the gap
largely persists, it is structural; if it shrinks substantially, the input
distribution is doing more of the work than we currently attribute to it.

<!-- [BIB NEEDED] Feng et al. (CHILDES-trained GPT) — no verified entry yet;
     named in text only. Daunizeau (2017) logistic-Gaussian approximation is
     used in the derivation note but not needed for this prose. -->
```

```md
## Connection to neural scaling laws  [draft for promotion — MCF to review; most speculative of the three]

The per-word picture above connects to a more familiar aggregate
phenomenon: neural scaling laws. A model's loss falls as a smooth power
law in the amount of training data, and this descent is remarkably
regular — different runs, and different model sizes, trace nearly the same
curve (Kaplan et al., 2020; Hoffmann et al., 2022). Our per-word analysis
is the micro-level counterpart of that regularity. Each word is acquired
along a sigmoid whose slope sits near the unit-accumulator value, and the
spread of those slopes — across words and across runs — is small. The
aggregate smoothness of scaling laws and the per-word $\kappa \approx 1$
we recover are two views of the same property: language models convert
experience into knowledge as near-pure accumulators, with little
instance-to-instance variation.

Children depart from this picture on both axes at once. Their per-instance
scaling is far steeper ($\kappa \approx 10$), and they vary substantially
from one another in that exponent ($\sigma_\kappa$ well above zero). The
"data gap" between children and language models [@frank2023; @warstadt2023]
is therefore not only that children reach adult competence from orders of
magnitude less input. It is that children convert experience into knowledge
through a different scaling relationship — steeper, and far more variable
across individuals — than the one that scaling laws describe for models.

This connection is analogical rather than formal: scaling laws characterize
aggregate loss, whereas our $\kappa$ characterizes per-item acquisition,
and the two need not move together. We offer it as a framing for why the
data gap may be harder to close than raw data counts suggest.

<!-- [BIB NEEDED] Kaplan et al. (2020) and Hoffmann et al. (2022) are named
     in text but have NO verified bib entries yet — do not promote with live
     @ markers until added. -->
```

# Generating the Model-Ladder Quantile Fans

*[STALE: this describes the retired glmer BLUP-bootstrap fan. The short paper's Fig 1 fan is instead drawn from the Bayesian M3 scalar posteriors (a synthetic MVN population; studies/bayes_long/03_fan.R). Rewrite to match, or remove.]*

```md
## How the model-ladder quantile bands are generated  [draft for promotion — MCF to review]

The percentile bands ("fans") overlaid on the empirical trajectories in the
model-ladder figures are not closed-form quantiles of a fitted distribution. For
each (dataset, model) fit we generate them by Monte-Carlo simulation: we build a
synthetic population of children and read off vocabulary percentiles at each age.

**Bootstrapping children from the fitted random effects.** For the models with a
per-child random effect (the C and D variants in both age bases) we draw
$N = 500$ synthetic children by resampling *with replacement* from the fit's
per-child BLUPs — the shrunken empirical-Bayes estimates of each observed child's
random intercept and, for the D models, random slope. The resampling uses a fixed
seed, so the bands are reproducible. The models with no per-child random effect
(the Rasch-style model A and the population-slope B models) have no child
distribution to resample and so collapse to a single median curve rather than a
fan.

We bootstrap the realised BLUPs rather than sampling parametrically from
$\mathcal{N}(0, \hat\Sigma)$. The two agree in expectation, but parametric
sampling assumes the *unshrunken* population covariance and routinely generates
extreme intercept–slope pairs that no real child exhibited; at ages where the
data are thin this inflates the upper percentiles. Resampling the BLUPs keeps the
simulated population conditional on the children we actually observed.

**Each child's expected vocabulary curve.** Ages run on a per-dataset integer
grid spanning the 5th–95th percentile of that dataset's observed ages, so the
bands are never extrapolated beyond the data window. For a bootstrapped child $i$
with random intercept $u^{0}_i$ and slope $u^{1}_i$, the linear predictor for item
$j$ at age $t$ is
$$\eta_{ij}(t) = \beta_0 + (\beta_{\text{age}} + u^{1}_i)\, g(t) + u^{0}_i + b_j,$$
where $\beta_0$ and $\beta_{\text{age}}$ are the fixed effects, $b_j$ is item $j$'s
fitted random effect, and $g(t) = \log(t / a_0)$ for the log-age models (and for
model A, whose offset pins $\beta_{\text{age}} \equiv 1$) or $g(t) = t - a_0$ for
the linear-age models. The child's expected productive vocabulary is the sum of
per-item production probabilities over the dataset's $J$ items,
$$V_i(t) = \sum_{j=1}^{J} \operatorname{logit}^{-1}\!\big(\eta_{ij}(t)\big),$$
clamped to $[0, J]$. Item random effects are held at their fitted values — only
children are resampled — and $V_i(t)$ is an *expected* count, so each simulated
child's curve is smooth and the width of the fan reflects between-child variation
alone.

**Quantiles.** At each (dataset, model, age) the plotted bands are the 10th,
25th, 50th, 75th, and 90th percentiles of $\{V_i(t)\}_{i=1}^{500}$. The simulation
is implemented in `studies/glmer_ladder/04a_simulate.R`.
```
