```{r si-setup, include=FALSE}
suppressPackageStartupMessages({
  library(here); library(dplyr); library(tidyr); library(ggplot2)
  library(patchwork); library(latex2exp); library(knitr); library(kableExtra)
  library(diptest)
})
source(here("paper", "_helpers.R"))
si_blups  <- readRDS(here("paper", "cache", "si_blups.rds"))   # per-child BLUPs (M3): xi, kappa
si_loo    <- readRDS(here("paper", "cache", "si_loo.rds"))     # LOO ladder (M0-M3)
si_loglin <- readRDS(here("paper", "cache", "si_loglin.rds"))  # log- vs linear-age (M3)
fig2      <- readRDS(here("paper", "cache", "fig2_efficiency.rds"))  # word efficiency + delta validation
si_ds     <- readRDS(here("paper", "cache", "si_datasets.rds"))      # datasets table (5 M3 bundles)
si_gamlss <- readRDS(here("paper", "cache", "si_gamlss.rds"))        # parametric vs GAMLSS overlay + per-child kappa
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", out.width = "100%", dpi = 300
)
```

# Generative Model

*[Pending prose for the generative accumulator model. The figure below shows the latent-ability ($\theta$) buildup across the model rungs M0-M3.]*

```{r fig-schematic-theta, fig.width=8, fig.height=2.6}
#| label: fig-schematic-theta
#| fig-cap: ""
# [MCF: caption on next pass.] Latent-ability (theta) version of the
# main-text model buildup (M0->M3). The main text shows only the vocab-
# space projection of these curves; here is theta_i(t) itself, before it
# is pushed through the word-difficulty distribution. Illustrative
# parameters, not estimated.
ages <- seq(8, 30, length.out = 80)
A0 <- 8; L <- log(ages / A0)
qs <- c(0.1, 0.25, 0.5, 0.75, 0.9); z <- qnorm(qs)
XI0 <- -2.0; KAPPA_A <- 2.5
variants <- tibble::tribble(
  ~name,                       ~kappa_pop, ~sigma_alpha, ~sigma_kappa, ~eq,
  "(M0) accumulator",          1,           0,            0,            r"($\theta = \log(r) + \log(t)$)",
  "(M1) + acceleration",       KAPPA_A,     0,            0,            r"($\theta = \log(r) + \kappa\,\log(t)$)",
  "(M2) + efficiency var.",    KAPPA_A,     1.1,          0,            r"($\theta = \xi_i + \kappa\,\log(t)$)",
  "(M3) + per-child accel.",   KAPPA_A,     1.1,          0.7,          r"($\theta = \xi_i + \kappa_i\,\log(t)$)"
)
sim <- variants |>
  mutate(name = factor(name, levels = name)) |>
  rowwise() |>
  do({ v <- .
    bind_rows(lapply(seq_along(qs), function(i) {
      tibble(name = v$name,
             qprob_f = factor(as.character(qs[i]),
                              levels = c("0.1","0.25","0.5","0.75","0.9")),
             age = ages,
             theta = XI0 + z[i]*v$sigma_alpha + (v$kappa_pop + z[i]*v$sigma_kappa)*L,
             eq = v$eq)
    })) }) |>
  ungroup()
eq_lab <- variants |> mutate(name = factor(name, levels = name)) |> select(name, eq)
lwidth <- function(qf) ifelse(as.character(qf) == "0.5", 0.95, 0.55)
ggplot(sim, aes(age, theta, color = qprob_f, group = qprob_f)) +
  geom_line(aes(linewidth = lwidth(qprob_f))) +
  geom_label(data = eq_lab, aes(x = -Inf, y = Inf,
             label = latex2exp::TeX(eq, output = "character")),
             inherit.aes = FALSE, parse = TRUE, hjust = -0.05, vjust = 1.4,
             size = 2.4, label.size = 0, fill = scales::alpha("white", 0.8)) +
  facet_wrap(~ name, nrow = 1) +
  scale_color_manual(values = WORDBANK_PALETTE, name = "Percentile",
                     labels = c("10th","25th","50th","75th","90th"), drop = FALSE) +
  scale_linewidth_identity() +
  coord_cartesian(ylim = c(-3.5, 4.5)) +
  labs(x = "age (months)", y = expression(theta[i](t))) +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(face = "bold", size = 8),
        legend.position = "bottom", panel.grid.minor = element_blank())
```

# Comparison with Prior Models

```md
## Relation to prior accumulator models  [draft for promotion — MCF to review]

Our model is one member of a family of accumulator models of early
vocabulary, and the family is easiest to compare in a single set of
coordinates. Writing the latent ability of child $i$ on word $j$ at
age $t$ as

$$\eta_{ij}(t) = \log r_i + \log \alpha_i + \kappa_i \log t - \delta_j,$$

every prior model in this tradition is a special case that fixes one or
more of three quantities: the between-child variance in efficiency
($\sigma_\alpha$), the population scaling exponent ($\kappa_{\text{pop}}$),
and — critically — the between-child variance in that exponent
($\sigma_\kappa$). Mapping each prior model into these coordinates makes
the comparison structural rather than notational.

**McMurray [-@mcmurray2007]** is a pure accumulator: every word fills at a
single unit rate ($\kappa = 1$) and children do not differ
($\sigma_\alpha = \sigma_\kappa = 0$). Its central insight is that the
*vocabulary* curve accelerates even when the underlying rate does not,
because ability sweeps through a distribution of word difficulties — the
construction reproduced in panel (A) of our schematic. Acceleration here
is a population-level property of the difficulty distribution, not a
change in the child. This corresponds to our M0.

**Mitchell & McMurray [-@mitchell2009]** add a leverage term: already-known
words speed the acquisition of new ones. They prove that this feedback
shifts *when* words are acquired but cannot manufacture acceleration that
is not already present in the difficulty distribution. In our coordinates
it moves the timing of acquisition without changing $\kappa$, and — being
non-linear in accumulated vocabulary — it is the one prior model that does
not embed cleanly in the linear predictor above.

**Hidaka [-@hidaka2013]** introduces a rate-change process in which ability
grows as $t^{D+1}$; his exponent $D$ is exactly $\kappa_{\text{pop}} - 1$
in our notation, so $D > 0$ is population-level acceleration
($\kappa_{\text{pop}} > 1$). This is our M1. The acceleration is shared
across children ($\sigma_\kappa = 0$). Hidaka relates $D$ to word
imageability — i.e. a *per-word* scaling exponent — which is an
interesting direction we do not pursue here.

**Mollica & Piantadosi [-@mollica2017]** model each word as a Gamma
waiting-time process in which acquisition occurs after roughly ten
"effective learning instances." In our coordinates this is again a pure
accumulator ($\kappa = 1$); the per-word variation they capture lives in
the Gamma shape parameter, which corresponds to a per-word discrimination
(a 2PL extension) rather than to per-child structure.

**Kachergis et al. [-@kachergis2022]** are the first to free
$\sigma_\alpha$: they cast vocabulary as a Rasch model in which each child
has their own latent ability, unifying the accumulator and item-response
traditions. This is our M2. But the scaling exponent remains shared across
children ($\sigma_\kappa = 0$), and the model is proposed as a framework
rather than fit to longitudinal individual data.

Reading down this list, the literature progressively frees parameters —
McMurray fixes all three, Hidaka frees $\kappa_{\text{pop}}$, Kachergis
frees $\sigma_\alpha$ — but every model holds $\sigma_\kappa = 0$.

| Model | $\sigma_\alpha$ | $\kappa_{\text{pop}}$ | $\sigma_\kappa$ | per-word $\delta_j$ |
|---|:--:|:--:|:--:|:--:|
| McMurray (2007)            | 0 | 1     | 0 | free |
| Mitchell & McMurray (2009) | 0 | 1     | 0 | free + leverage |
| Hidaka (2013), rate-change | 0 | $D+1$ | 0 | free |
| Mollica & Piantadosi (2017)| 0 | 1     | 0 | free |
| Kachergis et al. (2022)    | free | free | 0 | free |
| **This work (M3)**         | **free** | **free** | **free** | **free** |

Our contribution is to free $\sigma_\kappa$: children differ not only in
how efficient they are but in how steeply their learning accelerates. This
is the load-bearing addition — it is what lets the model fit the widening
fan of trajectories with age (M3 vs M2 in the main text). It is also a
falsifiable claim rather than a parameter convenience: if $\sigma_\kappa$
were near zero the model would collapse onto Kachergis et al., but the
estimated per-child slope variance is well above zero and large relative
to the efficiency variance $\sigma_\alpha$ (main text, Figure 2 and
Results).
```

# Non-identifiability in Cross-Sectional Data

```{r fig-cross-sectional-confound, fig.width=9, fig.height=4}
#| label: fig-cross-sectional-confound
#| fig-cap: "Non-identifiability of growth models in cross-sectional data. Panels show the same simulated data, reflecting the possibility of stable between-child intercept differences (A) vs. varying growth slopes (B)."
# Why longitudinal data are needed to identify per-child acceleration.
#
# A single cross-sectional snapshot — one ability estimate per child at
# their observed age — is consistent with qualitatively different
# individual-trajectory structures. We draw ONE set of cross-sectional
# points (identical in both panels) and overlay two generative stories
# that each pass through every point exactly:
#   (A) stable between-child differences: every child shares the
#       population slope and differs only in intercept (parallel lines);
#   (B) diverging trajectories: every child starts from a common early
#       ability and differs in slope (a fan).
# With one observation per child the two are indistinguishable — only
# repeated measurement of the same child reveals which holds.

set.seed(42)
N         <- 28
a0        <- 6                       # reference age; log(t/a0) > 0 over 8-30
kappa_pop <- 2.5                     # population acceleration (slope on log-age)
sigma_xi  <- 1.0                     # between-child intercept spread

ages_obs <- runif(N, 8, 30)
L_obs    <- log(ages_obs / a0)
# cross-sectional cloud: ability rises with log-age + between-child noise
theta_obs <- kappa_pop * L_obs + rnorm(N, 0, sigma_xi)
pts <- data.frame(age = ages_obs, theta = theta_obs)

age_grid <- seq(8, 30, length.out = 60)
L_grid   <- log(age_grid / a0)

# (A) parallel: each child same slope kappa_pop, intercept set to hit point
parallel <- bind_rows(lapply(seq_len(N), function(i) {
  xi <- theta_obs[i] - kappa_pop * L_obs[i]
  data.frame(child = i, age = age_grid, theta = xi + kappa_pop * L_grid)
}))

# (B) fanning: common early origin theta0, slope set to hit point
theta0 <- min(theta_obs) - 1.5       # origin below the cloud so all slopes > 0
fanning <- bind_rows(lapply(seq_len(N), function(i) {
  kappa_i <- (theta_obs[i] - theta0) / L_obs[i]
  data.frame(child = i, age = age_grid, theta = theta0 + kappa_i * L_grid)
}))

yr <- range(theta_obs) + c(-1, 1)
mk <- function(traj, ttl) {
  ggplot() +
    geom_line(data = traj, aes(age, theta, group = child),
              color = "#2b8cbe", alpha = 0.35, linewidth = 0.3) +
    geom_line(data = data.frame(age = age_grid,
                                theta = kappa_pop * L_grid +
                                        (mean(theta_obs) - kappa_pop * mean(L_obs))),
              aes(age, theta), linetype = "dashed",
              color = "grey45", linewidth = 0.4) +
    geom_point(data = pts, aes(age, theta), size = 1.3, color = "grey15") +
    coord_cartesian(ylim = yr) +
    labs(x = "age (months)", y = expression(ability~theta), title = ttl) +
    theme_minimal(base_size = 10) +
    theme(plot.title = element_text(size = 10, face = "bold"),
          panel.grid.minor = element_blank())
}

mk(parallel, "(A) stable between-child differences (intercepts vary)") +
mk(fanning,  "(B) diverging trajectories (slopes vary)")
```

# Datasets

```{r tbl-datasets}
#| label: tbl-datasets
#| tbl-cap: "Characteristics of all datasets used in the paper. N and ages reflect the analysis samples (post-exclusion). Cross-sectional rows pool all contributed Wordbank datasets for a language. For AM2018 and FMW2013, which enter both the input and processing analyses with slightly different inclusion criteria, N is the union of the two subsamples."
# Built by paper/build_cache_short.R (section 6, si_datasets.rds) from the five
# longitudinal M3 bundles -- the single source now that the io/proc/cross-sectional
# analyses are cut (replaces the retired build_table1.R / table1_datasets.csv).
# Citation column is plain text (kable can't render [@key]); bib keys live in the
# YAML `nocite:`.
# [MCF: (a) Thal/Smith/Marchman citation years are 20XX placeholders (@smith/
#  @marchman keys still missing from the bib). (b) the caption above still
#  describes the OLD multi-analysis table -- rewrite for the 5 longitudinal
#  datasets. (c) intro N_total says 3,135 but the a3 (3+-admin) analysis sample
#  is 1,834 -- reconcile.]
si_ds |>
  transmute(citation, language,
            n = format(n_kids, big.mark = ","),
            n_admins = format(n_admins, big.mark = ","),
            mean_age = sprintf("%.1f", mean_age),
            min_age = as.character(min_age), max_age = as.character(max_age),
            med_admins = as.character(med_admins)) |>
  knitr::kable(
    col.names = c("Citation", "Language", "N", "Admins",
                  "Mean", "Min", "Max", "Adm./kid"),
    align = "llrrrrrr") |>
  kableExtra::add_header_above(c(" " = 4, "Age (months)" = 3, " " = 1)) |>
  kableExtra::kable_styling(font_size = 9, latex_options = "hold_position")
```

# Full Model Comparison Results

```{r tbl-loo}
#| label: tbl-loo
#| tbl-cap: "Bayesian leave-one-out (LOO) model comparison. Each cell is the difference in expected log predictive density (ELPD) relative to the best model (M3), with the standard error of the difference in parentheses; more negative is worse. Every step of the ladder improves fit in every dataset, and the full model M3 (per-child efficiency *and* acceleration) is preferred throughout. Built from the per-fit LOO objects (paper/build_cache_short.R)."
short <- c("English (Thal)" = "Thal", "English (Smith)" = "Smith",
           "English (Marchman)" = "Marchman", "Norwegian" = "Norwegian",
           "Japanese" = "Japanese")
mod_desc <- c(M0 = "M0. accumulator ($\\kappa=1$)",
              M1 = "M1. + acceleration",
              M2 = "M2. + efficiency var.",
              M3 = "M3. + acceleration var.")
lang_order <- levels(si_loo$lang)
cell <- function(m, e, s) ifelse(m == "M3", "0",
  sprintf("%s (%s)", format(round(e), big.mark = ","), format(round(s))))
tab <- si_loo |>
  mutate(txt = cell(as.character(model), elpd_diff, se_diff)) |>
  select(model, lang, txt) |>
  tidyr::pivot_wider(names_from = lang, values_from = txt) |>
  arrange(model) |>
  mutate(Model = mod_desc[as.character(model)]) |>
  select(Model, all_of(lang_order))
knitr::kable(tab, align = c("l", rep("r", length(lang_order))), escape = FALSE,
             col.names = c("Model", unname(short[lang_order]))) |>
  kableExtra::add_header_above(
    c(" " = 1, "$\\\\Delta$ELPD vs M3 (SE)" = length(lang_order)), escape = FALSE) |>
  kableExtra::kable_styling(latex_options = "scale_down", font_size = 8)
```

# Linear vs. Logistic Age

```{r tbl-loglin}
#| label: tbl-loglin
#| tbl-cap: "Log-age vs linear-age accumulation. LOO ELPD advantage of the log-age model (M3) over its linear-age counterpart, with SE; positive favors log-age. Log-age is preferred in every dataset tested. (Norwegian's linear-age fit is pending.)"
short <- c("English (Thal)" = "Thal", "English (Smith)" = "Smith",
           "English (Marchman)" = "Marchman", "Norwegian" = "Norwegian",
           "Japanese" = "Japanese")
tab <- si_loglin |>
  mutate(Dataset = unname(short[as.character(lang)]),
         adv = sprintf("%s (%s)", format(round(-elpd_diff), big.mark = ","),
                       format(round(se_diff)))) |>
  select(Dataset, adv)
knitr::kable(tab, align = c("l", "r"), escape = FALSE,
             col.names = c("Dataset", "$\\Delta$ELPD (log $-$ linear)")) |>
  kableExtra::kable_styling(font_size = 9, latex_options = "hold_position")
```

# Characterizing Variation in Efficiency and Acceleration

```{r blup-prep, include=FALSE}
# Per-child BLUP distributions (efficiency xi, acceleration kappa) from the
# Bayesian M3 fits (per-child exports), used for the dip-test table and histogram
# below. We z-score WITHIN dataset (datasets differ in scale) and drop a handful
# of extreme BLUPs (|z| > 4) that come from sparsely-observed children rather
# than distinct subgroups.
bz <- si_blups |>
  group_by(lang) |>
  mutate(xz = as.numeric(scale(xi)), zz = as.numeric(scale(kappa))) |>
  ungroup() |>
  filter(is.finite(xz), is.finite(zz), abs(xz) <= 4, abs(zz) <= 4) |>
  rename(study = lang)
```

```{r tbl-blup-dip}
#| label: tbl-blup-dip
#| tbl-cap: ""
# Hartigan's dip test of unimodality on the per-child BLUPs, by dataset and
# pooled. Large p is consistent with a single mode — i.e. continuous
# between-child variation rather than discrete latent classes. Computed on
# the within-dataset z-scored values from blup-prep.
dip_row <- function(d, label) {
  tibble::tibble(Dataset = label, N = nrow(d),
                 eff = diptest::dip.test(d$xz)$p.value,
                 acc = diptest::dip.test(d$zz)$p.value)
}
diptab <- bind_rows(
  lapply(levels(bz$study), function(s)
    dip_row(filter(bz, study == s), s)),
  dip_row(bz, "Pooled"))
pfmt <- function(p) ifelse(p >= 0.995, ">0.99", sprintf("%.2f", p))
diptab |>
  mutate(eff = pfmt(eff), acc = pfmt(acc)) |>
  knitr::kable(align = c("l", "r", "r", "r"), escape = FALSE,
               col.names = c("Dataset", "N",
                             "Efficiency ($\\xi$)", "Acceleration ($\\kappa$)")) |>
  kableExtra::kable_styling(font_size = 9, latex_options = "hold_position")
```

```{r fig-blup-hist, fig.width=7, fig.height=6.5}
#| label: fig-blup-hist
#| fig-cap: ""
# Histograms of the per-child BLUPs (within-dataset z-scored): efficiency
# (xi) and acceleration (kappa), one row per dataset. Dashed curve is the
# standard-normal reference. Distributions are single-moded (cf. dip-test
# table), consistent with continuous between-child variation.
hist_df <- bz |>
  select(study, xz, zz) |>
  pivot_longer(c(xz, zz), names_to = "param", values_to = "z") |>
  mutate(param = factor(param, levels = c("xz", "zz"),
                        labels = c("Efficiency~(xi)", "Acceleration~(kappa)")))
ggplot(hist_df, aes(z)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = "#2b8cbe", colour = "white", linewidth = 0.1) +
  stat_function(fun = dnorm, colour = "grey30", linetype = "dashed",
                linewidth = 0.4) +
  facet_grid(study ~ param, scales = "free_y",
             labeller = labeller(param = label_parsed)) +
  labs(x = "per-child BLUP (z-scored within dataset)", y = "density") +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(size = 8, face = "bold"),
        panel.grid.minor = element_blank())
```

# Changes in Word Efficiency

```md
## Word-level efficiency: from difficulty to exposures  [draft for promotion — MCF to review]

The main-text efficiency figure translates the fitted model into a concrete
developmental claim: the number of exposures a word requires before it is learned
falls by orders of magnitude across early childhood. Two quantities underlie each
point.

**Age of acquisition.** For a word with fitted difficulty $\delta_j$, the age at
which a typical child (population intercept $\mu_\xi$, scaling $\kappa$) reaches
50% production probability is
$$t_{50}(j) = a_0 \exp\!\big((\delta_j - \log H - \mu_\xi)/\kappa\big),$$
read directly off the accumulator's linear predictor. Word difficulty here is the
*model's* $\delta_j$, not corpus frequency: high-frequency function words such as
"in" and "if" are acquired late, so frequency is a poor proxy for difficulty.

**Cumulative exposures.** The number of times a word has been heard by $t_{50}$ is
the input rate times the time-to-learn times the word's share of tokens,
$N_j = R \cdot t_{50}(j) \cdot p_j$, with $p_j$ the word's CHILDES token frequency
and $R$ the typical input rate. The descriptive model does not itself estimate an
input rate, so we anchor $R$ to the population input-rate estimates (see
Population Input Range Estimation): the figure uses the central value and the text
reports the range across sources.

**Validation.** Because $\delta_j$ drives the age axis, it must track how hard
words actually are for children. Fitted difficulty is strongly *negatively*
correlated with each word's empirical production rate in every dataset (easy words
are produced by more children; @fig-delta-validation) — the frequency-based check
that this replaces gives near-zero correlations and false-alarms.
```

```{r fig-delta-validation, fig.width=8, fig.height=2.4}
#| label: fig-delta-validation
#| fig-cap: ""
# Model item difficulty (delta_j) vs empirical per-item production rate, per
# dataset. The strong negative relationship confirms delta_j tracks how hard
# words are for children (validated against production, not frequency).
.rlab <- fig2$meta |> mutate(txt = sprintf("r = %.2f", cor_dj_prod),
                             lang = factor(lang, levels = levels(fig2$items$lang)))
ggplot(fig2$items, aes(emp_prod, delta_j)) +
  geom_point(alpha = 0.30, size = 0.6, colour = "#2b8cbe") +
  geom_smooth(method = "lm", se = FALSE, colour = "grey20", linewidth = 0.6, formula = y ~ x) +
  geom_text(data = .rlab, aes(x = Inf, y = Inf, label = txt), inherit.aes = FALSE,
            hjust = 1.1, vjust = 1.5, size = 2.6, colour = "grey25") +
  facet_wrap(~ lang, nrow = 1) +
  labs(x = "empirical production rate", y = expression("fitted difficulty  " * delta[j])) +
  theme_minimal(base_size = 9) +
  theme(strip.text = element_text(face = "bold", size = 8),
        panel.grid.minor = element_blank())
```

# Computing Acceleration in LMs

```md
## Comparing children and language models on one slope axis  [draft for promotion — MCF to review]

The comparison to language models in Figure 6 places children and models
on a single, literally shared axis: the slope of the acquisition sigmoid
in log-experience. Both populations have sigmoidal per-word acquisition
curves, and the slope of that sigmoid is directly comparable.

On the child side, our linear predictor makes this slope exact.
Differentiating $\eta_{ij}(t)$ with respect to $\log t$ gives $\kappa_i$:
the logit-probability that a child produces a word rises linearly in
log-age with slope equal to that child's scaling exponent. The per-child
sigmoid slope *is* $\kappa_i$, by construction.

On the model side, @chang2022 fit a four-parameter logistic to each
(model, word) pair, describing how the model's surprisal for a word falls
over training. Re-expressed as the fraction of a word's surprisal
reduction achieved so far, their fit is a sigmoid in
$\log_{10}(\text{training steps})$ with slope $1/\text{scale}_w$.
Dividing by $\ln 10$ converts this to natural-log units — the same units
as $\kappa_i$ — so the two slopes can be read off the same axis.

The result is a striking separation (Figure 6). Language models cluster
near a slope of one — the unit-accumulator boundary, where each log-unit
of additional experience yields a fixed increment of learning. Children
sit in a broad distribution roughly an order of magnitude higher. On the
shared axis, children's word learning accelerates with experience in a way
that these models' does not.

Three caveats bound this comparison. First, the variability is asymmetric:
children's slopes vary across *children*, whereas a single model is a
deterministic learner, so model slopes vary across *words* within a run
rather than across replicate learners. Second, the steep child slope is
the *per-child* quantity; a researcher fitting a single sigmoid to
cross-sectional Wordbank data per word would recover a shallower slope,
because marginalizing over the population attenuates it. Third, the models
in @chang2022 were trained on adult written text rather than
child-directed speech, so part of the gap could reflect the input
distribution rather than the learner. Models trained directly on CHILDES
(Feng et al., forthcoming) provide the matched-input test: if the gap
largely persists, it is structural; if it shrinks substantially, the input
distribution is doing more of the work than we currently attribute to it.

<!-- [BIB NEEDED] Feng et al. (CHILDES-trained GPT) — no verified entry yet;
     named in text only. Daunizeau (2017) logistic-Gaussian approximation is
     used in the derivation note but not needed for this prose. -->
```

```md
## Relation to scaling laws  [draft for promotion — MCF to review]

The main-text figure places children and language models on one axis by
re-expressing both as a *fractional return to experience*. Here we derive
that axis from the scaling law, which makes the contrast exact rather than
descriptive.

Take the model's aggregate competence to be its loss on the target
vocabulary: the mean surprisal $L(D)$ of the CDI words after training on
$D$ word tokens of input. On the ladder this follows the Chinchilla form
$$L(D) = E + B\,D^{-\beta},$$
with $E$ the irreducible entropy floor and $B, \beta$ fit across budgets
(here $E \approx 2.94$ nats, $B \approx 312$, $\beta \approx 0.32$,
$R^2 = 0.998$). The *excess loss* $L(D) - E = B\,D^{-\beta}$ is the reducible
part that training removes; it falls as a power law of slope $-\beta$
(panel B), with the per-seed points scattering tightly around it.

The model's *marginal return* to experience is the loss it removes per
$e$-fold of data,
$$R(D) = -\frac{\mathrm{d}L}{\mathrm{d}\ln D} = \beta\,\big(L(D) - E\big),$$
obtained by differentiating the scaling law. This is a genuine return, but
it is measured in nats of loss, not in the logit units of a child's
$\kappa$; we do *not* call it an acceleration, and its numerical value — and
where it would cross any fixed threshold — depends on the arbitrary choice
of nats versus bits. What *is* unit-free is its ratio to the excess loss:
the *fractional return*, the share of the remaining reducible loss removed
per $e$-fold,
$$\gamma_{\text{LM}}(D) = \frac{R(D)}{L(D) - E} = \beta.$$
For a power law this is constant — $\gamma_{\text{LM}} = \beta \approx 0.32$
at every budget. That constancy *is* the scaling law.

The same construction applies to a child. Take competence to be production
loss, $L_i(t) = \operatorname{mean}_j[-\log p_{ij}(t)]$, whose floor is zero
(a known word is produced, $p \to 1$). Its marginal return is
$R_i(t) = -\mathrm{d}L_i/\mathrm{d}\log t = \kappa_i\,\operatorname{mean}_j(1 - p_{ij})$,
so the fractional return is
$$\gamma_i(t) = \frac{R_i(t)}{L_i(t)}
 = \kappa_i\,\frac{\operatorname{mean}_j(1 - p_{ij})}{\operatorname{mean}_j(-\log p_{ij})},$$
which rises toward the child's own $\kappa_i$ as the vocabulary saturates.
Panel C plots $\gamma$ for both populations against the *same absolute*
amount of input — training tokens for the models, and for each child the
cumulative tokens implied by their age under the pooled input rate (see
Population Input Range Estimation), so that age serves as experience. The
models sit at $\gamma = \beta \approx 0.32$, a third of a unit accumulator,
uniformly across the token axis; children rise from near the accumulator
boundary toward $\kappa_i \sim 10$ — an order of magnitude higher — and vary
widely across individuals.

The contrast is now formal, and it is about functional *form*. The model's
reducible loss follows a single-exponent power law, so its fractional return
is pinned at $\beta$; the child's does not, so its fractional return climbs.
Neither learner has flat returns in absolute terms — both slow as they
approach their floor — but the model spends its returns down along a fixed
exponent while the child's exponent grows.

Three caveats bound the comparison. First, the identity is *aggregate*: $L$
is the mean over CDI words and $\gamma$ is the return of that aggregate, not
of any single word — the quantization account (Michaud et al., 2023) is what
links the smooth aggregate curve to discrete per-word acquisition. Second,
the two $\gamma$'s share a *definition* — reducible loss removed per $e$-fold
of experience — but not a *task*: the model's loss is next-token surprisal
and the child's is production surprisal, so the comparison is of the rate and
shape of returns, not of one identical quantity. Third, the child's *rising*
$\gamma$ is the loss-currency reflection of a **constant** $\kappa_i$
(production probability saturating), not of $\kappa$ increasing over
development; the invariant claim is the constant-versus-rising exponent. The
horizontal placement of the child curves carries the input-rate uncertainty
($\pm 1$ SD $\approx \times 1.7$; panel C ribbons), which shifts children
along the token axis without changing $\gamma$.

<!-- [BIB NEEDED] Hoffmann et al. (2022) / Chinchilla and Michaud et al. (2023)
     are named in plain text (as in the block above). The main text already uses
     @hoffmann2022 and @michaud2023 markers; if those bib entries are verified,
     these can be promoted to live @ markers. Fitted constants (E, B, beta, R^2)
     and the panel C series are from build_cache_short.R section 8 — wire to
     inline values on promotion rather than hard-coding. -->
```

# Generating the Model-Ladder Quantile Fans

```md
## How the model-implied quantile bands are generated  [draft for promotion — MCF to review]

The percentile bands ("fans") overlaid on the empirical trajectories in the model
figure are not closed-form quantiles; for each dataset we generate them by
Monte-Carlo simulation from the fitted M3 posterior — a synthetic population of
children whose vocabulary percentiles we read off at each age.

**Drawing a synthetic population.** From each dataset's M3 fit we take the
posterior medians of the population parameters: mean efficiency $\mu_\xi$,
population scaling $\kappa_{\text{pop}} = 1 + \delta$, the between-child SDs of
efficiency and acceleration ($\sigma_a$, $\sigma_b$), their correlation $\rho$, and
the item-difficulty SD $\tau_\delta$. We draw $N = 500$ synthetic children by
sampling their (efficiency, acceleration) deviations from the fitted bivariate
normal,
$(a_i, b_i) \sim \mathcal{N}\!\big(\mathbf{0}, \Sigma(\sigma_a, \sigma_b, \rho)\big)$,
so child $i$ has $\xi_i = \mu_\xi + a_i$ and $\kappa_i = \kappa_{\text{pop}} + b_i$;
a fixed seed makes the bands reproducible. Because the Bayesian hierarchy estimates
the population covariance $\Sigma$ directly (with partial pooling), sampling from it
*is* the population draw — unlike a frequentist point-estimate fit, there are no
unshrunken per-child BLUPs to resample.

**Each child's expected vocabulary curve.** Ages run on a per-dataset grid spanning
the full range of that dataset's observed ages, so the bands cover the plotted
trajectories without extrapolating beyond the data. We draw $M$ synthetic item difficulties
$\delta_j \sim \mathcal{N}(0, \tau_\delta)$ to stand in for a typical item set;
child $i$'s expected productive vocabulary, as a proportion of items, at age $t$ is
$$V_i(t) = \frac{1}{M}\sum_{j=1}^{M} \operatorname{logit}^{-1}\!\big(\xi_i + \kappa_i \log(t/a_0) + \log H - \delta_j\big),$$
an *expected* proportion, so each simulated child's curve is smooth and the width
of the fan reflects between-child variation alone.

**Quantiles.** At each (dataset, age) the plotted bands are the 10th, 25th, 50th,
75th, and 90th percentiles of $\{V_i(t)\}_{i=1}^{500}$. The simulation is
implemented in `studies/bayes_long/03_fan.R` (mirrored in `paper/build_cache_short.R`).
```

# Comparison to a Non-Parametric Quantile Model

## The parametric fan reproduces the data's quantile structure  [draft for promotion — MCF to review]

```md
[draft for promotion — MCF to review]

To check that the accumulator model is not imposing its shape on the data, we compare
its model-implied quantile fan to a *non-parametric* benchmark: a GAMLSS beta-regression
of proportion-produced on age (a penalized spline on the mean and on the scale; the model
used for the CDI norms), fit to the same per-administration sumscores. Across all five
datasets the parametric fan (@fig-gamlss-overlay, blue) tracks both the non-parametric
GAMLSS fit (orange) and the empirical quantiles (circles) — the accelerating accumulator
is a faithful, low-dimensional compression of the data's quantile structure, not an
artifact of its functional form.

Two implementation notes. First, we let the GAMLSS choose its smoothing by marginal
likelihood; over-smoothing the mean spline makes the non-parametric fit artificially
tight at older ages. Second, we draw each simulated child's curve from the model's
*fitted* per-child acceleration $\kappa_i$ rather than a fresh Gaussian resample of the
population variance. This matters because the fitted acceleration distribution is
right-skewed with a light lower tail (@fig-kappa-skew): a symmetric resample invents
low-acceleration children who do not exist and over-disperses the lower tail at older
ages, whereas the fitted-$\kappa$ fan matches the empirical quantiles at every age.
```

```{r fig-gamlss-overlay, fig.width=13, fig.height=3.5}
#| label: fig-gamlss-overlay
#| fig-cap: "[draft caption — MCF] Parametric accumulator fan (blue, drawn from fitted per-child acceleration) vs. a non-parametric GAMLSS beta-regression (orange), with empirical 10th/50th/90th percentiles (open circles). Lines are the 10/50/90 percentiles of each model. Color denotes model."
ll <- si_gamlss$overlay$lines |> dplyr::filter(model != "M3 (MVN draw)") |>
  tidyr::pivot_longer(c(q10, q50, q90), names_to = "pct", values_to = "v") |>
  dplyr::mutate(dataset = factor(dataset, levels = si_gamlss$levels))
empd <- si_gamlss$overlay$emp |> dplyr::mutate(dataset = factor(dataset, levels = si_gamlss$levels))
ggplot() +
  geom_point(data = empd, aes(age, v), shape = 1, size = 1.2, stroke = .6, color = "grey25") +
  geom_line(data = ll, aes(age, v, color = model, group = interaction(model, pct)), linewidth = .7) +
  facet_wrap(~dataset, nrow = 1) +
  scale_color_manual(values = c("M3 (fitted-κ)" = "#2c7fb8", "GAMLSS (auto-λ)" = "#d95f02"), name = NULL) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Age (months)", y = "Proportion of items produced") +
  theme_minimal(base_size = 9) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(), strip.text = element_text(face = "bold"))
```

```{r fig-kappa-skew, fig.width=13, fig.height=3}
#| label: fig-kappa-skew
#| fig-cap: "[draft caption — MCF] Distribution of fitted per-child acceleration (kappa) in each dataset (grey), with a matched Normal (red). Every dataset is right-skewed with a lighter lower tail than the Normal, so a Gaussian resample of the population variance would place too many children at low acceleration. Dotted line: kappa = 1 (pure accumulation, the LLM regime)."
K <- si_gamlss$kappa |> dplyr::mutate(dataset = factor(dataset, levels = si_gamlss$levels))
nrm <- K |> dplyr::group_by(dataset) |> dplyr::summarise(m = mean(kappa), s = sd(kappa), .groups = "drop") |>
  dplyr::rowwise() |> dplyr::mutate(x = list(seq(-6, 35, .5))) |> tidyr::unnest(x) |> dplyr::mutate(dens = dnorm(x, m, s))
ggplot(K, aes(kappa)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 1.5, fill = "grey72", color = "white") +
  geom_line(data = nrm, aes(x, dens), color = "#c41e37", linewidth = .7) +
  geom_vline(xintercept = 1, linetype = "dotted") +
  facet_wrap(~dataset, nrow = 1, scales = "free_y") + coord_cartesian(xlim = c(-6, 32)) +
  labs(x = "Fitted per-child acceleration (κ)", y = "Density") +
  theme_minimal(base_size = 9) +
  theme(panel.grid.minor = element_blank(), strip.text = element_text(face = "bold"))
```
